### 2021-04-19 papers/final/paper.tex
start individual progress report

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,authorversion]{acmart}
%% NOTE that a single column version may be required for
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to
%% \documentclass[manuscript,screen,review]{acmart}

\begin{document}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\title{Musical Deep Learning}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Nhut Minh Phan}
\email{phan92@uw.edu}
\affiliation{%
  \institution{University of Washington}
  \city{Bothell}
  \state{WA}
  \country{USA}
}

\author{Alex Kyllo}
\email{akyllo@uw.edu}
\affiliation{%
  \institution{University of Washington}
  \city{Bothell}
  \state{WA}
  \country{USA}
}

\begin{abstract}
  The abstract will go here.
\end{abstract}

\keywords{deep learning, neural networks, music}

\maketitle

\section{Introduction}

Example citation \cite{huang_music_2018}.

\section{Methods}

\section{Results}

\section{Discussion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{paper}

\end{document}
\endinput

### 2021-04-19 papers/progress/alex/progress-alex.tex
start individual progress report

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,authorversion]{acmart}
%% NOTE that a single column version may be required for
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to
%% \documentclass[manuscript,screen,review]{acmart}

\begin{document}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\title{Progress Report for CSS 586: \\
Modeling Latent Patterns in Music}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

\author{Alex Kyllo}
\email{akyllo@uw.edu}
\affiliation{%
  \institution{University of Washington}
  \city{Bothell}
  \state{WA}
  \country{USA}
}

\begin{abstract}
This report explores recent research in modeling music with deep learning and
provides a progress report on a project to train a generative model of classical
music on the MusicNet dataset.
\end{abstract}

\keywords{deep learning, neural networks, music}

\maketitle

\section{Introduction}

Machine learning models of music have interesting applications in music
information retrieval and creative tools for musical artists and educators.
Music is complex and challenging to model because it exhibits a hierarchy of
recurring patterns.

Depending on the task, machine learning models of music may be trained on the
audio signal itself, either in a time domain or a frequency domain
representation, or they may be trained on a digital symbolic representation of
music, the most common of which is MIDI (Musical Instrument Digital Interface)
notation. MIDI is an encoding of music as streams of bytes in one or more tracks
or channels, each representing a sequence of 128 possible pitch values, along
with timing, pressure and instrument values. A music transcription model may
convert an audio signal into MIDI (which can then be transcribed into sheet
music), while a synthesizer model can convert MIDI representations into audio
signals.

\section{Related Work}

Google's \href{https://magenta.tensorflow.org/}{Magenta} is an umbrella project
for music deep learning research and development of software tools to expose
these models for use by creative artists and students.

MusicVAE is a variational LSTM autoencoder for MIDI that incorporates a
novel hierarchical structure using a "composer" recurrent layer in its encoder
model to better capture structure at multiple
levels \cite{roberts_hierarchical_2018}.

Music Transformer is a generative model that borrows its approach from the
Natural Language Processing (NLP) domain, using an attention network to model
MIDI music as a sequence of discrete tokens with relative positional dependencies
\cite{huang_music_2018}.

\section{Planned Methods}

\section{Progress}

\section{Future Work}

\bibliographystyle{ACM-Reference-Format}
\bibliography{progress-alex}

\end{document}
\endinput

### 2021-04-19 papers/progress/alex/progress-alex.tex
save progress report paper work

@@
 with timing, pressure and instrument values. A music transcription model may
-convert an audio signal into MIDI (which can then be transcribed into sheet
-music), while a synthesizer model can convert MIDI representations into audio
-signals.
+convert an audio signal into MIDI, which can easily be converted into other
+symbolic representations such as sheet music, while a synthesizer model can
+convert MIDI representations into audio signals.
 
@@
 
+Other research has focused on modeling raw audio waveforms directly. WaveNet is
+a causal convolutional neural network for generating raw audio waveforms,
+developed by Google DeepMind, which achieves state of the art performance in
+generating natural sounding speech from text, but is also capable of generating
+short, realistic snippets of audio music \cite{oord_wavenet_2016}. Another model
+named SampleRNN generates raw audio waveforms using a three-tier hierarchy of
+gated recurrent units (GRU) to model recurrent structure at multiple temporal
+resolutions \cite{mehri_samplernn_2017}.
+
+A paper from Boston University describes an effort to combine the symbolic and
+waveform approaches to music modeling \cite{manzelli_conditioning_2018}.
+
 \section{Planned Methods}

### 2021-04-19 papers/progress/alex/progress-alex.tex
save report updates

@@
 MusicVAE is a variational LSTM autoencoder for MIDI that incorporates a
-novel hierarchical structure using a "composer" recurrent layer in its encoder
+novel hierarchical structure using a ``composer'' recurrent layer in its encoder
 model to better capture structure at multiple
@@
 
+The major advantage of working with the symbolic representation of music is that
+it is of far lower dimensionality than the raw audio waveforms of a recorded
+performance,
+
 Other research has focused on modeling raw audio waveforms directly. WaveNet is
@@
 generating natural sounding speech from text, but is also capable of generating
-short, realistic snippets of audio music \cite{oord_wavenet_2016}. Another model
-named SampleRNN generates raw audio waveforms using a three-tier hierarchy of
-gated recurrent units (GRU) to model recurrent structure at multiple temporal
-resolutions \cite{mehri_samplernn_2017}.
+short, realistic snippets of audio music \cite{oord_wavenet_2016}.
+
+Another model named SampleRNN generates raw audio waveforms using a three-tier
+hierarchy of gated recurrent units (GRU) to model recurrent structure at
+multiple temporal resolutions \cite{mehri_samplernn_2017}.
+
+Jukebox by OpenAI utilizes a vector-quantized variational autoencoder (VQ-VAE)
+to compress raw audio into a sequence of discrete codes and models these
+sequences using autoregressive transformers to generate music
+\cite{dhariwal2020jukebox}.
+
+Audio data can also be modeled in the frequency domain through the use of
+Fourier analysis. The recent MelNet model is trained on spectrograms and can
+learn musical structures such as melody and harmony and variations in volume,
+timbre and rhythm \cite{vasquez2019melnet}.
 
 A paper from Boston University describes an effort to combine the symbolic and
-waveform approaches to music modeling \cite{manzelli_conditioning_2018}.
+waveform approaches to music modeling, by training an LSTM to learn melodic
+structure of different styles of music, then providing generations from this
+model as conditioning inputs to a WaveNet-based raw audio generator
+\cite{manzelli_conditioning_2018}.
 

### 2021-04-19 papers/progress/alex/progress-alex.tex
save report

@@
 
-The major advantage of working with the symbolic representation of music is that
+A major advantage of working with the symbolic representation of music is that
 it is of far lower dimensionality than the raw audio waveforms of a recorded
-performance,
+performance, which makes it less computationally expensive. However, there are
+many aspects of musical performance that are not captured by a symbolic
+representation, so the expressiveness of symbolic generative models is
+constrained \cite{manzelli_conditioning_2018}.
 

### 2021-04-22 musiclearn/processing.py
add note-seq dependency for reading MIDI files into Python

@@ -10,6 +10,7 @@ from scipy.io import wavfile
 import numpy as np
 
 
+
 class WavDataGenerator(keras.utils.Sequence):
     """A Keras Sequence for providing audio (.wav) file data in batches."""
 
@@ -83,3 +84,7 @@ def test_batch_maxlen():
     wdg = WavDataGenerator(dir, batch_size=batch_size, max_len=sample_rate)
     batch = wdg[0]
     assert batch.shape == (batch_size, sample_rate)
+
+
+class MIDIDataGenerator(keras.utils.Sequence):
+    """Generates note sequence tensors from MIDI files"""

### 2021-04-24 papers/progress/jack/progress-jack.tex
Add jacks report

--- 2021-04-19 papers/progress/alex/progress-alex.tex
@@
 \title{Progress Report for CSS 586: \\
-Modeling Latent Patterns in Music}
+Music Transcription}
 
@@
 
-\author{Alex Kyllo}
-\email{akyllo@uw.edu}
+\author{Jack Phan}
+\email{phan92@uw.edu}
 \affiliation{%
@@
 \begin{abstract}
-This report explores recent research in modeling music with deep learning and
-provides a progress report on a project to train a generative model of classical
-music on the MusicNet dataset.
+
 \end{abstract}
@@
 
-Machine learning models of music have interesting applications in music
-information retrieval and creative tools for musical artists and educators.
-Music is complex and challenging to model because it exhibits a hierarchy of
-recurring patterns.
-
-Depending on the task, machine learning models of music may be trained on the
-audio signal itself, either in a time domain or a frequency domain
-representation, or they may be trained on a digital symbolic representation of
-music, the most common of which is MIDI (Musical Instrument Digital Interface)
-notation. MIDI is an encoding of music as streams of bytes in one or more tracks
-or channels, each representing a sequence of 128 possible pitch values, along
-with timing, pressure and instrument values. A music transcription model may
-convert an audio signal into MIDI, which can easily be converted into other
-symbolic representations such as sheet music, while a synthesizer model can
-convert MIDI representations into audio signals.
-
 \section{Related Work}
-
-Google's \href{https://magenta.tensorflow.org/}{Magenta} is an umbrella project
-for music deep learning research and development of software tools to expose
-these models for use by creative artists and students.
-
-MusicVAE is a variational LSTM autoencoder for MIDI that incorporates a
-novel hierarchical structure using a "composer" recurrent layer in its encoder
-model to better capture structure at multiple
-levels \cite{roberts_hierarchical_2018}.
-
-Music Transformer is a generative model that borrows its approach from the
-Natural Language Processing (NLP) domain, using an attention network to model
-MIDI music as a sequence of discrete tokens with relative positional dependencies
-\cite{huang_music_2018}.
-
-Other research has focused on modeling raw audio waveforms directly. WaveNet is
-a causal convolutional neural network for generating raw audio waveforms,
-developed by Google DeepMind, which achieves state of the art performance in
-generating natural sounding speech from text, but is also capable of generating
-short, realistic snippets of audio music \cite{oord_wavenet_2016}. Another model
-named SampleRNN generates raw audio waveforms using a three-tier hierarchy of
-gated recurrent units (GRU) to model recurrent structure at multiple temporal
-resolutions \cite{mehri_samplernn_2017}.
-
-A paper from Boston University describes an effort to combine the symbolic and
-waveform approaches to music modeling \cite{manzelli_conditioning_2018}.
 
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{progress-alex}
+\bibliography{progress-jack}
 

### 2021-04-24 musiclearn/processing.py
start a notebook to play with midi

@@ -10,7 +10,6 @@ from scipy.io import wavfile
 import numpy as np
 
 
-
 class WavDataGenerator(keras.utils.Sequence):
     """A Keras Sequence for providing audio (.wav) file data in batches."""
 
@@ -61,7 +60,7 @@ class WavDataGenerator(keras.utils.Sequence):
 def test_batch_filenames():
     """Test that it can create shuffled batches of filenames"""
     dotenv.load_dotenv()
-    dir = Path(os.getenv("MUSIC_DATA_DIR")) / "train_data"
+    dir = Path(os.getenv("MUSICNET_DIR")) / "train_data"
     wdg = WavDataGenerator(dir, batch_size=2)
     batch = wdg._get_batch_filenames(0)
     assert Path(batch[0]).is_file
@@ -77,9 +76,9 @@ def test_batch_filenames():
 def test_batch_maxlen():
     """Test that it retrieves data up to a max length"""
     dotenv.load_dotenv()
-    dir = Path(os.getenv("MUSIC_DATA_DIR")) / "train_data"
+    dir = Path(os.getenv("MUSICNET_DIR")) / "train_data"
     # get 1 second of data per file
-    sample_rate = int(os.getenv("SAMPLE_RATE"))
+    sample_rate = int(os.getenv("MUSICNET_SAMPLE_RATE"))
     batch_size = 2
     wdg = WavDataGenerator(dir, batch_size=batch_size, max_len=sample_rate)
     batch = wdg[0]

### 2021-04-24 notebooks/working_with_midi.py

@@ -1,0 +1,30 @@
+# %%
+import os
+import pathlib
+import sys
+
+sys.path.append("..")
+import note_seq
+import music21
+import dotenv
+
+dotenv.load_dotenv()
+
+# %% [markdown]
+# ## Using Magenta note_seq package
+
+# %%
+midi_dir = pathlib.Path(os.getenv("MUSICNET_MIDI_DIR"))
+mid_2494 = midi_dir / "Beethoven" / "2494_qt11_1.mid"
+
+# %%
+notes_2494 = note_seq.midi_io.midi_file_to_note_sequence(mid_2494)
+print(notes_2494.instrument_infos)
+note_seq.plot_sequence(notes_2494)
+
+# %% [markdown]
+# ## Using Music21 package
+parse_2494 = music21.converter.parse(mid_2494)
+parts = music21.instrument.partitionByInstrument(parse_2494)
+print(parts)
+print(list(parts.parts[0])[1:20])

### 2021-04-25 notebooks/working_with_midi.py
pretty_midi piano roll example

@@ -7,15 +7,42 @@ sys.path.append("..")
 import note_seq
 import music21
 import dotenv
+import pretty_midi
+import pandas as pd
 
 dotenv.load_dotenv()
-
-# %% [markdown]
-# ## Using Magenta note_seq package
-
 # %%
 midi_dir = pathlib.Path(os.getenv("MUSICNET_MIDI_DIR"))
 mid_2494 = midi_dir / "Beethoven" / "2494_qt11_1.mid"
+
+# %% [markdown]
+# ## Using the pretty_midi package
+
+# %%
+pretty_2494 = pretty_midi.PrettyMIDI(str(mid_2494))
+print(pretty_2494.instruments)
+print(pretty_2494.instruments[0].notes)
+
+# %% [markdown]
+# One representation is a piano roll, which is a numpy 2D array of
+# notes over time steps at a given frames per second rate. There are 128 notes
+# (0-127) and if the note is playing during that time step it will be 1, else 0.
+# This allows chords but assumes a single instrument. So it loses instrument
+# information unless you produce separate per-instrument piano rolls.
+
+# %%
+# returns a (notes, time) dimension numpy array
+piano_2494 = pretty_2494.get_piano_roll(fs=10)
+# Show as pandas DataFrame where rows are time steps and columns are note values
+pd.DataFrame(piano_2494.T)
+
+# %% [markdown]
+# ## Using Magenta note_seq package
+#
+# note_seq is a very powerful but undocumented package created by the Magenta
+# project. It includes tools for converting between MIDI and a Protocol Buffer
+# serializeable representation of symbolic music. It uses pretty_midi internally
+# but adds some bells and whistles like plotting.
 
 # %%
 notes_2494 = note_seq.midi_io.midi_file_to_note_sequence(mid_2494)
@@ -28,3 +55,5 @@ parse_2494 = music21.converter.parse(mid_2494)
 parts = music21.instrument.partitionByInstrument(parse_2494)
 print(parts)
 print(list(parts.parts[0])[1:20])
+
+# %%

### 2021-04-25 musiclearn/config.py
add config.py

@@ -1,0 +1,8 @@
+import os
+import dotenv
+
+dotenv.load_dotenv()
+
+MUSICNET_DIR = os.getenv("MUSICNET_DIR")
+MUSICNET_MIDI_DIR = os.getenv("MUSICNET_MIDI_DIR")
+MUSICNET_SAMPLE_RATE = os.getenv("MUSICNET_SAMPLE_RATE")

### 2021-04-25 papers/progress/alex/progress-alex.tex
add note-seq dependency for reading MIDI files into Python
add some midi deps and paper work

@@
 This report explores recent research in modeling music with deep learning and
-provides a progress report on a project to train a generative model of classical
-music on the MusicNet dataset.
+provides a progress report on a course project to train a generative model of
+classical music on the MusicNet dataset.
 \end{abstract}
@@
 convert an audio signal into MIDI, which can easily be converted into other
-symbolic representations such as sheet music, while a synthesizer model can
-convert MIDI representations into audio signals.
+symbolic representations such as sheet music for human performers to read from,
+while a synthesizer model can convert MIDI representations into audio signals.
 
@@
 
-A paper from Boston University describes an effort to combine the symbolic and
-waveform approaches to music modeling, by training an LSTM to learn melodic
-structure of different styles of music, then providing generations from this
-model as conditioning inputs to a WaveNet-based raw audio generator
-\cite{manzelli_conditioning_2018}.
+Prior work points out that the division between symbolic music notes and the
+sounds of music is analogous to the division between symbolic language and
+utterances in speech, which may inspire ideas for combining the two approaches
+\cite{hawthorne2019enabling}. A paper from Boston University describes an effort
+to combine the symbolic and waveform approaches to music modeling, by training
+an LSTM to learn melodic structure of different styles of music, then providing
+generations from this model as conditioning inputs to a WaveNet-based raw audio
+generator \cite{manzelli_conditioning_2018}.
 
-\section{Planned Methods}
+\section{Planned Methods and Progress}
 
-\section{Progress}
+\subsection{Data Preprocessing}
+
+This project will focus on the generative modeling of symbolic music using MIDI
+data, because of the advantages of symbolic music models in representing
+long-term structure in musical compositions to produce generations with coherent
+use of repetition over long time scales.
+
+\subsection{Model Fitting}
+
+We will explore several modeling approaches to generating symbolic music:
+
+\begin{itemize}
+  \item{Sliding window sequence prediction with RNNs (LSTM/GRU)}
+  \item{Sliding window sequence prediction with Transformers}
+  \item{Latent space interpolation with Sequential VAEs}
+  \item{Latent space interpolation with Sequential GANs}
+\end{itemize}
+
+\subsection{Model Evaluation}
+
+Evaluation of generative models is challenging because there is no equivalent of
+an accuracy metric like what is used in supervised learning. Generative models
+are typically evaluated using a combination of qualitative metrics whereby human
+judges rate the quality of the generated examples (essentially a Turing test),
+and quantitative metrics that assess the differences in the parametric
+distributions of generated and real examples. Yang and Lerch (2020) proposes a
+set of metrics informed by music theory, for probabilistically evaluating how
+similar the generations are to known sample distributions of real music
+\cite{yang_evaluation_2020}. These metrics include counts, ranges, histograms
+and transition matrices of pitches and note lengths, then the Kullback-Leibler
+divergence and overlapping area of the probability density functions are use to
+compare against known reference distributions per musical genre
+\cite{yang_evaluation_2020}. Due to the cost and time requirements associated
+with designing a human subjects experiment, we plan to utilize this quantitative
+approach to generation quality assessment.
 

### 2021-04-25 notebooks/working_with_midi.py

@@ -1,34 +1,40 @@
 # %%
-import os
 import pathlib
 import sys
-
-sys.path.append("..")
 import note_seq
 import music21
-import dotenv
 import pretty_midi
 import pandas as pd
 
-dotenv.load_dotenv()
+sys.path.append("..")
+from musiclearn import config
+
 # %%
-midi_dir = pathlib.Path(os.getenv("MUSICNET_MIDI_DIR"))
+midi_dir = pathlib.Path(config.MUSICNET_MIDI_DIR)
 mid_2494 = midi_dir / "Beethoven" / "2494_qt11_1.mid"
 
 # %% [markdown]
 # ## Using the pretty_midi package
-
+#
+# `pretty_midi` encodes the note pitch values, velocities and start and end
+# times of each instrument track. Gaps in timing between notes are rests. It
+# includes methods for estimating the tempo and time signature, and converting
+# between representations.
 # %%
 pretty_2494 = pretty_midi.PrettyMIDI(str(mid_2494))
-print(pretty_2494.instruments)
-print(pretty_2494.instruments[0].notes)
+dir(pretty_2494)
 
-# %% [markdown]
-# One representation is a piano roll, which is a numpy 2D array of
+# %%
+print(pretty_2494.instruments)
+print(pretty_2494.instruments[0].notes[0:50])
+
+# %% [markdown] One representation is a piano roll, which is a numpy 2D array of
 # notes over time steps at a given frames per second rate. There are 128 notes
 # (0-127) and if the note is playing during that time step it will be 1, else 0.
 # This allows chords but assumes a single instrument. So it loses instrument
-# information unless you produce separate per-instrument piano rolls.
+# information unless you produce separate per-instrument piano rolls. It
+# produces a sparse matrix since most notes aren't being played at most
+# time steps.
 
 # %%
 # returns a (notes, time) dimension numpy array
@@ -51,6 +57,12 @@ note_seq.plot_sequence(notes_2494)
 
 # %% [markdown]
 # ## Using Music21 package
+#
+# Another package similar to pretty_midi is Music21. Instead of encoding the
+# start and end time of each note, it encodes a metronome in beats per minute,
+# and encodes rest symbols. This is closer to a musical score.
+
+# %%
 parse_2494 = music21.converter.parse(mid_2494)
 parts = music21.instrument.partitionByInstrument(parse_2494)
 print(parts)

### 2021-04-25 papers/progress/alex/progress-alex.tex
paper and deps updates

@@
 \pagestyle{plain}
-\title{Progress Report for CSS 586: \\
+\title{Progress Report for CSS 586 Course Project: \\
 Modeling Latent Patterns in Music}
@@
 provides a progress report on a course project to train a generative model of
-classical music on the MusicNet dataset.
+classical and popular music on the MusicNet and Lakh MIDI datasets.
 \end{abstract}
 
-\keywords{deep learning, neural networks, music}
+\keywords{deep learning, sequence learning, generative modeling, recurrent neural networks, music, MIDI}
 
@@
 levels \cite{roberts_hierarchical_2018}.
+
+MuseGAN  \cite{dong2017musegan}.
 
@@
 
+\subsection{Datasets}
+
 \subsection{Data Preprocessing}

### 2021-04-26 notebooks/working_with_midi.py
add miditoolkit dep

@@ -5,6 +5,8 @@ import note_seq
 import music21
 import pretty_midi
 import pandas as pd
+import pypianoroll
+from miditoolkit import midi
 
 sys.path.append("..")
 from musiclearn import config
@@ -28,19 +30,43 @@ dir(pretty_2494)
 print(pretty_2494.instruments)
 print(pretty_2494.instruments[0].notes[0:50])
 
-# %% [markdown] One representation is a piano roll, which is a numpy 2D array of
-# notes over time steps at a given frames per second rate. There are 128 notes
-# (0-127) and if the note is playing during that time step it will be 1, else 0.
-# This allows chords but assumes a single instrument. So it loses instrument
-# information unless you produce separate per-instrument piano rolls. It
-# produces a sparse matrix since most notes aren't being played at most
-# time steps.
+# %% [markdown]
+#
+# One representation is a piano roll, which is a numpy 2D array of notes over
+# time steps at a given frames per second rate (symbolic timing). There are 128
+# notes (0-127) and if the note is playing during that time step it will be 1,
+# else 0. This allows chords but assumes a single instrument. So it loses
+# instrument information unless you produce separate per-instrument piano rolls.
+# It produces a sparse matrix since most notes aren't being played at most time
+# steps.
 
 # %%
 # returns a (notes, time) dimension numpy array
 piano_2494 = pretty_2494.get_piano_roll(fs=10)
 # Show as pandas DataFrame where rows are time steps and columns are note values
 pd.DataFrame(piano_2494.T)
+
+# %% [markdown]
+# ## Using miditoolkit
+#
+# miditoolkit is similar to pretty_midi but uses symbolic (tick) timing instead
+# of absolute (wall clock) timing. It can also convert to and from pianorolls.
+# %%
+mid_obj = midi.parser.MidiFile(mid_2494)
+notes = mid_obj.instruments[0].notes
+print(notes[0:50])
+
+# %% [markdown]
+# ## Using pypianoroll
+#
+# Yet another package, `pypianoroll`, can read MIDI into multi-track pianoroll
+# sparse tensors, and also provides handy visualizations.
+
+# %%
+multitrack = pypianoroll.read(mid_2494)
+print(multitrack)
+multitrack.plot(mode="separate")
+multitrack.plot(mode="blended")
 
 # %% [markdown]
 # ## Using Magenta note_seq package
[…]

### 2021-04-26 papers/progress/alex/progress-alex.tex
start working on midi preprocess

@@
 
-\section{Planned Methods and Progress}
+\section{Planned Methods}
 
@@
 use of repetition over long time scales.
+
+Several choices must be made in how to preprocess MIDI files into training
+examples for a neural network. In order to accommodate polyphonic music, we will
+convert each MIDI file into a piano roll representation, wherein each instrument
+track is a sparse matrix of the velocity values for each of 128 possible pitch
+levels at each timestep. Because songs are typically at least a few minutes long
+and of widely varying length, we will crop songs into equal-length phrases.
+
+The result of this preprocessing is that each training example will be a 3D
+tensor of shape (tracks x notes x ticks).
 
@@
 
-\section{Future Work}
+\section{Progress and Remaining Work}
 

### 2021-04-26 musiclearn/processing.py

@@ -3,11 +3,12 @@ Preprocessing audio data (.wav files) for analysis.
 """
 import os
 from pathlib import Path
-import dotenv
 from tensorflow import keras
 from tensorflow.keras.preprocessing.sequence import pad_sequences
 from scipy.io import wavfile
 import numpy as np
+import pypianoroll
+from musiclearn import config
 
 
 class WavDataGenerator(keras.utils.Sequence):
@@ -59,8 +60,7 @@ class WavDataGenerator(keras.utils.Sequence):
 
 def test_batch_filenames():
     """Test that it can create shuffled batches of filenames"""
-    dotenv.load_dotenv()
-    dir = Path(os.getenv("MUSICNET_DIR")) / "train_data"
+    dir = Path(config.MUSICNET_DIR) / "train_data"
     wdg = WavDataGenerator(dir, batch_size=2)
     batch = wdg._get_batch_filenames(0)
     assert Path(batch[0]).is_file
@@ -75,8 +75,7 @@ def test_batch_filenames():
 
 def test_batch_maxlen():
     """Test that it retrieves data up to a max length"""
-    dotenv.load_dotenv()
-    dir = Path(os.getenv("MUSICNET_DIR")) / "train_data"
+    dir = Path(config.MUSICNET_DIR) / "train_data"
     # get 1 second of data per file
     sample_rate = int(os.getenv("MUSICNET_SAMPLE_RATE"))
     batch_size = 2
@@ -86,4 +85,33 @@ def test_batch_maxlen():
 
 
 class MIDIDataGenerator(keras.utils.Sequence):
-    """Generates note sequence tensors from MIDI files"""
+    """Generates multitrack piano roll tensors from MIDI files"""
+
+    def __init__(self, directory: os.PathLike):
+        """constructor"""
+
+
+def midi_to_multitrack(path: os.PathLike):
+    """"""
+    assert os.path.isfile(path)
+    mt = pypianoroll.Multitrack(str(path))
+    return mt
+
+
+def num_bars(mt: pypianoroll.Multitrack, resolution: int = 12) -> int:
+    """Get the integer number of bars in a Multitrack piano roll"""
+    return len(mt.downbeat) // resolution
+
+
+def phrase(
+    mt: pypianoroll.Multitrack,
+    idx: int,
+    resolution: int = 12,
+    bars_per_phrase: int = 4,
+):
+    """Get the phrase starting at the given index, where 0 is the first
+    phrase."""
[…]

### 2021-04-26 notebooks/working_with_midi.py

@@ -6,7 +6,7 @@ import music21
 import pretty_midi
 import pandas as pd
 import pypianoroll
-from miditoolkit import midi
+from miditoolkit import midi, pianoroll
 
 sys.path.append("..")
 from musiclearn import config
@@ -34,11 +34,11 @@ print(pretty_2494.instruments[0].notes[0:50])
 #
 # One representation is a piano roll, which is a numpy 2D array of notes over
 # time steps at a given frames per second rate (symbolic timing). There are 128
-# notes (0-127) and if the note is playing during that time step it will be 1,
-# else 0. This allows chords but assumes a single instrument. So it loses
-# instrument information unless you produce separate per-instrument piano rolls.
-# It produces a sparse matrix since most notes aren't being played at most time
-# steps.
+# notes (0-127) and the values are the note velocities. Or, if binarized, the
+# values are 1 for playing, else 0 for resting. This format allows chords but
+# assumes a single instrument. So it loses instrument information unless you
+# produce separate per-instrument piano rolls. It produces a sparse matrix since
+# most notes aren't being played at most time steps.
 
 # %%
 # returns a (notes, time) dimension numpy array
@@ -55,6 +55,8 @@ pd.DataFrame(piano_2494.T)
 mid_obj = midi.parser.MidiFile(mid_2494)
 notes = mid_obj.instruments[0].notes
 print(notes[0:50])
+pianoroll_2494 = pianoroll.notes2pianoroll(notes, ticks_per_beat=24)
+pd.DataFrame(pianoroll_2494)
 
 # %% [markdown]
 # ## Using pypianoroll
@@ -66,12 +68,12 @@ print(notes[0:50])
 multitrack = pypianoroll.read(mid_2494)
 print(multitrack)
 multitrack.plot(mode="separate")
-multitrack.plot(mode="blended")
+# multitrack.plot(mode="blended")
 
 # %% [markdown]
 # ## Using Magenta note_seq package
 #
-# note_seq is a very powerful but undocumented package created by the Magenta
+# note_seq is a powerful but undocumented package created by the Magenta
 # project. It includes tools for converting between MIDI and a Protocol Buffer
 # serializeable representation of symbolic music. It uses pretty_midi internally
 # but adds some bells and whistles like plotting.
@@ -93,3 +95,11 @@ parse_2494 = music21.converter.parse(mid_2494)
 parts = music21.instrument.partitionByInstrument(parse_2494)
 print(parts)
 print(list(parts.parts[0])[1:20])
[…]

### 2021-05-01 papers/progress/alex/progress-alex.tex
pianoroll bar cropping

@@
 
-MuseGAN  \cite{dong2017musegan}.
+MuseGAN \cite{dong2017musegan} is an application of Generative Adversarial
+Networks to polyphonic MIDI music generation, trained on four-bar phrases of a
+multi-track pianoroll representation of rock songs from the Lakh Midi Dataset
+\cite{raffel_learning-based_2016}.
 
@@
 examples for a neural network. In order to accommodate polyphonic music, we will
-convert each MIDI file into a piano roll representation, wherein each instrument
+convert each MIDI file into a pianoroll representation, wherein each instrument
 track is a sparse matrix of the velocity values for each of 128 possible pitch
@@
 The result of this preprocessing is that each training example will be a 3D
-tensor of shape (tracks x notes x ticks).
+tensor of shape (tracks x notes x ticks) and stacking the training examples will
+produce a 4D tensor, which is also typically the dimension of image data
+tensors in machine learning.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{first_four_bars.png}
+  \caption{The first four bars of Beethoven's Serioso String Quartet}
+  \Description{A pianoroll representation of cello, viola and violin parts.}
+\end{figure}
+
+Data augmentation is also possible--the literature suggests augmentation via
+pitch shifting the entire training example up or down by up to six semitones,
+and increasing or reducing the speed by up to 10\% in order to create additional
+training examples and reduce overfitting \cite{oore_this_2018}.
 

### 2021-05-01 notebooks/working_with_midi.py

@@ -6,6 +6,7 @@ import music21
 import pretty_midi
 import pandas as pd
 import pypianoroll
+import matplotlib.pyplot as plt
 from miditoolkit import midi, pianoroll
 
 sys.path.append("..")
@@ -24,7 +25,6 @@ mid_2494 = midi_dir / "Beethoven" / "2494_qt11_1.mid"
 # between representations.
 # %%
 pretty_2494 = pretty_midi.PrettyMIDI(str(mid_2494))
-dir(pretty_2494)
 
 # %%
 print(pretty_2494.instruments)
@@ -65,7 +65,7 @@ pd.DataFrame(pianoroll_2494)
 # sparse tensors, and also provides handy visualizations.
 
 # %%
-multitrack = pypianoroll.read(mid_2494)
+multitrack = pypianoroll.read(mid_2494, resolution=24)
 print(multitrack)
 multitrack.plot(mode="separate")
 # multitrack.plot(mode="blended")
@@ -103,3 +103,51 @@ print(list(parts.parts[0])[1:20])
 # Because an entire song is too long music is generally composed in bars or
 # multi-bar phrases, we plan to read MIDI files and segment them into phrases,
 # one training example per phrase.
+
+# %%
+def get_num_beats(multitrack, resolution):
+    return len(multitrack.downbeat) // resolution
+
+
+def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
+    start = bar_index * resolution
+    end = start + (num_bars * beats_per_bar) * resolution
+    return (start, end)
+
+
+# %%
+def bars(multitrack, start_index, num_bars, beats_per_bar, resolution):
+    start, end = get_bar_bounds(
+        start_index, num_bars, beats_per_bar, resolution
+    )
+    print(start)
+    print(end)
+    tracks = []
+    for track in multitrack.tracks:
+        print(track[start:end].shape)
+        tracks.append(
+            pypianoroll.Track(
+                name=track.name,
+                program=track.program,
+                pianoroll=track[start:end],
+            )
+        )
+    print(tracks)
+    return pypianoroll.Multitrack(tracks=tracks, resolution=resolution)
+
+
+# %% [markdown]
+# Plot four bars of Beethoven:
+# %%
+resolution = 24
+four_bars = bars(multitrack, 0, 4, 4, resolution)
+four_bars.plot()
+fig = plt.gcf()
[…]

### 2021-05-01 papers/progress/alex/progress-alex.tex
progress paper work

--- 2021-04-25 papers/progress/alex/progress-alex.tex
@@
 \begin{abstract}
-This report explores recent research in modeling music with deep learning and
-provides a progress report on a course project to train a generative model of
-classical and popular music on the MusicNet and Lakh MIDI datasets.
+This report explores recent research in symbolic music modeling with
+deep learning and provides a progress report on a course project to
+train a generative model of classical and popular music on the
+MusicNet and Lakh MIDI datasets.
 \end{abstract}
 
-\keywords{deep learning, sequence learning, generative modeling, recurrent neural networks, music, MIDI}
+\keywords{deep learning, sequence learning, generative modeling,
+  recurrent neural networks, music, MIDI}
 
@@
 
-Machine learning models of music have interesting applications in music
-information retrieval and creative tools for musical artists and educators.
-Music is complex and challenging to model because it exhibits a hierarchy of
-recurring patterns.
+Machine learning models of music have interesting applications in
+music information retrieval and creative tools for musical artists and
+educators. Generative models can create accompaniments for music,
+transfer styles between clips of music, and even generate entirely new
+music. Music is challenging to model because it exhibits a complex
+hierarchy of recurring patterns and long-range temporal dependencies,
+and because both musical scores and performances have multiple
+possible digital representations.
 
-Depending on the task, machine learning models of music may be trained on the
-audio signal itself, either in a time domain or a frequency domain
-representation, or they may be trained on a digital symbolic representation of
-music, the most common of which is MIDI (Musical Instrument Digital Interface)
-notation. MIDI is an encoding of music as streams of bytes in one or more tracks
-or channels, each representing a sequence of 128 possible pitch values, along
-with timing, pressure and instrument values. A music transcription model may
-convert an audio signal into MIDI, which can easily be converted into other
-symbolic representations such as sheet music for human performers to read from,
-while a synthesizer model can convert MIDI representations into audio signals.
+Depending on the task, machine learning models of music may be trained
+on the audio signal itself, either in a time domain or a frequency
+domain representation, or they may be trained on a digital symbolic
+representation of music, the most common of which is MIDI (Musical
+Instrument Digital Interface) notation. MIDI is an encoding of music
+as streams of bytes in one or more tracks or channels, each
+representing a sequence of 128 possible pitch values, along with
+timing, pressure and instrument values. A music transcription model
+may transcribe an audio signal as a MIDI score, which can easily be
+converted into other symbolic representations such as sheet music for
+human performers to read from, while a synthesizer model can convert
+MIDI representations into audio signals. A generative music model can
+be either trained to generate raw audio, or trained to produce a score
+that must be played by either a synthesizer or by humans to produce an
+audio music performance.
 
@@
 
-Google's \href{https://magenta.tensorflow.org/}{Magenta} is an umbrella project
-for music deep learning research and development of software tools to expose
-these models for use by creative artists and students.
+The state of the art in music generation has a long way to go before
+it can consistently generate music scores or performances that would
+be enjoyable and popular to listen to. There are a number of
+interesting, recent papers and projects in the area of music machine
+learning, several of which focus on learning long-term dependencies in
+music to produce longer clips with coherent structure.
 
-MusicVAE is a variational LSTM autoencoder for MIDI that incorporates a
-novel hierarchical structure using a ``composer'' recurrent layer in its encoder
-model to better capture structure at multiple
-levels \cite{roberts_hierarchical_2018}.
+Google's \href{https://magenta.tensorflow.org/}{Magenta} is an
+umbrella project for music deep learning research and development of
+software tools to expose these models for use by creative artists and
+students.
 
-MuseGAN  \cite{dong2017musegan}.
+MusicVAE, part of the Magenta project, is a variational LSTM
+autoencoder for MIDI that incorporates a novel hierarchical structure
+using a ``conductor'' recurrent layer in its decoder model to better
+capture structure at multiple levels and avoid the problem of
+``posterior/mode collapse'' whereby a generative model learns to
+ignore its latent code and rely on autoregression
+\cite{roberts_hierarchical_2018}.
 
-Music Transformer is a generative model that borrows its approach from the
-Natural Language Processing (NLP) domain, using an attention network to model
-MIDI music as a sequence of discrete tokens with relative positional dependencies
-\cite{huang_music_2018}.
+Another Magenta model called Music Transformer is a generative model
+that borrows its approach from the Natural Language Processing (NLP)
+domain, using a self-attention network to model MIDI music as a
+sequence of discrete tokens with relative positional dependencies
+\cite{huang_music_2018}. This model was trained on the MAESTRO dataset
+of Piano-e-competition performances \cite{hawthorne2019enabling} and
+its generated piano music received favorable qualitative (Likert
+scale) ratings from human listeners \cite{huang_music_2018}.
+
+MuseGAN \cite{dong2017musegan} is an application of Generative Adversarial
+Networks to polyphonic MIDI music generation, trained on four-bar phrases of a
+multi-track pianoroll representation of rock songs from the Lakh Midi Dataset
+\cite{raffel_learning-based_2016}.
 
@@
 
-Jukebox by OpenAI utilizes a vector-quantized variational autoencoder (VQ-VAE)
-to compress raw audio into a sequence of discrete codes and models these
-sequences using autoregressive transformers to generate music
+Jukebox by OpenAI utilizes a vector-quantized variational autoencoder
+(VQ-VAE) to compress raw audio into a sequence of discrete codes and
+models these sequences using autoregressive transformers to generate
+music of various popular genres, including singing and lyrics
 \cite{dhariwal2020jukebox}.
@@
 
-Prior work points out that the division between symbolic music notes and the
-sounds of music is analogous to the division between symbolic language and
-utterances in speech, which may inspire ideas for combining the two approaches
-\cite{hawthorne2019enabling}. A paper from Boston University describes an effort
-to combine the symbolic and waveform approaches to music modeling, by training
-an LSTM to learn melodic structure of different styles of music, then providing
-generations from this model as conditioning inputs to a WaveNet-based raw audio
-generator \cite{manzelli_conditioning_2018}.
+Prior work points out that the division between symbolic music notes
+and the sounds of music is analogous to the division between symbolic
+language and utterances in speech, which may inspire ideas for
+combining the two approaches \cite{hawthorne2019enabling}. A paper
+from Boston University describes an effort to combine the symbolic and
+waveform approaches to music modeling, by training an LSTM to learn
+melodic structure of different styles of music, then providing
+generations from this model as conditioning inputs to a WaveNet-based
+raw audio generator \cite{manzelli_conditioning_2018}.
 
-\section{Planned Methods and Progress}
+\section{Planned Methods}
+
+This research project will focus on the generative modeling of
+symbolic music using MIDI data as inputs, because of the advantages of
+symbolic music models in representing long-term structure in musical
+compositions to produce generations with coherent structure and use of
+repetition over long time scales.
 
@@
 
+Two primary datasets will be used for this research project: MusicNet,
+which is a collection of 330 freely licensed European classical music
+recordings with aligned MIDI scores \cite{thickstun2017learning}, and
+the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
+Commons licensed MIDI files of popular music songs that have been
+matched and aligned to MP3 recordings, and of which we will use a
+subset for model training and evaluation
+\cite{raffel_learning-based_2016}.
+
 \subsection{Data Preprocessing}
 
-This project will focus on the generative modeling of symbolic music using MIDI
-data, because of the advantages of symbolic music models in representing
-long-term structure in musical compositions to produce generations with coherent
-use of repetition over long time scales.
+Several choices must be made in how to preprocess binary MIDI files
+into training examples for a neural network. There are multiple
+open-source Python packages capable of reading MIDI files into NumPy
+array-based representations, such as pretty\_midi
+\cite{raffel_pretty_midi_2014} and Pypianoroll
+\cite{dong_pypianoroll_2018}. In order to accommodate polyphonic
+music, we will convert each MIDI file into a pianoroll representation,
+wherein each instrument track is a sparse matrix that multi-hot
+encodes the velocity values for each of 128 possible pitch levels at
+each timestep. Because songs are typically at least a few minutes long
+and of varying length, it will not be feasible to train with entire
+songs as examples, so we will crop songs into phrases of equal numbers
+of measures, to use as training data.
+
+The result of this preprocessing is that each training example will be
+a 3D tensor of shape (tracks x notes x ticks) and stacking the
+training examples will produce a 4D tensor.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{first_four_bars.png}
+  \caption{Pianoroll visualization of the first four measures of
+    Beethoven's Serioso String Quartet}
+  \Description{A pianoroll representation of cello, viola and violin
+    parts from a Beethoven piece.}
+\end{figure}
+
+Data augmentation is also possible and we plan to test its impact on
+our results--the literature suggests augmentation via pitch shifting
+the entire training example up or down by up to six semitones, and
+increasing or reducing the speed by up to 10\% in order to create
+additional training examples and reduce overfitting
+\cite{oore_this_2018}.
 
@@
 
-Evaluation of generative models is challenging because there is no equivalent of
-an accuracy metric like what is used in supervised learning. Generative models
-are typically evaluated using a combination of qualitative metrics whereby human
-judges rate the quality of the generated examples (essentially a Turing test),
-and quantitative metrics that assess the differences in the parametric
-distributions of generated and real examples. Yang and Lerch (2020) proposes a
-set of metrics informed by music theory, for probabilistically evaluating how
-similar the generations are to known sample distributions of real music
-\cite{yang_evaluation_2020}. These metrics include counts, ranges, histograms
-and transition matrices of pitches and note lengths, then the Kullback-Leibler
-divergence and overlapping area of the probability density functions are use to
+Evaluation of generative models is challenging because there is no
+equivalent of an accuracy metric like what is used in supervised
+learning. For autoencoder models we can measure how accurately the
+model can reconstruct its own inputs, but this does not tell us the
+quality of the interpolated examples. Generative models are typically
+evaluated using a combination of qualitative metrics whereby human
+judges rate the quality of the generated examples (essentially a
+Turing test), and quantitative metrics that assess the differences in
+the parametric distributions of generated and real examples. Yang and
+Lerch (2020) proposes a set of metrics informed by music theory, for
+probabilistically evaluating how similar the generations are to known
+sample distributions of real music \cite{yang_evaluation_2020}. These
+metrics include counts, ranges, histograms and transition matrices of
+pitches and note lengths, then the Kullback-Leibler divergence and
+overlapping area of the probability density functions are use to
 compare against known reference distributions per musical genre
-\cite{yang_evaluation_2020}. Due to the cost and time requirements associated
-with designing a human subjects experiment, we plan to utilize this quantitative
-approach to generation quality assessment.
+\cite{yang_evaluation_2020}. Due to the cost and time requirements
+associated with designing a human subjects experiment, we plan to
+utilize this quantitative approach to generation quality assessment.
 
-\section{Future Work}
+\section{Progress and Remaining Work}
+
+At this stage, we have decided on the modeling problem, downloaded the
+MIDI datasets, and begun exploring them to learn how to interpret,
+process, and visualize them. We have started writing a training data
+processing pipeline that will provide batches of training examples to
+Keras model layers for fitting.
 

### 2021-05-01 papers/progress/alex/progress-alex.tex
save work on MIDI pianoroll processing functions

@@
 ignore its latent code and rely on autoregression
-\cite{roberts_hierarchical_2018}.
+\cite{roberts_hierarchical_2018}. This model is capable of generating
+new melodies that blend two given melodies via latent space
+interpolation.
 
@@
 short, realistic snippets of audio music \cite{oord_wavenet_2016}.
-
 Another model named SampleRNN generates raw audio waveforms using a three-tier
@@
 subset for model training and evaluation
-\cite{raffel_learning-based_2016}.
+\cite{raffel_learning-based_2016}. Including this second dataset may
+help us to generalize the modeling approach beyond European classical
+music to include other popular genres and associated instruments.
 
@@
 music, we will convert each MIDI file into a pianoroll representation,
-wherein each instrument track is a sparse matrix that multi-hot
-encodes the velocity values for each of 128 possible pitch levels at
-each timestep. Because songs are typically at least a few minutes long
-and of varying length, it will not be feasible to train with entire
-songs as examples, so we will crop songs into phrases of equal numbers
-of measures, to use as training data.
+as visualized in Figure \ref{pianoroll}, wherein each instrument track
+is a sparse matrix that multi-hot encodes the velocity values for each
+of 128 possible pitch levels at each timestep. Because songs are
+typically at least a few minutes long and of varying length, it will
+not be feasible to train with entire songs as examples, so we will
+crop songs into phrases of equal numbers of measures to use as
+training data.
 
@@
     Beethoven's Serioso String Quartet}
+  \label{pianoroll}
   \Description{A pianoroll representation of cello, viola and violin

### 2021-05-01 musiclearn/config.py

@@ -3,6 +3,6 @@ import dotenv
 
 dotenv.load_dotenv()
 
-MUSICNET_DIR = os.getenv("MUSICNET_DIR")
-MUSICNET_MIDI_DIR = os.getenv("MUSICNET_MIDI_DIR")
-MUSICNET_SAMPLE_RATE = os.getenv("MUSICNET_SAMPLE_RATE")
+MUSICNET_DIR = str(os.getenv("MUSICNET_DIR"))
+MUSICNET_MIDI_DIR = str(os.getenv("MUSICNET_MIDI_DIR"))
+MUSICNET_SAMPLE_RATE = str(os.getenv("MUSICNET_SAMPLE_RATE"))

### 2021-05-01 musiclearn/processing.py

@@ -2,6 +2,7 @@
 Preprocessing audio data (.wav files) for analysis.
 """
 import os
+from typing import List
 from pathlib import Path
 from tensorflow import keras
 from tensorflow.keras.preprocessing.sequence import pad_sequences
@@ -77,7 +78,7 @@ def test_batch_maxlen():
     """Test that it retrieves data up to a max length"""
     dir = Path(config.MUSICNET_DIR) / "train_data"
     # get 1 second of data per file
-    sample_rate = int(os.getenv("MUSICNET_SAMPLE_RATE"))
+    sample_rate = int(config.MUSICNET_SAMPLE_RATE)
     batch_size = 2
     wdg = WavDataGenerator(dir, batch_size=batch_size, max_len=sample_rate)
     batch = wdg[0]
@@ -85,33 +86,73 @@ def test_batch_maxlen():
 
 
 class MIDIDataGenerator(keras.utils.Sequence):
-    """Generates multitrack piano roll tensors from MIDI files"""
+    """Generates multitrack pianoroll tensors from MIDI files."""
 
     def __init__(self, directory: os.PathLike):
         """constructor"""
+        self.directory = directory
+
+    def __getitem__(self, index):
+        """[] accessor. Get one batch as a numpy array."""
 
 
-def midi_to_multitrack(path: os.PathLike):
+def midi_to_multitrack(path: os.PathLike, resolution: int) -> pypianoroll.Multitrack:
     """"""
     assert os.path.isfile(path)
-    mt = pypianoroll.Multitrack(str(path))
+    mt = pypianoroll.read(str(path))
     return mt
 
 
-def num_bars(mt: pypianoroll.Multitrack, resolution: int = 12) -> int:
+def num_bars(multitrack: pypianoroll.Multitrack) -> int:
     """Get the integer number of bars in a Multitrack piano roll"""
-    return len(mt.downbeat) // resolution
+    assert multitrack.downbeat is not None and multitrack.resolution is not None
+    return len(multitrack.downbeat) // multitrack.resolution
 
 
-def phrase(
-    mt: pypianoroll.Multitrack,
-    idx: int,
-    resolution: int = 12,
-    bars_per_phrase: int = 4,
+def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
+    start = bar_index * resolution
+    end = start + (num_bars * beats_per_bar) * resolution
+    return (start, end)
+
+
[…]

### 2021-05-02 musiclearn/processing.py
add split_phrases function

@@ -103,9 +103,10 @@ def midi_to_multitrack(path: os.PathLike, resolution: int) -> pypianoroll.Multit
     return mt
 
 
-def num_bars(multitrack: pypianoroll.Multitrack) -> int:
-    """Get the integer number of bars in a Multitrack piano roll"""
-    assert multitrack.downbeat is not None and multitrack.resolution is not None
+def num_beats(multitrack: pypianoroll.Multitrack) -> int:
+    """Get the integer number of beats in a Multitrack piano roll"""
+    assert multitrack.downbeat is not None
+    assert multitrack.resolution is not None
     return len(multitrack.downbeat) // multitrack.resolution
 
 
@@ -133,21 +134,32 @@ def get_phrase(
         for track in multitrack.tracks
     ]
 
-    return pypianoroll.Multitrack(tracks=tracks, resolution=multitrack.resolution)
+    return pypianoroll.Multitrack(
+        tracks=tracks,
+        resolution=multitrack.resolution,
+        downbeat=multitrack.downbeat[start:end],
+        tempo=multitrack.tempo[start:end],
+    )
 
 
-def get_phrase_array(
-    multitrack: pypianoroll.multitrack, bars_per_phrase: int, beats_per_bar: int
-) -> np.array:
+def split_phrases(
+    multitrack: pypianoroll.Multitrack, bars_per_phrase: int, beats_per_bar: int
+):
     """get an array of equal length multitrack phrases from a single multitrack"""
-    # TODO
+    total_bars = num_beats(multitrack) // beats_per_bar
+    num_phrases = total_bars // bars_per_phrase
+    phrases = [
+        get_phrase(multitrack, ph * bars_per_phrase, bars_per_phrase, beats_per_bar)
+        for ph in range(num_phrases)
+    ]
+    return phrases
 
 
 def test_num_bars():
     path = Path(config.MUSICNET_MIDI_DIR)
     mid_file = list(path.glob("Beethoven/2494*.mid"))[0]
     mt = midi_to_multitrack(mid_file, 24)
-    assert num_bars(mt) == 609
+    assert num_beats(mt) == 609
 
 
 def test_get_phrase():
@@ -156,3 +168,14 @@ def test_get_phrase():
     mt = midi_to_multitrack(mid_file, 24)
     first_four_bars = get_phrase(mt, 0, 4, 4)
[…]

### 2021-05-02 papers/progress/alex/progress-alex.tex
add pianoroll to numpy function

@@
 The result of this preprocessing is that each training example will be
-a 3D tensor of shape (tracks x notes x ticks) and stacking the
+a 3D tensor of shape (tracks x ticks x pitches) and stacking the
 training examples will produce a 4D tensor.

### 2021-05-02 musiclearn/processing.py

@@ -110,6 +110,13 @@ def num_beats(multitrack: pypianoroll.Multitrack) -> int:
     return len(multitrack.downbeat) // multitrack.resolution
 
 
+def test_num_beats():
+    path = Path(config.MUSICNET_MIDI_DIR)
+    mid_file = list(path.glob("Beethoven/2494*.mid"))[0]
+    mt = midi_to_multitrack(mid_file, 24)
+    assert num_beats(mt) == 609
+
+
 def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
     start = bar_index * resolution
     end = start + (num_bars * beats_per_bar) * resolution
@@ -142,6 +149,14 @@ def get_phrase(
     )
 
 
+def test_get_phrase():
+    path = Path(config.MUSICNET_MIDI_DIR)
+    mid_file = list(path.glob("Beethoven/2494*.mid"))[0]
+    mt = midi_to_multitrack(mid_file, 24)
+    first_four_bars = get_phrase(mt, 0, 4, 4)
+    assert first_four_bars.tracks[0].pianoroll.shape == (384, 128)
+
+
 def split_phrases(
     multitrack: pypianoroll.Multitrack, bars_per_phrase: int, beats_per_bar: int
 ):
@@ -155,21 +170,6 @@ def split_phrases(
     return phrases
 
 
-def test_num_bars():
-    path = Path(config.MUSICNET_MIDI_DIR)
-    mid_file = list(path.glob("Beethoven/2494*.mid"))[0]
-    mt = midi_to_multitrack(mid_file, 24)
-    assert num_beats(mt) == 609
-
-
-def test_get_phrase():
-    path = Path(config.MUSICNET_MIDI_DIR)
-    mid_file = list(path.glob("Beethoven/2494*.mid"))[0]
-    mt = midi_to_multitrack(mid_file, 24)
-    first_four_bars = get_phrase(mt, 0, 4, 4)
-    assert first_four_bars.tracks[0].pianoroll.shape == (384, 128)
-
-
 def test_split_phrases():
     path = Path(config.MUSICNET_MIDI_DIR)
     mid_file = list(path.glob("Beethoven/2494*.mid"))[0]
@@ -179,3 +179,18 @@ def test_split_phrases():
     shapes = [m.tracks[0].pianoroll.shape for m in four_phrases]
     assert len(four_phrases) == 4
     # assert shapes == [(384, 128), (384, 128), (384, 128), (384, 128)]
+
+
[…]

### 2021-05-04 papers/progress/jack/progress-jack.tex

@@
 \pagestyle{plain}
-\title{Progress Report for CSS 586: \\
-Music Transcription}
+\title{Progress Report for CSS 586 Project: \\
+Music Generation Using Deep Learning}
 
@@
 
+The original goal of the project is music transcription, which is the translation of audio data
+to symbolic representation of musical notations. With further consideration, we decided to focus on 
+a different problem to better align our work. The new objective is music generation using deep learning.
+There are several methods that could be used to generate new music from a corpus of training data: MIDI, 
+time-domain sound pressure data, frequency-domain representation of the sound pressure. Regardless of the 
+data type, music generation involves the application of a sequence model such as recurrent neural network 
+(RNN). The focus of my contribution is using a combination of time-domain and frequency-domain information. 
+The current proges includes literature review, data exploration, learning about different representations 
+of an audio signal. I am working on building a data pipeline and experiencing with RNN models.
 \end{abstract}
 
-\keywords{deep learning, neural networks, music}
+\keywords{deep learning, neural networks, music generation, RNN}
 
@@
 
+Recurrent Neural Network (RNN) has shown its capability in recent years to model time-series sequences. 
+Advanced version of RNN such as Gated Recurrent Unit (GRU) and Long Short-Term Memory (LSTM) have been 
+successfully used for applications such as voice recognition, text to speech, and other applications 
+that involve time-series data. RNN could also be used to generate novel sequences of texts or sounds.
+
+A music recording is a digitized time-varying sequence of pressure displacement. Each point in the 
+sequence represents the amplitude of the pressure strength at a moment in time. The sequence could 
+also be converted to the frequency-domain representation, which describes its sound compoments in terms 
+of amplitude and frequency. A popular representation of a sound segment is spectrogram. The goal of this 
+project is generative modeling for short polyphonic music segments. We will build and train deep learning 
+models using spectrogram as input, and generate novel music segments using these models.
+
 \section{Related Work}
@@
 
+\subsection{Datasets}
+
+\subsection{Data Processing}
+
+\subsection{Model Building}
+
 \section{Progress}
+Talk about learning of RNN, GRU, and LSTM
+Talk about data exploration
 

### 2021-05-04 papers/progress/jack/progress-jack.tex
Update to Jack's report

@@
 
+Musical generative modeling is an interesting area of research. Recent innovations in this domain involve
+deep learning using a variety of modelling techniques. This section presents some of the recent works in
+this domain.
+
+Historically, the modelling of raw audio data is extremely challenging because a single second of audio recording
+could contains up to 44,100 samples in the case of music. Google DeepMind's WaveNet is one of the first models
+that successfully learn from raw audio data \cite{oord_wavenet_2016}. It is designed using a technique called 
+1D dialated causal convolution that allows an output to capture information from many inputs with a minimal
+computational cost. WaveNet was designed for the primary purpose of generating speech which mimics any human voice.
+However, since the architecture can be used to model any raw audio signal, it was also used to generate music.
+Although the outputs were far from being any masterpieces, the results show the possibility of generating musical
+pieces from the raw audio data.
+
 \section{Planned Methods}
@@
 \section{Progress}
-Talk about learning of RNN, GRU, and LSTM
-Talk about data exploration
 

### 2021-05-04 papers/progress/jack/progress-jack.tex
Update to Jack's report

@@
 The current proges includes literature review, data exploration, learning about different representations 
-of an audio signal. I am working on building a data pipeline and experiencing with RNN models.
+of an audio signal. We are working on building a data pipeline and experimenting with various RNN models.
 \end{abstract}
@@
 
-Historically, the modelling of raw audio data is extremely challenging because a single second of audio recording
-could contains up to 44,100 samples in the case of music. Google DeepMind's WaveNet is one of the first models
+Historically, the modeling of raw audio data is extremely challenging because a single second of audio recording
+could contains up to 44,100 samples. Google DeepMind's WaveNet is one of the first models
 that successfully learn from raw audio data \cite{oord_wavenet_2016}. It is designed using a technique called 
@@
 computational cost. WaveNet was designed for the primary purpose of generating speech which mimics any human voice.
-However, since the architecture can be used to model any raw audio signal, it was also used to generate music.
+Since the architecture can be used to model any raw audio signal, it was also used to generate music.
 Although the outputs were far from being any masterpieces, the results show the possibility of generating musical
@@
 
-\section{Planned Methods}
+An alternatve representation of the time-series signal is the frequency-domain representation. Existing generative
+models for audio have largely focused on the time-domain waveforms. In 2019, a generative model for audio in the
+frequency domain called MelNet was introduced \cite{vasquez2019melnet}. MelNet models the spectrograms, which are
+time-frequency representations of audio. The advantage of spectrograms is that they are compressed expressions of 
+the time-domain signal. MelNet leverages this advantage to generate high-fidelity audio samples, which capture 
+structures that are still challenging for time-domain models.
+
+\section{Planned Work}
+The focus of this project is generating short musical pieces by modeling the spectrograms of input audio. The
+following subsections describe our plan toward accomplishing this objective.
 
 \subsection{Datasets}
+We use a dataset called MusicNet published by University of Washington, the department of Com[uter Science 
+\cite{thickstun2017learning}. This dataset is a collection of 330 freely-licensed classical music recordings 
+captured under various studio and microphone conditions. The recordings are in .wav raw audio format and are
+polyphonic, meaning that multiple musical notes could present at a single point in time.
 

### 2021-05-04 papers/progress/jack/progress-jack.tex
Update to Jack's report

@@
 \subsection{Datasets}
-We use a dataset called MusicNet published by University of Washington, the department of Com[uter Science 
+We use a dataset called MusicNet published by University of Washington, the department of Computer Science 
 \cite{thickstun2017learning}. This dataset is a collection of 330 freely-licensed classical music recordings 
 captured under various studio and microphone conditions. The recordings are in .wav raw audio format and are
-polyphonic, meaning that multiple musical notes could present at a single point in time.
+polyphonic, meaning that multiple musical notes could present at a single time point. Furthermore, the music is
+performed by multiple instruments, which have unique sound signatures.
 
 \subsection{Data Processing}
+Each training example is a raw sound fragment, which has a length of several seconds, taken from an audio recording.
+We will transform the time-domain sample to the mel spectrogram representation. The mel scale is a logarithmic
+transformation of a signal frequency. The idea of this transformation is to replicate the human perception of the 
+difference between two sounds of different frequencies. For example, a person can easily tell the difference between 
+100 Hz and 200 Hz, but cannot distinguish 1000 Hz and 1100 Hz. This transformation is done by splitting the sample
+into overlapping windows, performing the Short Time Fourier Transformation on each window, converting the frequencies
+to mel scale, and finally converting the amplitude to decibels. Librosa, a python package for music analysis, has
+built-in functions for performing this tranformation. An example of a mel spectrogam is shown in Figure \ref{spectrogram}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{mel_spec.png}
+  \caption{Mel spectrogram of a 10 second sound segment from the dataset. The color represents the amplitude in dB.}
+  \label{spectrogram}
+\end{figure}
 
 \subsection{Model Building}
+We will implement sequence models such as RNN, GRU, and LSTM to model the mel spectrogram and use these models to
+generate novel spectrograms that could be converted back to audio.
 
 \section{Progress}
-
-\section{Future Work}
+Initially, a lot of time was spent on understanding different representation of the input data. We also learned 
+about the basics of sequence models such as RNN, GRU, and LSTM. Because the dataset is large and cannot be loaded into 
+the memory at once, we are working on a data pipeline that processes the raw audio files as described in the Section 
+Data Processing and presents training examples one by one to a model.
 

### 2021-05-04 papers/progress/alex/progress-alex.tex
save progress report

@@
 \title{Progress Report for CSS 586 Course Project: \\
-Modeling Latent Patterns in Music}
+Modeling Latent Patterns to Generate Symbolic Music}
 
@@
   deep learning and provides a progress report on a course project to
-train a generative model of classical and popular music on the
-MusicNet and Lakh MIDI datasets.
+  train and evaluate generative models of classical and popular music
+  on the MusicNet and Lakh MIDI datasets.
 \end{abstract}
 
-\keywords{deep learning, sequence learning, generative modeling,
-  recurrent neural networks, music, MIDI}
+\keywords{
+  generative modeling,
+  midi,
+  music,
+  recurrent neural networks,
+  sequence learning,
+  symbolic music,
+  deep learning
+}
 
@@
 as streams of bytes in one or more tracks or channels, each
-representing a sequence of 128 possible pitch values, along with
-timing, pressure and instrument values. A music transcription model
-may transcribe an audio signal as a MIDI score, which can easily be
+representing a sequence of 128 possible pitch values (where 0 is the
+lowest and 127 is the highest), along with timing, pressure and
+instrument identifier values. A music transcription model may
+transcribe an audio signal as a MIDI score, which can easily be
 converted into other symbolic representations such as sheet music for
@@
 MIDI representations into audio signals. A generative music model can
-be either trained to generate raw audio, or trained to produce a score
-that must be played by either a synthesizer or by humans to produce an
-audio music performance.
+be either trained to generate raw audio, or trained to produce a
+symbolic score that must be played by either a synthesizer or by
+humans to produce an audio music performance. This project focuses on
+generative modeling of symbolic (MIDI) music to generate original
+musical compositions.
 
@@
 it can consistently generate music scores or performances that would
-be enjoyable and popular to listen to. There are a number of
-interesting, recent papers and projects in the area of music machine
-learning, several of which focus on learning long-term dependencies in
-music to produce longer clips with coherent structure.
+be enjoyable and popular for humans to listen to, but a number of
+recent research projects have shown promising progress in this area.
 
@@
 
-MusicVAE, part of the Magenta project, is a variational LSTM
-autoencoder for MIDI that incorporates a novel hierarchical structure
-using a ``conductor'' recurrent layer in its decoder model to better
-capture structure at multiple levels and avoid the problem of
-``posterior/mode collapse'' whereby a generative model learns to
-ignore its latent code and rely on autoregression
+MusicVAE, part of the Magenta project, is a variational Long
+Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
+novel hierarchical structure using a ``conductor'' recurrent layer in
+its decoder model to better capture structure at multiple levels and
+avoid the problem of ``posterior/mode collapse'' whereby a generative
+model learns to ignore its latent code and rely on autoregression
 \cite{roberts_hierarchical_2018}. This model is capable of generating
@@
 sequence of discrete tokens with relative positional dependencies
-\cite{huang_music_2018}. This model was trained on the MAESTRO dataset
-of Piano-e-competition performances \cite{hawthorne2019enabling} and
-its generated piano music received favorable qualitative (Likert
-scale) ratings from human listeners \cite{huang_music_2018}.
+\cite{huang_music_2018}. The focus of this model is on learning
+long-term dependencies in music to produce longer clips of music with
+coherent structure. Music Transformer was trained on a dataset of
+Piano-e-competition performances \cite{hawthorne2019enabling} and its
+generated piano music received favorable qualitative (Likert scale)
+ratings from human listeners for its resemblance to human-composed
+music \cite{huang_music_2018}.
 
-MuseGAN \cite{dong2017musegan} is an application of Generative Adversarial
-Networks to polyphonic MIDI music generation, trained on four-bar phrases of a
-multi-track pianoroll representation of rock songs from the Lakh Midi Dataset
-\cite{raffel_learning-based_2016}.
+MuseGAN \cite{dong2017musegan} is an application of Generative
+Adversarial Networks (GAN) to polyphonic MIDI music generation,
+trained on four-bar phrases of a multi-track pianoroll representation
+of rock songs from the Lakh Midi Dataset
+\cite{raffel_learning-based_2016}. Like MusicVAE, MuseGAN includes a
+two-level generator that first samples latent codes at the phrase or
+bar level, then generates notes within the bars, to produce
+longer-term structural patterns.
 
-A major advantage of working with the symbolic representation of music is that
-it is of far lower dimensionality than the raw audio waveforms of a recorded
-performance, which makes it less computationally expensive. However, there are
-many aspects of musical performance that are not captured by a symbolic
-representation, so the expressiveness of symbolic generative models is
-constrained \cite{manzelli_conditioning_2018}.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{song_structure.png}
+  \caption{Diagram of the hierarchical structure of a musical
+    composition, from the MuseGAN paper \cite{dong2017musegan}.}
+  \label{song_structure}
+  \Description{A diagram showing the hierarchical structure of songs,
+    to paragraphs, to phrases, to bars, to, beats, to pixels.}
+\end{figure}
+
+A major advantage of working with the symbolic representation of music
+is that it is of far lower dimensionality than the raw audio waveforms
+of a recorded performance, which makes it less computationally
+expensive. However, there are many aspects of musical performance
+style that are not captured by a symbolic representation, so the
+expressiveness of symbolic generative models is limited
+\cite{manzelli_conditioning_2018}.
 
@@
 
-Jukebox by OpenAI utilizes a vector-quantized variational autoencoder
-(VQ-VAE) to compress raw audio into a sequence of discrete codes and
-models these sequences using autoregressive transformers to generate
-music of various popular genres, including singing and lyrics
-\cite{dhariwal2020jukebox}.
-
-Audio data can also be modeled in the frequency domain through the use of
-Fourier analysis. The recent MelNet model is trained on spectrograms and can
-learn musical structures such as melody and harmony and variations in volume,
-timbre and rhythm \cite{vasquez2019melnet}.
-
 Prior work points out that the division between symbolic music notes
-and the sounds of music is analogous to the division between symbolic
+and music performances, is analogous to the division between symbolic
 language and utterances in speech, which may inspire ideas for
@@
 symbolic music using MIDI data as inputs, because of the advantages of
-symbolic music models in representing long-term structure in musical
+symbolic music models in representing long-range patterns in musical
 compositions to produce generations with coherent structure and use of
@@
 Commons licensed MIDI files of popular music songs that have been
-matched and aligned to MP3 recordings, and of which we will use a
+matched and aligned to MP3 recordings, and of which I plan to use a
 subset for model training and evaluation
@@
 into training examples for a neural network. There are multiple
-open-source Python packages capable of reading MIDI files into NumPy
-array-based representations, such as pretty\_midi
-\cite{raffel_pretty_midi_2014} and Pypianoroll
-\cite{dong_pypianoroll_2018}. In order to accommodate polyphonic
-music, we will convert each MIDI file into a pianoroll representation,
-as visualized in Figure \ref{pianoroll}, wherein each instrument track
-is a sparse matrix that multi-hot encodes the velocity values for each
-of 128 possible pitch levels at each timestep. Because songs are
-typically at least a few minutes long and of varying length, it will
-not be feasible to train with entire songs as examples, so we will
-crop songs into phrases of equal numbers of measures to use as
-training data.
+open-source Python packages that assist with the process of reading
+MIDI files from their binary on-disk representations into Python
+objects, such as pretty\_midi \cite{raffel_pretty_midi_2014} and
+Pypianoroll \cite{dong_pypianoroll_2018}.
+
+In order to accommodate polyphonic music, I will convert each MIDI
+file into a pianoroll representation, as visualized in Figure
+\ref{pianoroll}, wherein each instrument track is a sparse matrix that
+multi-hot encodes the velocity values for each of 128 possible pitch
+levels at each timestep. To reduce dimensionality, I will clip the
+note pitch values to a narrower range, excluding the rarely used notes
+in the extreme high and low octaves, and encode chords (combinations
+of notes played simultaneously on the same instrument) as distinct
+tokens, so that
+
+Because songs are typically at least a few minutes long and of varying
+length, it will not be feasible to train with entire songs as
+examples, so we will crop songs into phrases of equal numbers of
+measures to use as training data.
 
@@
 
+While I plan to model music with multiple instrument tracks, I
+anticipate the need to model only a fixed selection of instrument
+parts, similar to how MusicVAE models three-part (drum, bass and
+melody) \cite{roberts_hierarchical_2018} and MuseGAN models five-part
+(drum, bass, guitar, string, piano) arrangements.
+
 \begin{figure}[h]
   \centering
-  \includegraphics[width=\linewidth]{first_four_bars.png}
-  \caption{Pianoroll visualization of the first four measures of
+  \includegraphics[width=\linewidth]{first_bar.png}
+  \caption{Pianoroll visualization of the first measure of
     Beethoven's Serioso String Quartet}
@@
 
-Data augmentation is also possible and we plan to test its impact on
-our results--the literature suggests augmentation via pitch shifting
-the entire training example up or down by up to six semitones, and
+Data augmentation is also possible and I plan to test its impact on
+the results--the literature suggests augmentation via pitch shifting
+each training example up or down by up to six semitones, and
 increasing or reducing the speed by up to 10\% in order to create
@@
 
-\subsection{Model Fitting}
+\subsection{Model Design}
 
-We will explore several modeling approaches to generating symbolic music:
+I plan to utilize recurrent neural networks with LSTM or GRU cells to
+model the sequential structure of music, stacking multiple layers to
+allow the network to learn at multiple levels of abstraction.
 
-\begin{itemize}
-  \item{Sliding window sequence prediction with RNNs (LSTM/GRU)}
-  \item{Sliding window sequence prediction with Transformers}
-  \item{Latent space interpolation with Sequential VAEs}
-  \item{Latent space interpolation with Sequential GANs}
-\end{itemize}
+There are two general methods of generating sequences from a model:
+sliding window prediction, where a model is trained on a fixed length
+sequence and then used to predict the next item in the sequence
+repeatedly; and latent space interpolation, where a model learns
+random distributions of latent code embeddings for the sequences and
+draws from the distributions to generate new samples. I plan to
+attempt both of these approaches and compare the results.
 
@@
 pitches and note lengths, then the Kullback-Leibler divergence and
-overlapping area of the probability density functions are use to
+overlapping area of the probability density functions are used to
 compare against known reference distributions per musical genre
@@
 associated with designing a human subjects experiment, we plan to
-utilize this quantitative approach to generation quality assessment.
+utilize this quantitative approach to the quality assessment of
+generated samples.
 
@@
 
-At this stage, we have decided on the modeling problem, downloaded the
-MIDI datasets, and begun exploring them to learn how to interpret,
-process, and visualize them. We have started writing a training data
-processing pipeline that will provide batches of training examples to
-Keras model layers for fitting.
+At this stage, I have decided on MIDI generation as the modeling
+problem, downloaded several MIDI datasets, and begun exploring them to
+learn how to interpret, process, and visualize them.
+
+I have implemented several pianoroll data processing functions that
+crop sections of MIDI files into fixed-length training examples
+consisting of multi-track note sequences in symbolic time (beats) at a
+given time resolution. I have started implementing a training data
+processing pipeline that will read MIDI files from disk and transform
+them into the pianoroll-style tensor representation, providing batches
+of training examples to Keras model layers for fitting.
+
+I have started designing a simple LSTM autoencoder to model short
+(e.g. 16 beats or 4 bars) clips of monophonic (single instrument
+track) music, in order to learn the basics of sequence model
+architecture. I will create a sequence predictor model with a similar
+architecture and compare it to the autoencoder. Next I will add
+support for polyphonic music with 3-5 instrument tracks. Once
+polyphonic music generation is working well, I plan to increase the
+length of the sequences and make adjustments to the model architecture
+to permit it to model longer phrases and paragraphs of music.
+
+I aim to build upon the hierarchical decoding approach taken in the
+MusicVAE \cite{roberts_hierarchical_2018} and MuseGAN
+\cite{dong2017musegan} papers, possibly adding more than two recurrent
+layers in the decoder, to address a key challenge in generative music
+modeling: the ability to learn the layered patterns in music to
+produce compositions with convincing long-term structure.
 

### 2021-05-04 notebooks/working_with_midi.py

@@ -125,9 +125,7 @@ def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
 
 # %%
 def bars(multitrack, start_index, num_bars, beats_per_bar, resolution):
-    start, end = get_bar_bounds(
-        start_index, num_bars, beats_per_bar, resolution
-    )
+    start, end = get_bar_bounds(start_index, num_bars, beats_per_bar, resolution)
     tracks = []
     for track in multitrack.tracks:
         tracks.append(
@@ -141,16 +139,19 @@ def bars(multitrack, start_index, num_bars, beats_per_bar, resolution):
 
 
 # %% [markdown]
-# Plot four bars of Beethoven:
+# Plot a bar of Beethoven:
+
 # %%
 resolution = 24
-four_bars = bars(multitrack, 0, 4, 4, resolution)
-four_bars.plot()
-fig = plt.gcf()
-for ax in fig.axes:
-    ax.set_ylim(24, 96)
+fig, axes = plt.subplots(4, sharex=True, sharey=True)
+first_bar = bars(multitrack, 0, 1, 4, resolution)
+first_bar.plot(axes)
+for ax in axes:
+    ax.set_ylim(24, 72)
+plt.savefig("../papers/progress/alex/first_bar.png")
+
 # %%
-viola_track = four_bars.tracks[2]
+viola_track = first_bar.tracks[2]
 viola_track.plot()
 ax = plt.gca()
 ax.set_title("Viola Track")
@@ -226,9 +227,7 @@ model.add(
 )
 model.add(layers.RepeatVector(n_timesteps))
 model.add(layers.LSTM(256, activation="relu", return_sequences=True))
-model.add(
-    layers.TimeDistributed(layers.Dense(n_features, activation="softmax"))
-)
+model.add(layers.TimeDistributed(layers.Dense(n_features, activation="softmax")))
 model.compile(optimizer="adam", loss="categorical_crossentropy")
 utils.plot_model(model, show_shapes=True)
 

### 2021-05-04 papers/progress/jack/progress-jack.tex
Update to Jack's report

@@
   \includegraphics[width=\linewidth]{mel_spec.png}
-  \caption{Mel spectrogram of a 10 second sound segment from the dataset. The color represents the amplitude in dB.}
+  \caption{Mel spectrogram of a 10 second sound segment from the dataset. The color represents the magnitude in dB.}
   \label{spectrogram}
@@
 \section{Progress}
-Initially, a lot of time was spent on understanding different representation of the input data. We also learned 
-about the basics of sequence models such as RNN, GRU, and LSTM. Because the dataset is large and cannot be loaded into 
-the memory at once, we are working on a data pipeline that processes the raw audio files as described in the Section 
-Data Processing and presents training examples one by one to a model.
+Initially, a lot of time was spent on understanding different representations of the input data and on selecting a
+viable representation that could be used to train a model. The key is that we can also reliably convert this
+representation back to a sound wave. A number of representations are available: the raw waveform, the complex numbers,
+the FFT of the waveform, the spectrogram, the mel spectrogram, and the mel frequency ceptral coefficients (MFCC). We decided
+to use the mel spectrogram because it is a compressed version of the sound wave, which could be reliably convert back to 
+audio. We also learned about the basics of sequence models such as RNN, GRU, and LSTM. Because the dataset is large and 
+cannot be loaded into the memory at once, we are working on a data pipeline that processes the raw audio files and presents training examples 
+one by one to a model. This pipeline will slice the input files into short audio fragments to be used as input. Then,
+it will process the fragments as described in Section Data Processing. The resulting spectrograms will be save
+to disk. The training code will load them into memory as needed to train the model.
 

### 2021-05-05 papers/progress/alex/progress-alex.tex
playing with an LSTM-AE model for viola track

@@
 %% \documentclass[manuscript,screen,review]{acmart}
-
+\copyrightyear{2021}
+\acmYear{2021}
 \begin{document}
@@
   deep learning and provides a progress report on a course project to
-  train and evaluate generative models of classical and popular music
-  on the MusicNet and Lakh MIDI datasets.
+  train and evaluate generative models of classical and popular music.
 \end{abstract}
@@
 MIDI representations into audio signals. A generative music model can
-be either trained to generate raw audio, or trained to produce a
-symbolic score that must be played by either a synthesizer or by
-humans to produce an audio music performance. This project focuses on
-generative modeling of symbolic (MIDI) music to generate original
-musical compositions.
+be trained either to generate raw audio, or to produce a symbolic
+score that must be played by a synthesizer or by humans to produce an
+audio music performance. This project focuses on the latter type of
+modeling: generative modeling of symbolic (MIDI) music to compose
+original musical scores.
 
@@
 model learns to ignore its latent code and rely on autoregression
-\cite{roberts_hierarchical_2018}. This model is capable of generating
-new melodies that blend two given melodies via latent space
-interpolation.
+\cite{roberts_hierarchical_2018}. This model is trained on 16-bar
+paragraphs of music and is capable of generating new melodies that
+blend two given melodies via latent space interpolation.
 
@@
 of a recorded performance, which makes it less computationally
-expensive. However, there are many aspects of musical performance
-style that are not captured by a symbolic representation, so the
-expressiveness of symbolic generative models is limited
+expensive. However, there are many stylistic aspects of musical
+performance that are not captured by a symbolic representation, and
+may be specific to a particular performer, so the expressiveness of
+symbolic generative models is limited in comparison
 \cite{manzelli_conditioning_2018}.
@@
 \cite{raffel_learning-based_2016}. Including this second dataset may
-help us to generalize the modeling approach beyond European classical
+help to generalize the modeling approach beyond European classical
 music to include other popular genres and associated instruments.
@@
 of notes played simultaneously on the same instrument) as distinct
-tokens, so that
+tokens, so that they can be one-hot encoded
 
@@
 \cite{yang_evaluation_2020}. Due to the cost and time requirements
-associated with designing a human subjects experiment, we plan to
+associated with designing a human subjects experiment, I plan to
 utilize this quantitative approach to the quality assessment of
@@
 problem, downloaded several MIDI datasets, and begun exploring them to
-learn how to interpret, process, and visualize them.
+learn how to interpret, process, and visualize them. Due to the
+complexity of the MIDI format and the multiple options for
+representing it in memory, data representation has been a
+time-consuming portion of the work in this project.
 
-I have implemented several pianoroll data processing functions that
-crop sections of MIDI files into fixed-length training examples
-consisting of multi-track note sequences in symbolic time (beats) at a
-given time resolution. I have started implementing a training data
-processing pipeline that will read MIDI files from disk and transform
-them into the pianoroll-style tensor representation, providing batches
-of training examples to Keras model layers for fitting.
+So far I have implemented in Python several pianoroll data processing
+functions that crop sections of MIDI files into fixed-length training
+examples consisting of multi-track note sequences in symbolic time
+(beats) at a given time resolution. I have also started implementing a
+training data processing pipeline that will read MIDI files from disk
+and transform them into the pianoroll-style tensor representation,
+providing batches of training examples to Keras model layers for
+fitting.
 
-I have started designing a simple LSTM autoencoder to model short
+I have begun designing a simple LSTM autoencoder to model short
 (e.g. 16 beats or 4 bars) clips of monophonic (single instrument
 track) music, in order to learn the basics of sequence model
-architecture. I will create a sequence predictor model with a similar
-architecture and compare it to the autoencoder. Next I will add
+architecture and determine how complex the model must be in order to
+be capable of reconstructing music note sequences. I will create a
+sequence predictor model with a similar structure and compare its
+generative capability to that of the autoencoder. Next I will add
 support for polyphonic music with 3-5 instrument tracks. Once
@@
 length of the sequences and make adjustments to the model architecture
-to permit it to model longer phrases and paragraphs of music.
-
-I aim to build upon the hierarchical decoding approach taken in the
-MusicVAE \cite{roberts_hierarchical_2018} and MuseGAN
-\cite{dong2017musegan} papers, possibly adding more than two recurrent
-layers in the decoder, to address a key challenge in generative music
-modeling: the ability to learn the layered patterns in music to
-produce compositions with convincing long-term structure.
+to permit it to model longer phrases and paragraphs of music. I aim to
+improve upon the hierarchical decoding approach taken in the MusicVAE
+\cite{roberts_hierarchical_2018} and MuseGAN \cite{dong2017musegan}
+papers, possibly adding more than two recurrent layers in the decoder
+and providing even longer training examples, to address a key
+challenge in generative music modeling: the ability to learn from the
+layered patterns in music to produce compositions with convincing
+long-term structure. A stretch goal that would significantly
+differentiate my work is to produce the first model capable of
+generating not just separate phrases of symbolic music, but a coherent
+song from beginning to end.
 

### 2021-05-05 notebooks/midi_model_playground.py

@@ -1,0 +1,169 @@
+# %% [markdown]
+# # MIDI Model Playground
+#
+# An exploratory notebook for playing with model architecture ideas.
+# %%
+import pathlib
+import sys
+import numpy as np
+import pypianoroll
+import matplotlib.pyplot as plt
+from tensorflow.keras import utils, layers, Model, optimizers, Sequential
+
+sys.path.append("..")
+from musiclearn import config
+
+
+# %%
+# Picking one song from MusicNet dataset
+midi_dir = pathlib.Path(config.MUSICNET_MIDI_DIR)
+mid_2494 = midi_dir / "Beethoven" / "2494_qt11_1.mid"
+multitrack = pypianoroll.read(mid_2494, resolution=24)
+
+# %% [markdown]
+#
+# ## Training data representation
+#
+# Because an entire song is too long music is generally composed in bars or
+# multi-bar phrases, we plan to read MIDI files and segment them into phrases,
+# one training example per phrase.
+
+# %%
+def get_num_beats(multitrack, resolution):
+    return len(multitrack.downbeat) // resolution
+
+
+def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
+    start = bar_index * resolution
+    end = start + (num_bars * beats_per_bar) * resolution
+    return (start, end)
+
+
+# %%
+def bars(multitrack, start_index, num_bars, beats_per_bar, resolution):
+    start, end = get_bar_bounds(
+        start_index, num_bars, beats_per_bar, resolution
+    )
+    tracks = []
+    for track in multitrack.tracks:
+        tracks.append(
+            pypianoroll.Track(
+                name=track.name,
+                program=track.program,
+                pianoroll=track[start:end],
+            )
+        )
+    return pypianoroll.Multitrack(tracks=tracks, resolution=resolution)
+
+
+# %% [markdown]
+# Plot a bar of Beethoven:
+
+# %%
+resolution = 24
+fig, axes = plt.subplots(4, sharex=True, sharey=True)
+first_bar = bars(multitrack, 0, 1, 4, resolution)
+first_bar.plot(axes)
+for ax in axes:
+    ax.set_ylim(24, 72)
+plt.savefig("../papers/progress/alex/first_bar.png")
+
+# %%
+viola_track = first_bar.tracks[2]
+viola_track.plot()
[…]

### 2021-05-05 notebooks/working_with_midi.py

@@ -125,7 +125,9 @@ def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
 
 # %%
 def bars(multitrack, start_index, num_bars, beats_per_bar, resolution):
-    start, end = get_bar_bounds(start_index, num_bars, beats_per_bar, resolution)
+    start, end = get_bar_bounds(
+        start_index, num_bars, beats_per_bar, resolution
+    )
     tracks = []
     for track in multitrack.tracks:
         tracks.append(
@@ -156,88 +158,3 @@ viola_track.plot()
 ax = plt.gca()
 ax.set_title("Viola Track")
 ax.set_ylim(24, 96)
-
-# %% [markdown]
-#
-# ## Pianoroll sequence autoencoder model
-#
-# Testing out training a simple autoencoder model on a single track.
-
-# %%
-# Prepare training examples from the binarized cello track
-resolution = 12
-beats_per_bar = 4
-bars_per_phrase = 2
-total_bars = get_num_beats(multitrack, resolution) // beats_per_bar
-total_phrases = total_bars // bars_per_phrase
-
-# %%
-viola_clips = [
-    bars(
-        multitrack,
-        i * bars_per_phrase,
-        bars_per_phrase,
-        beats_per_bar,
-        resolution,
-    )
-    .tracks[2]
-    .binarize()
-    .pianoroll.astype(int)
-    for i in range(total_phrases)
-]
-np_viola = np.array(viola_clips)
-
-# %%
-np_viola
-
-# %% [markdown]
-#
-# The viola only plays notes 48-70, notes outside this range are all zeroes
-# We can trim the note space to reduce dimensionality and avoid the model trying
-# to fit notes outside the viola's range.
-# %%
-np_viola.sum(axis=(0, 1))
-
-# %%
-np_viola = np_viola[:, :, 48:70]
-np_viola.sum(axis=(0, 1))
-
-# %%
-# make the features ordinal
-np_viola = np.argmax(np_viola, axis=2).astype(float)
-np_viola = np_viola.reshape(np_viola.shape[0], np_viola.shape[1], 1)
-
-# %%
-np_viola
-# %%
-from tensorflow.keras import utils, layers, Model, optimizers, Sequential
-
-# %%
-n_timesteps = np_viola.shape[1]
-n_features = np_viola.shape[2]
-
-# %%
-model = Sequential()
-model.add(
-    layers.LSTM(
-        256,
[…]

### 2021-05-05 notebooks/midi_model_playground.py
playing with lstm-ae

@@ -41,9 +41,7 @@ def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
 
 # %%
 def bars(multitrack, start_index, num_bars, beats_per_bar, resolution):
-    start, end = get_bar_bounds(
-        start_index, num_bars, beats_per_bar, resolution
-    )
+    start, end = get_bar_bounds(start_index, num_bars, beats_per_bar, resolution)
     tracks = []
     for track in multitrack.tracks:
         tracks.append(
@@ -61,19 +59,9 @@ def bars(multitrack, start_index, num_bars, beats_per_bar, resolution):
 
 # %%
 resolution = 24
-fig, axes = plt.subplots(4, sharex=True, sharey=True)
 first_bar = bars(multitrack, 0, 1, 4, resolution)
-first_bar.plot(axes)
-for ax in axes:
-    ax.set_ylim(24, 72)
-plt.savefig("../papers/progress/alex/first_bar.png")
-
 # %%
 viola_track = first_bar.tracks[2]
-viola_track.plot()
-ax = plt.gca()
-ax.set_title("Viola Track")
-ax.set_ylim(24, 96)
 
 # %% [markdown]
 #
@@ -136,34 +124,38 @@ print(shape)
 
 # %%
 opt = optimizers.Adam(learning_rate=0.0001)
-# opt = optimizers.RMSprop(learning_rate=0.0001)
-# opt = optimizers.SGD()
+act = "tanh"
 model = Sequential()
 model.add(
     layers.LSTM(
-        128,
+        512,
         activation="tanh",
         input_shape=(n_timesteps, n_features),
         return_sequences=True,
     )
 )
-model.add(layers.LSTM(64, activation="tanh"))
-model.add(layers.RepeatVector(n_timesteps))
-model.add(layers.LSTM(64, activation="tanh", return_sequences=True))
-model.add(layers.LSTM(128, activation="tanh", return_sequences=True))
-model.add(
-    layers.TimeDistributed(layers.Dense(n_features, activation="softmax"))
-)
+model.add(layers.LSTM(128, activation=act, return_sequences=True))
+model.add(layers.LSTM(128, activation=act, return_sequences=True))
+model.add(layers.LSTM(512, activation=act, return_sequences=True))
+model.add(layers.TimeDistributed(layers.Dense(n_features, activation="softmax")))
 model.compile(
[…]

### 2021-05-15 papers/final/paper.tex
transfer some writing & bibliography to final paper

@@
 \begin{abstract}
-  The abstract will go here.
+  This paper presents a generative model of multi-instrument symbolic
+  music.
 \end{abstract}
@@
 
-Example citation \cite{huang_music_2018}.
+Machine learning models of music have interesting applications in
+music information retrieval and creative tools for musical artists and
+educators. Generative models can create accompaniments for music,
+transfer styles between clips of music, and even generate entirely new
+music. Music is challenging to model because it exhibits a complex
+hierarchy of recurring patterns and long-range temporal dependencies,
+and because both musical scores and performances have multiple
+possible digital representations.
+
+Depending on the task, machine learning models of music may be trained
+on the audio signal itself, either in a time domain or a frequency
+domain representation, or they may be trained on a digital symbolic
+representation of music, the most common of which is MIDI (Musical
+Instrument Digital Interface) notation. MIDI is an encoding of music
+as streams of bytes in one or more tracks or channels, each
+representing a sequence of 128 possible pitch values (where 0 is the
+lowest and 127 is the highest), along with timing, pressure and
+instrument identifier values. A music transcription model may
+transcribe an audio signal as a MIDI score, which can easily be
+converted into other symbolic representations such as sheet music for
+human performers to read from, while a synthesizer model can convert
+MIDI representations into audio signals. A generative music model can
+be trained either to generate raw audio, or to produce a symbolic
+score that must be played by a synthesizer or by humans to produce an
+audio music performance. This project focuses on the latter type of
+modeling: generative modeling of symbolic (MIDI) music to compose
+original musical scores.
+
+\section{Related Work}
+
+The state of the art in music generation has a long way to go before
+it can consistently generate music scores or performances that would
+be enjoyable and popular for humans to listen to, but a number of
+recent research projects have shown promising progress in this area.
+
+Google's \href{https://magenta.tensorflow.org/}{Magenta} is an
+umbrella project for music deep learning research and development of
+software tools to expose these models for use by creative artists and
+students.
+
+MusicVAE, part of the Magenta project, is a variational Long
+Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
+novel hierarchical structure using a ``conductor'' recurrent layer in
+its decoder model to better capture structure at multiple levels and
+avoid the problem of ``posterior/mode collapse'' whereby a generative
+model learns to ignore its latent code and rely on autoregression
+\cite{roberts_hierarchical_2018}. This model is trained on 16-bar
+paragraphs of music and is capable of generating new melodies that
+blend two given melodies via latent space interpolation.
+
+Another Magenta model called Music Transformer is a generative model
+that borrows its approach from the Natural Language Processing (NLP)
+domain, using a self-attention network to model MIDI music as a
+sequence of discrete tokens with relative positional dependencies
+\cite{huang_music_2018}. The focus of this model is on learning
+long-term dependencies in music to produce longer clips of music with
+coherent structure. Music Transformer was trained on a dataset of
+Piano-e-competition performances \cite{hawthorne2019enabling} and its
+generated piano music received favorable qualitative (Likert scale)
+ratings from human listeners for its resemblance to human-composed
+music \cite{huang_music_2018}.
+
+MuseGAN \cite{dong2017musegan} is an application of Generative
+Adversarial Networks (GAN) to polyphonic MIDI music generation,
+trained on four-bar phrases of a multi-track pianoroll representation
+of rock songs from the Lakh Midi Dataset
+\cite{raffel_learning-based_2016}. Like MusicVAE, MuseGAN includes a
+two-level generator that first samples latent codes at the phrase or
+bar level, then generates notes within the bars, to produce
+longer-term structural patterns.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{song_structure.png}
+  \caption{Diagram of the hierarchical structure of a musical
+    composition, from the MuseGAN paper \cite{dong2017musegan}.}
+  \label{song_structure}
+  \Description{A diagram showing the hierarchical structure of songs,
+    to paragraphs, to phrases, to bars, to, beats, to pixels.}
+\end{figure}
+
+A major advantage of working with the symbolic representation of music
+is that it is of far lower dimensionality than the raw audio waveforms
+of a recorded performance, which makes it less computationally
+expensive. However, there are many stylistic aspects of musical
+performance that are not captured by a symbolic representation, and
+may be specific to a particular performer, so the expressiveness of
+symbolic generative models is limited in comparison
+\cite{manzelli_conditioning_2018}.
+
+Other research has focused on modeling raw audio waveforms directly. WaveNet is
+a causal convolutional neural network for generating raw audio waveforms,
+developed by Google DeepMind, which achieves state of the art performance in
+generating natural sounding speech from text, but is also capable of generating
+short, realistic snippets of audio music \cite{oord_wavenet_2016}.
+Another model named SampleRNN generates raw audio waveforms using a three-tier
+hierarchy of gated recurrent units (GRU) to model recurrent structure at
+multiple temporal resolutions \cite{mehri_samplernn_2017}.
+
+Prior work points out that the division between symbolic music notes
+and music performances, is analogous to the division between symbolic
+language and utterances in speech, which may inspire ideas for
+combining the two approaches \cite{hawthorne2019enabling}. A paper
+from Boston University describes an effort to combine the symbolic and
+waveform approaches to music modeling, by training an LSTM to learn
+melodic structure of different styles of music, then providing
+generations from this model as conditioning inputs to a WaveNet-based
+raw audio generator \cite{manzelli_conditioning_2018}.
 
 \section{Methods}
+
+\subsection{Datasets}
+
+Two primary datasets will be used for this research project: MusicNet,
+which is a collection of 330 freely licensed European classical music
+recordings with aligned MIDI scores \cite{thickstun2017learning}, and
+the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
+Commons licensed MIDI files of popular music songs that have been
+matched and aligned to MP3 recordings, and of which I plan to use a
+subset for model training and evaluation
+\cite{raffel_learning-based_2016}. Including this second dataset may
+help to generalize the modeling approach beyond European classical
+music to include other popular genres and associated instruments.
+
+\subsection{Data Preprocessing}
+
+Several choices must be made in how to preprocess binary MIDI files
+into training examples for a neural network. There are multiple
+open-source Python packages that assist with the process of reading
+MIDI files from their binary on-disk representations into Python
+objects, such as pretty\_midi \cite{raffel_pretty_midi_2014} and
+Pypianoroll \cite{dong_pypianoroll_2018}.
+
+In order to accommodate polyphonic music, I will convert each MIDI
+file into a pianoroll representation, as visualized in Figure
+\ref{pianoroll}, wherein each instrument track is a sparse matrix that
+multi-hot encodes the velocity values for each of 128 possible pitch
+levels at each timestep. To reduce dimensionality, I will clip the
+note pitch values to a narrower range, excluding the rarely used notes
+in the extreme high and low octaves, and encode chords (combinations
+of notes played simultaneously on the same instrument) as distinct
+tokens, so that they can be one-hot encoded
+
+Because songs are typically at least a few minutes long and of varying
+length, it will not be feasible to train with entire songs as
+examples, so we will crop songs into phrases of equal numbers of
+measures to use as training data.
+
+The result of this preprocessing is that each training example will be
+a 3D tensor of shape (tracks x ticks x pitches) and stacking the
+training examples will produce a 4D tensor.
+
+While we plan to model music with multiple instrument tracks, we
+anticipate the need to model only a fixed selection of instrument
+parts, similar to how MusicVAE models three-part (drum, bass and
+melody) \cite{roberts_hierarchical_2018} and MuseGAN models five-part
+(drum, bass, guitar, string, piano) arrangements.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{first_bar.png}
+  \caption{Pianoroll visualization of the first measure of
+    Beethoven's Serioso String Quartet}
+  \label{pianoroll}
+  \Description{A pianoroll representation of cello, viola and violin
+    parts from a Beethoven piece.}
+\end{figure}
+
+Data augmentation is also possible and we assess its impact on
+the results--the literature suggests augmentation via pitch shifting
+each training example up or down by up to six semitones, and
+increasing or reducing the speed by up to 10\% in order to create
+additional training examples and reduce overfitting
+\cite{oore_this_2018}.
+
+The multi-stream representation discussed in
+\cite{kumar2019polyphonic} addresses the issue of extreme class
+imbalance and sparsity in the pianoroll representation.
+
+\subsection{Model Design}
+
+\subsection{Model Evaluation}
+
+Evaluation of generative models is challenging because there is no
+equivalent of an accuracy metric like what is used in supervised
+learning. For autoencoder models we can measure how accurately the
+model can reconstruct its own inputs, but this does not tell us the
+quality of the interpolated examples. Generative models are typically
+evaluated using a combination of qualitative metrics whereby human
+judges rate the quality of the generated examples (essentially a
+Turing test), and quantitative metrics that assess the differences in
+the parametric distributions of generated and real examples. Yang and
+Lerch (2020) proposes a set of metrics informed by music theory, for
+probabilistically evaluating how similar the generations are to known
+sample distributions of real music \cite{yang_evaluation_2020}. These
+metrics include counts, ranges, histograms and transition matrices of
+pitches and note lengths, then the Kullback-Leibler divergence and
+overlapping area of the probability density functions are used to
+compare against known reference distributions per musical genre
+\cite{yang_evaluation_2020}. Due to the cost and time requirements
+associated with designing a human subjects experiment, we
+utilize this quantitative approach to the quality assessment of
+generated samples.
 

### 2021-05-15 notebooks/midi_quartet_model.py

@@ -7,6 +7,10 @@
 # An exploratory notebook for playing with model architecture ideas.
 
 # %%
+%load_ext autoreload
+%autoreload 2
+
+# %%
 import pathlib
 import os
 import sys

### 2021-05-15 papers/final/paper.tex
save paper

@@
 MIDI files from their binary on-disk representations into Python
-objects, such as pretty\_midi \cite{raffel_pretty_midi_2014} and
-Pypianoroll \cite{dong_pypianoroll_2018}.
+objects, such as pretty\_midi \cite{raffel_pretty_midi_2014},
+Pypianoroll \cite{dong_pypianoroll_2018} and music21
+\cite{cuthbert_music21_2010}.
 

### 2021-05-17 papers/final/paper.tex
some music21 processing functions

@@
 Commons licensed MIDI files of popular music songs that have been
-matched and aligned to MP3 recordings, and of which I plan to use a
+matched and aligned to MP3 recordings, and of which we plan to use a
 subset for model training and evaluation
@@
 
-In order to accommodate polyphonic music, I will convert each MIDI
-file into a pianoroll representation, as visualized in Figure
+In order to accommodate polyphonic music, we convert each MIDI file
+into a pianoroll representation, as visualized in Figure
 \ref{pianoroll}, wherein each instrument track is a sparse matrix that

### 2021-05-17 musiclearn/processing.py

@@ -13,6 +13,7 @@ from scipy.io import wavfile
 import numpy as np
 import pypianoroll
 from musiclearn import config
+from music21 import converter, instrument, note, stream, chord
 
 
 class WavDataGenerator(keras.utils.Sequence):
@@ -241,7 +242,7 @@ def list_musicnet_instruments(path: os.PathLike) -> pd.DataFrame:
 def multitracks_by_instruments(
     path: os.PathLike,
     program_ids: List[int],
-    resolution: int = 24,
+    resolution: int,
 ) -> Tuple[List[pypianoroll.Multitrack], List[str]]:
     """Get a list of multitracks with the specified program IDs (instrument #s) from a directory"""
     if not os.path.isdir(path):
@@ -266,7 +267,7 @@ def multitracks_by_instruments(
 
 
 def multitracks_by_instruments_musicnet(
-    path: os.PathLike, program_ids: List[int]
+    path: os.PathLike, program_ids: List[int], resolution: int
 ) -> Tuple[List[pypianoroll.Multitrack], List[str]]:
     """Get a list of musicnet multitracks with the specified program IDs"""
     if not os.path.isdir(path):
@@ -277,7 +278,7 @@ def multitracks_by_instruments_musicnet(
     fnames = []
     for composer in composers:
         composer_tracks, composer_files = multitracks_by_instruments(
-            path / composer, program_ids
+            path / composer, program_ids, resolution
         )
         if len(composer_tracks) > 0:
             multitracks.extend(composer_tracks)
@@ -285,11 +286,19 @@ def multitracks_by_instruments_musicnet(
     return multitracks, fnames
 
 
-def multitracks_musicnet_quartets() -> List[pypianoroll.Multitrack]:
+def multitracks_musicnet_quartets(resolution: int = 12) -> List[pypianoroll.Multitrack]:
     """Get a pypianoroll multitrack for each string quartet in the MusicNet dataset."""
     path = Path(config.MUSICNET_MIDI_DIR)
     string_quartet = [40, 40, 41, 42]
-    tracks, _ = multitracks_by_instruments_musicnet(path, string_quartet)
+    tracks, _ = multitracks_by_instruments_musicnet(path, string_quartet, resolution)
+    return tracks
+
+
+def multitracks_musicnet_quartets(resolution: int = 12) -> List[pypianoroll.Multitrack]:
[…]

### 2021-05-31 musiclearn/sequential_models.py
Update python files

@@ -49,12 +49,6 @@ def lstm_model(input_shape, n_vocab):
     return model
 
 
-def load_lstm_model():
-    """ Load the saved LSTM model """
-    model = tf.keras.models.load_model('lstm_saved_weights.hdf5')
-    return model
-
-
 def bidirectional_lstm_model(input_shape, n_vocab):
     """ Build a bidirectional LSTM model """
     model = Sequential()
@@ -77,13 +71,6 @@ def bidirectional_lstm_model(input_shape, n_vocab):
     
     return model
 
-
-def load_bidirectional_lstm_model():
-    """ Load the saved LSTM model """
-    
-    model = tf.keras.models.load_model('bidirect_lstm_saved_weights.hdf5')
-    
-    return model
 
 
 class Customized_Attention(Layer):
@@ -150,38 +137,6 @@ def attention_lstm_model(input_shape, n_vocab):
     return model
 
 
-def load_attention_lstm_model(input_shape, n_vocab):
-    """ Load the saved weights """
-    #model = tf.keras.models.load_model("attention_lstm_saved_weight.hdf5")
-    
-    model = Sequential()
-    
-    # Bidirectional LSTM layer with attention
-    model.add(Bidirectional(LSTM(512, return_sequences=True), 
-                            input_shape=(input_shape[1], input_shape[2])
-                           )
-             )
-    model.add(Customized_Attention(return_sequences=True))
-    model.add(Dropout(0.3))
-    
-    # Second bidirectional LSTM layer with attention
-    model.add(Bidirectional(LSTM(512, return_sequences=True)))
-    model.add(Customized_Attention(return_sequences=True))
-    model.add(Dropout(0.3))
-    
-    model.add(Bidirectional(LSTM(512)))
-    model.add(Dense(256))
-    model.add(Dropout(0.3))
-    model.add(Dense(n_vocab))
-    model.add(Activation('softmax'))
-    
-    # compile
-    model.compile(loss='sparse_categorical_crossentropy', optimizer='rmsprop')
-    
-    model.load_weights("attention_lstm_saved_weights.hdf5")
-    return model
-
-
 def simplified_wavenet(n_vocab, sequence_length, input_shape):
     """ A simplified version of WaveNet without residual and skip connection """
     model = Sequential()
@@ -214,14 +169,9 @@ def simplified_wavenet(n_vocab, sequence_length, input_shape):
     return model
 
 
-def load_wavenet_model():
-    """ Load the saved weights of the WaveNet model """
[…]

### 2021-05-31 musiclearn/sequential_models.py
Add commandline interface for training sequential models

@@ -137,7 +137,7 @@ def attention_lstm_model(input_shape, n_vocab):
     return model
 
 
-def simplified_wavenet(n_vocab, sequence_length, input_shape):
+def simplified_wavenet(input_shape, n_vocab):
     """ A simplified version of WaveNet without residual and skip connection """
     model = Sequential()
     
@@ -178,8 +178,8 @@ def train_model(model, sequence_length, model_name, epochs=20, batch_size=128):
     network_input, network_output = prepare_sequences(sequence_length, notes)
     
     # train the network
-    #filepath = model_name + "_saved_weights.hdf5"                       # uncomment later
-    filepath = "test_" + model_name + "_saved_weights.hdf5"
+    cwd = os.getcwd()
+    filepath = cwd + "/" + model_name + "_saved_weights.hdf5"
     
     checkpoint_cb = ModelCheckpoint(
         filepath,

### 2021-06-01 musiclearn/sequential_models.py
Add a commandline interface for generate music from sequential models

@@ -209,11 +209,11 @@ def train_model(model, sequence_length, model_name, epochs=20, batch_size=128):
 
 def load_lstm_model(weights_path:str):
     """ Load the saved LSTM and bidirectional LSTM model """
-    model = tf.keras.models.load_model(weight_path)
+    model = tf.keras.models.load_model(weights_path)
     return model
 
 
-def load_attention_lstm_model(weight_path:str, input_shape, n_vocab):
+def load_attention_lstm_model(weights_path:str, input_shape, n_vocab):
     """ Load the saved weights """
     model = Sequential()
     
@@ -238,17 +238,17 @@ def load_attention_lstm_model(weight_path:str, input_shape, n_vocab):
 
     # compile
     model.compile(loss='sparse_categorical_crossentropy', optimizer='rmsprop')
-    model.load_weights("attention_lstm_saved_weights.hdf5")
+    model.load_weights(weights_path)
     return model
 
 
 def load_wavenet_model(weights_path:str):
     """ Load the saved weights of the WaveNet model """
-    model = tf.keras.models.load_model('wavenet_saved_weights.hdf5')
+    model = tf.keras.models.load_model(weight_path)
     return model
 
 
-def generate_notes(model, network_input, int_to_note, n_vocab):
+def generate_notes(model, network_input, int_to_note, n_vocab, num_notes):
     """ Generate notes from the neural network based on a sequence of notes """
     # pick a random sequence from the input as a starting point for the prediction
     start = np.random.randint(0, len(network_input)-1)
@@ -257,7 +257,7 @@ def generate_notes(model, network_input, int_to_note, n_vocab):
     prediction_output = []
 
     # generate 100 notes
-    for note_index in range(100):
+    for note_index in range(num_notes):
         prediction_input = np.reshape(pattern, (1, len(pattern), 1))
 
         prediction = model.predict(prediction_input, verbose=0)
@@ -273,22 +273,22 @@ def generate_notes(model, network_input, int_to_note, n_vocab):
     return prediction_output
 
 
-def generate_mid_sample(model, file_name):
+def generate_midi_sample(model, data_path, outputFile_name, num_notes):
     """ Generate a new mid sample """
     # get notes from dataset
-    notes = read_midi(schubert_dir)
+    notes = read_midi(data_path)
     
     network_input, _ = prepare_sequences(SEQUENCE_LENGTH, notes)
     
-    int_to_note = map_int_to_note(notes)
[…]

### 2021-06-01 musiclearn/single_note_processing.py

@@ -141,7 +141,7 @@ def create_midi(notes_list, filename):
 
     midi_stream = stream.Stream(output_notes)
 
-    midi_stream.write('midi', fp=filename)
+    midi_stream.write('midi', fp=filename + '.mid')
  
 
 

### 2021-06-01 musiclearn/models/vae.py
model saving enhancement

@@ -183,6 +183,7 @@ class MultiTrackVAE:
                 batch_size=self.batch_size,
                 learning_rate=self.learning_rate,
                 trained_epochs=self.trained_epochs,
+                rest_code=self.rest_code,
             )
             joblib.dump(train_state, directory / "train_state.joblib")
         joblib.dump(hparams, directory / "hparams.joblib")
@@ -202,6 +203,7 @@ class MultiTrackVAE:
         model.n_tracks = train_state["n_tracks"]
         model.n_timesteps = train_state["n_timesteps"]
         model.n_notes = train_state["n_notes"]
+        model.rest_code = train_state["rest_code"]
         model.batch_size = train_state["batch_size"]
         model.learning_rate = train_state["learning_rate"]
         model.trained_epochs = train_state["trained_epochs"]
@@ -251,9 +253,12 @@ class MultiTrackVAE:
         x = processing.split_array(
             x, beats_per_phrase=beats_per_phrase, resolution=ticks_per_beat, fill=self.rest_code
         )
-        xout = self.vae_model.predict(x)
-        xout = np.vstack(xout)
-        return xout
+        x = self.vae_model.predict(x)
+        x = np.reshape(x, (x.shape[0], x.shape[1], self.n_tracks, self.n_notes))
+        x = np.argmax(x, axis=3)
+        x = np.vstack(x)
+        x = self.ord_enc.inverse_transform(x)
+        return x
 
     def interpolate(self, start, stop, n):
         """Interpolate n samples from the latent space between two inputs."""

### 2021-06-01 musiclearn/processing.py

@@ -6,7 +6,7 @@ import os
 import sys
 from functools import reduce
 from pathlib import Path
-from typing import List, Tuple
+from typing import Any, List, Tuple
 
 import joblib
 import numpy as np
@@ -436,8 +436,12 @@ def score_to_str_array(score: stream.Score, resolution: int = 12) -> np.array:
     return arr
 
 
-def str_array_to_score(
-    arr: np.array, programs: List[int] = None, resolution: int = 12
+def array_to_score(
+    arr: np.array,
+    programs: List[int] = None,
+    resolution: int = 12,
+    rest: Any = REST,
+    sustain: Any = SUSTAIN,
 ) -> stream.Score:
     """Convert str array back into a score so we can output MIDI.
     Parameters
@@ -459,13 +463,13 @@ def str_array_to_score(
         inst = instrument.instrumentFromMidiProgram(mid_program)
         part.insert(0, inst)
         dfp = df[[p, "offset"]]
-        dfp = dfp[dfp[p] != SUSTAIN_STR]
+        dfp = dfp[dfp[p] != sustain]
         dfp["duration"] = -dfp["offset"].diff(-1)
         dfp["duration"] = dfp["duration"].fillna(0)
         for i, row in dfp.iterrows():
-            if row[p] == REST_STR:  # rest
+            if row[p] == rest:
                 part.append(note.Rest(quarterLength=row["duration"]))
-            elif "." in row[p]:  # chord
+            elif isinstance(row[p], str) and "." in row[p]:  # chord
                 part.append(chord.Chord(str_to_chord(row[p]), quarterLength=row["duration"]))
             else:  # note
                 part.append(note.Note(row[p], quarterLength=row["duration"]))

### 2021-06-01 musiclearn/training.py

@@ -72,7 +72,7 @@ def train_mtvae(
     model_name = type(mtvae).__name__
     start_time = datetime.now()
     start_str = start_time.strftime("%Y-%m-%dT%H:%M:%S")
-    exp_path = Path(f"experiments/{exp_name}/{start_str})")
+    exp_path = Path(f"experiments/{exp_name}/{start_str}")
     cbacks = build_callbacks(exp_path, patience=patience)
     log_start(
         exp_name,
@@ -92,5 +92,5 @@ def train_mtvae(
     mtvae.train(x, ticks_per_beat, beats_per_phrase, epochs, callbacks=cbacks)
     log_end(exp_name, model_name, start_time)
     mtvae.save(exp_path / "saved_model")
-    LOG.info(f"Model configuration saved to {exp_path}")
+    LOG.info(f"Model configuration saved to {exp_path / 'saved_model'}")
     return mtvae

### 2021-06-01 run_experiments.sh

@@ -9,4 +9,8 @@
 #python musiclearn_cli.py fit-mtvae mtvae_0003 --beats-per-phrase 16 --gru --embedding-dim=32 --patience 50
 #python musiclearn_cli.py fit-mtvae mtvae_0003 --beats-per-phrase 16 --gru --embedding-dim=48 --patience 50
 #python musiclearn_cli.py fit-mtvae mtvae_0003 --beats-per-phrase 16 --lstm --embedding-dim=48 --patience 50
-python musiclearn_cli.py fit-mtvae mtvae_0004 --beats-per-phrase 16 --lstm-units=256 --embedding-dim=16 --dropout-rate 0.4 --patience 50
+#python musiclearn_cli.py fit-mtvae mtvae_0004 --beats-per-phrase 16 --lstm-units=256 --embedding-dim=16 --dropout-rate 0.4 --patience 50
+#python musiclearn_cli.py fit-mtvae mtvae_0005 --beats-per-phrase 4 --lstm-units 512 --embedding-dim 0 --patience 50
+#python musiclearn_cli.py fit-mtvae mtvae_0006 --beats-per-phrase 4 --lstm-units 256 --latent-dim 1024 --embedding-dim 128 --patience 100
+#python musiclearn_cli.py fit-mtvae mtvae_0007 --beats-per-phrase 4 --lstm-units 512 --latent-dim 1024 --embedding-dim 32 --patience 200
+python musiclearn_cli.py fit-mtvae mtvae_0008 --beats-per-phrase 4 --lstm-units 512 --latent-dim 2048 --embedding-dim 32 --patience 100

### 2021-06-01 musiclearn/models/vae.py
remove concatenate layer from vae model

@@ -109,12 +109,12 @@ def build_multi_track_vae(
     decoder = layers.Dropout(dropout_rate)(decoder)
     decoder = rnn(lstm_units, return_sequences=True)(decoder)
     outputs = [
-        layers.TimeDistributed(layers.Dense(n_notes, activation="softmax"))(decoder)
-        for _ in range(n_tracks)
+        layers.TimeDistributed(layers.Dense(n_notes, activation="softmax", name=f"track_{i}"))(
+            decoder
+        )
+        for i in range(n_tracks)
     ]
-    concat_outputs = layers.Concatenate(axis=2)(outputs)
-    decoder_model = keras.Model(decoder_input, concat_outputs)
-
+    decoder_model = keras.Model(decoder_input, outputs)
     # connect encoder and decoder together
     decoder_outputs = decoder_model(z)
     vae_model = keras.Model(inputs=inputs, outputs=decoder_outputs)
@@ -254,7 +254,7 @@ class MultiTrackVAE:
             x, beats_per_phrase=beats_per_phrase, resolution=ticks_per_beat, fill=self.rest_code
         )
         x = self.vae_model.predict(x)
-        x = np.reshape(x, (x.shape[0], x.shape[1], self.n_tracks, self.n_notes))
+        x = np.stack(x, axis=2)
         x = np.argmax(x, axis=3)
         x = np.vstack(x)
         x = self.ord_enc.inverse_transform(x)
@@ -267,7 +267,7 @@ class MultiTrackVAE:
         new_points = self.decoder_model.predict(axis)
         return new_points
 
-    def generate():
+    def generate(self):
         """TODO: write a function to generate MIDI output"""
 
         raise NotImplementedError()

### 2021-06-01 notebooks/midi_quartet_inference.py

@@ -1,0 +1,52 @@
+# %% [markdown]
+# # Quartet model generation and interpolation
+
+# %%
+import os
+import sys
+from pathlib import Path
+
+sys.path.append("..")
+import numpy as np
+from musiclearn import config, processing
+from musiclearn.models import vae
+
+# %%
+midis = Path(config.MUSICNET_MIDI_DIR)
+beethoven_2494 = midis / "Beethoven" / "2494_qt11_1.mid"
+mozart_1788 = midis / "Mozart" / "1788_kv_465_1.mid"
+
+# %%
+ticks_per_beat = 4
+# beats_per_phrase = 16
+beats_per_phrase = 4
+programs = [40, 40, 41, 42]
+bee = processing.score_to_array(processing.midi_to_music21(beethoven_2494), ticks_per_beat)
+moz = processing.score_to_array(processing.midi_to_music21(mozart_1788), ticks_per_beat)
+
+# %%
+# saved_model_path = "../experiments/mtvae_0005/2021-05-31T15:28:18/saved_model"
+# saved_model_path = "../experiments/mtvae_0005/2021-05-31T16:06:39/saved_model"
+# saved_model_path = "../experiments/mtvae_0006/2021-05-31T16:36:10/saved_model"
+# saved_model_path = "../experiments/mtvae_0007/2021-05-31T16:59:11/saved_model"
+# saved_model_path = "../experiments/mtvae_0008/2021-05-31T17:16:16/saved_model"
+saved_model_path = "../experiments/mtvae_0009/2021-05-31T20:24:32/saved_model"
+model = vae.MultiTrackVAE.from_saved(saved_model_path)
+
+# %% [markdown]
+# Use the model to autoencode / reconstruct an entire piece of music
+
+# %%
+bee_reconst = model.reconstruct(bee, ticks_per_beat, beats_per_phrase)
+bee_reconst_score = processing.array_to_score(
+    bee_reconst, programs=programs, resolution=ticks_per_beat
+)
+
+bee_reconst_score.write("midi", "../outputs/2494_qt11_1_reconst.mid")
+
+# %%
+moz_reconst = model.reconstruct(moz, ticks_per_beat, beats_per_phrase)
+moz_reconst_score = processing.array_to_score(
+    moz_reconst, programs=programs, resolution=ticks_per_beat
+)
+moz_reconst_score.write("midi", "../outputs/1788_kv_465_1_reconst.mid")

### 2021-06-01 run_experiments.sh

@@ -13,4 +13,5 @@
 #python musiclearn_cli.py fit-mtvae mtvae_0005 --beats-per-phrase 4 --lstm-units 512 --embedding-dim 0 --patience 50
 #python musiclearn_cli.py fit-mtvae mtvae_0006 --beats-per-phrase 4 --lstm-units 256 --latent-dim 1024 --embedding-dim 128 --patience 100
 #python musiclearn_cli.py fit-mtvae mtvae_0007 --beats-per-phrase 4 --lstm-units 512 --latent-dim 1024 --embedding-dim 32 --patience 200
-python musiclearn_cli.py fit-mtvae mtvae_0008 --beats-per-phrase 4 --lstm-units 512 --latent-dim 2048 --embedding-dim 32 --patience 100
+#python musiclearn_cli.py fit-mtvae mtvae_0008 --beats-per-phrase 4 --lstm-units 512 --latent-dim 2048 --embedding-dim 32 --patience 100
+python musiclearn_cli.py fit-mtvae mtvae_0009 --beats-per-phrase 4 --lstm-units 128

### 2021-06-01 musiclearn/models/vae.py
add interpolate method

@@ -260,12 +260,31 @@ class MultiTrackVAE:
         x = self.ord_enc.inverse_transform(x)
         return x
 
-    def interpolate(self, start, stop, n):
+    def interpolate(self, start, stop, n, ticks_per_beat, beats_per_phrase):
         """Interpolate n samples from the latent space between two inputs."""
-        space = np.linspace(start, stop, n)
-        axis = space[:, np.newaxis]
-        new_points = self.decoder_model.predict(axis)
-        return new_points
+        start = self.ord_enc.transform(start).astype(int)
+        stop = self.ord_enc.transform(stop).astype(int)
+        start = processing.split_array(
+            start, beats_per_phrase=beats_per_phrase, resolution=ticks_per_beat, fill=self.rest_code
+        )
+        stop = processing.split_array(
+            stop, beats_per_phrase=beats_per_phrase, resolution=ticks_per_beat, fill=self.rest_code
+        )
+        min_len = min(start.shape[0], stop.shape[0])
+        start = start[0:min_len, :, :]
+        stop = stop[0:min_len, :, :]
+        start_mu, start_sigma, start_z = self.encoder_model.predict(start)
+        stop_mu, stop_sigma, stop_z = self.encoder_model.predict(stop)
+        space = np.linspace(start_z, stop_z, 3)
+        results = []
+        for x in space:
+            x = self.decoder_model.predict(x)
+            x = np.stack(x, axis=2)
+            x = np.argmax(x, axis=3)
+            x = np.vstack(x)
+            x = self.ord_enc.inverse_transform(x)
+            results.append(x)
+        return results
 
     def generate(self):
         """TODO: write a function to generate MIDI output"""

### 2021-06-01 notebooks/midi_quartet_inference.py

@@ -18,23 +18,22 @@ mozart_1788 = midis / "Mozart" / "1788_kv_465_1.mid"
 
 # %%
 ticks_per_beat = 4
-# beats_per_phrase = 16
 beats_per_phrase = 4
 programs = [40, 40, 41, 42]
 bee = processing.score_to_array(processing.midi_to_music21(beethoven_2494), ticks_per_beat)
 moz = processing.score_to_array(processing.midi_to_music21(mozart_1788), ticks_per_beat)
 
 # %%
-# saved_model_path = "../experiments/mtvae_0005/2021-05-31T15:28:18/saved_model"
-# saved_model_path = "../experiments/mtvae_0005/2021-05-31T16:06:39/saved_model"
-# saved_model_path = "../experiments/mtvae_0006/2021-05-31T16:36:10/saved_model"
-# saved_model_path = "../experiments/mtvae_0007/2021-05-31T16:59:11/saved_model"
-# saved_model_path = "../experiments/mtvae_0008/2021-05-31T17:16:16/saved_model"
-saved_model_path = "../experiments/mtvae_0009/2021-05-31T20:24:32/saved_model"
+# saved_model_path = "../experiments/mtvae_0009/2021-05-31T20:24:32/saved_model"
+# saved_model_path = "../experiments/mtvae_0013/2021-05-31T22:15:37/saved_model"
+# saved_model_path = "../experiments/mtvae_0015/2021-05-31T22:38:59/saved_model"
+saved_model_path = "../experiments/mtvae_0016/2021-05-31T23:01:43/saved_model"
 model = vae.MultiTrackVAE.from_saved(saved_model_path)
 
 # %% [markdown]
-# Use the model to autoencode / reconstruct an entire piece of music
+# ## Reconstruction
+#
+# Use the model to reconstruct an entire piece of music
 
 # %%
 bee_reconst = model.reconstruct(bee, ticks_per_beat, beats_per_phrase)
@@ -42,11 +41,48 @@ bee_reconst_score = processing.array_to_score(
     bee_reconst, programs=programs, resolution=ticks_per_beat
 )
 
-bee_reconst_score.write("midi", "../outputs/2494_qt11_1_reconst.mid")
+bee_reconst_score.write("midi", "../outputs/2494_qt11_1_reconst_0016.mid")
 
 # %%
 moz_reconst = model.reconstruct(moz, ticks_per_beat, beats_per_phrase)
 moz_reconst_score = processing.array_to_score(
[…]

### 2021-06-01 run_experiments.sh

@@ -14,4 +14,11 @@
 #python musiclearn_cli.py fit-mtvae mtvae_0006 --beats-per-phrase 4 --lstm-units 256 --latent-dim 1024 --embedding-dim 128 --patience 100
 #python musiclearn_cli.py fit-mtvae mtvae_0007 --beats-per-phrase 4 --lstm-units 512 --latent-dim 1024 --embedding-dim 32 --patience 200
 #python musiclearn_cli.py fit-mtvae mtvae_0008 --beats-per-phrase 4 --lstm-units 512 --latent-dim 2048 --embedding-dim 32 --patience 100
-python musiclearn_cli.py fit-mtvae mtvae_0009 --beats-per-phrase 4 --lstm-units 128
+#python musiclearn_cli.py fit-mtvae mtvae_0009 --beats-per-phrase 4 --lstm-units 128
+#python musiclearn_cli.py fit-mtvae mtvae_0010 --beats-per-phrase 16 --lstm-units 128 --latent-dim 512 --embedding-dim 32  --dropout-rate 0.4 --patience 50
+#python musiclearn_cli.py fit-mtvae mtvae_0011 --beats-per-phrase 16 --gru --lstm-units 128 --latent-dim 512 --embedding-dim 32  --dropout-rate 0.4 --patience 50
+#python musiclearn_cli.py fit-mtvae mtvae_0012 --beats-per-phrase 16 --lstm-units 512 --latent-dim 2048 --embedding-dim 8 --dropout-rate 0.5 --patience 20
+#python musiclearn_cli.py fit-mtvae mtvae_0013 --beats-per-phrase 8 --lstm-units 128 --latent-dim 512 --dropout-rate 0.4 --patience 20
+#python musiclearn_cli.py fit-mtvae mtvae_0014 --beats-per-phrase 8 --lstm-units 256 --latent-dim 512 --dropout-rate 0.5 --patience 20
+#python musiclearn_cli.py fit-mtvae mtvae_0015 --beats-per-phrase 8 --lstm-units 128 --latent-dim 128 --dropout-rate 0.5 --patience 20
+python musiclearn_cli.py fit-mtvae mtvae_0016 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20

### 2021-06-01 notebooks/midi_quartet_inference.py
save quartet notebooks

@@ -20,8 +20,11 @@ mozart_1788 = midis / "Mozart" / "1788_kv_465_1.mid"
 ticks_per_beat = 4
 beats_per_phrase = 4
 programs = [40, 40, 41, 42]
-bee = processing.score_to_array(processing.midi_to_music21(beethoven_2494), ticks_per_beat)
-moz = processing.score_to_array(processing.midi_to_music21(mozart_1788), ticks_per_beat)
+bee_score = processing.midi_to_music21(beethoven_2494)
+
+bee = processing.score_to_array(bee_score, ticks_per_beat)
+moz_score = processing.midi_to_music21(mozart_1788)
+moz = processing.score_to_array(moz_score, ticks_per_beat)
 
 # %%
 # saved_model_path = "../experiments/mtvae_0009/2021-05-31T20:24:32/saved_model"
@@ -62,27 +65,3 @@ stop = moz
 beemoz = model.interpolate(start, stop, 3, ticks_per_beat, beats_per_phrase)
 beemoz_score = processing.array_to_score(beemoz[1], programs=programs, resolution=ticks_per_beat)
 beemoz_score.write("midi", "../outputs/beemoz_0016.mid")
-
-# %%
-start = model.ord_enc.transform(start).astype(int)
-stop = model.ord_enc.transform(stop).astype(int)
-start = processing.split_array(
-    start, beats_per_phrase=beats_per_phrase, resolution=ticks_per_beat, fill=model.rest_code
-)
-stop = processing.split_array(
-    stop, beats_per_phrase=beats_per_phrase, resolution=ticks_per_beat, fill=model.rest_code
-)
-min_len = min(start.shape[0], stop.shape[0])
-start = start[0:min_len, :, :]
-stop = stop[0:min_len, :, :]
-start_mu, start_sigma, start_z = model.encoder_model.predict(start)
-stop_mu, stop_sigma, stop_z = model.encoder_model.predict(stop)
-space = np.linspace(start_z, stop_z, 3)
-interpolations = []
-for x in space:
-    x = model.decoder_model.predict(x)
-    x = np.stack(x, axis=2)
-    x = np.argmax(x, axis=3)
-    x = np.vstack(x)
-    x = model.ord_enc.inverse_transform(x)
-    interpolations.append(x)

### 2021-06-01 notebooks/working_with_midi.py

@@ -5,13 +5,14 @@
 # %%
 import pathlib
 import sys
-import numpy as np
-import note_seq
-import music21
-import pretty_midi
-import pandas as pd
-import pypianoroll
+
 import matplotlib.pyplot as plt
+import music21
+import note_seq
+import numpy as np
+import pandas as pd
+import pretty_midi
+import pypianoroll
 from miditoolkit import midi, pianoroll
 from tensorflow.python.keras.layers.core import RepeatVector
 from tensorflow.python.keras.layers.wrappers import TimeDistributed
@@ -125,9 +126,7 @@ def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
 
 # %%
 def bars(multitrack, start_index, num_bars, beats_per_bar, resolution):
-    start, end = get_bar_bounds(
-        start_index, num_bars, beats_per_bar, resolution
-    )
+    start, end = get_bar_bounds(start_index, num_bars, beats_per_bar, resolution)
     tracks = []
     for track in multitrack.tracks:
         tracks.append(
@@ -151,6 +150,16 @@ first_bar.plot(axes)
 for ax in axes:
     ax.set_ylim(24, 72)
 plt.savefig("../papers/progress/alex/first_bar.png")
+
+# %%
+resolution = 24
+fig, axes = plt.subplots(4, sharex=True, sharey=True)
+first_bar = bars(multitrack, 0, 2, 4, resolution)
+first_bar.plot(axes)
+for ax in axes:
+    ax.set_ylim(24, 72)
+plt.savefig("first_two_bars.png")
+
 
 # %%
 viola_track = first_bar.tracks[2]

### 2021-06-01 papers/final/jack/paper.tex
separate final paper to individual

--- 2021-05-17 papers/final/paper.tex
@@
 
-\keywords{deep learning, neural networks, music}
+\keywords{deep learning, neural networks, sequential models, music}
 
@@
 
+Modelling raw audio waveforms is challenging since a single second in
+modern recording spans thousands of timesteps. Most commonly music is
+recorded at 44.1 kHz (44,100 samples per second). Since a piece of music
+is at least a few seconds in length, capturing long-range dependencies
+is difficult in time domain. A model named MelNet was introduced to address
+this limitation of time-domain models \cite{vasquez2019melnet}. MelNet 
+leverages a two-dimentional time-frequency representation of audio called 
+spectrogram, which reduces the dimentionality of the audio waveforms. This 
+model was able to generate high-fidelity audio samples using structures at 
+long time scales.
+
 \section{Methods}

### 2021-06-01 musiclearn/models/vae.py
fix interpolation method, interpolate 5 points from beethoven to mozart

@@ -275,7 +275,7 @@ class MultiTrackVAE:
         stop = stop[0:min_len, :, :]
         start_mu, start_sigma, start_z = self.encoder_model.predict(start)
         stop_mu, stop_sigma, stop_z = self.encoder_model.predict(stop)
-        space = np.linspace(start_z, stop_z, 3)
+        space = np.linspace(start_z, stop_z, n)
         results = []
         for x in space:
             x = self.decoder_model.predict(x)

### 2021-06-01 notebooks/midi_quartet_inference.py

@@ -20,8 +20,9 @@ mozart_1788 = midis / "Mozart" / "1788_kv_465_1.mid"
 ticks_per_beat = 4
 beats_per_phrase = 4
 programs = [40, 40, 41, 42]
-bee_score = processing.midi_to_music21(beethoven_2494)
 
+# %%
+bee_score = processing.midi_to_music21(beethoven_2494)
 bee = processing.score_to_array(bee_score, ticks_per_beat)
 moz_score = processing.midi_to_music21(mozart_1788)
 moz = processing.score_to_array(moz_score, ticks_per_beat)
@@ -60,8 +61,14 @@ moz_reconst_score.write("midi", "../outputs/1788_kv_465_1_reconst_0016.mid")
 # latent codes
 
 # %%
-start = bee
-stop = moz
-beemoz = model.interpolate(start, stop, 3, ticks_per_beat, beats_per_phrase)
-beemoz_score = processing.array_to_score(beemoz[1], programs=programs, resolution=ticks_per_beat)
-beemoz_score.write("midi", "../outputs/beemoz_0016.mid")
+# Get first 8 measures of each
+bee_8 = processing.score_to_array(bee_score.measures(0, 8), ticks_per_beat)
+moz_8 = processing.score_to_array(moz_score.measures(0, 8), ticks_per_beat)
+
+# %%
+start = bee_8
+stop = moz_8
+beemoz = model.interpolate(start, stop, 5, ticks_per_beat, beats_per_phrase)
+for i, score in enumerate(beemoz):
+    beemoz_score = processing.array_to_score(score, programs=programs, resolution=ticks_per_beat)
+    beemoz_score.write("midi", f"../outputs/beemoz_8_{i}.mid")

### 2021-06-01 papers/final/alex/paper.tex
separate final paper to individual
alex paper updates

--- 2021-05-17 papers/final/paper.tex
@@
 \pagestyle{plain}
-\title{Musical Deep Learning}
+\title{Modeling the Latent Space of Symbolic String Quartet Music}
 
@@
 %% used to denote shared contribution to the research.
-\author{Nhut Minh Phan}
-\email{phan92@uw.edu}
-\affiliation{%
-  \institution{University of Washington}
-  \city{Bothell}
-  \state{WA}
-  \country{USA}
-}
 
@@
 \begin{abstract}
-  This paper presents a generative model of multi-instrument symbolic
-  music.
+  This paper presents the results of a project to apply a recurrent
+  variational autoencoder to the task of modeling a latent space for
+  generating pieces of multi-instrument symbolic classical music
+  through interpolation.
 \end{abstract}
 
-\keywords{deep learning, neural networks, music}
+\keywords{deep learning, recurrent neural networks, sequential models,
+  music modeling, latent space modeling, generative modeling}
 
@@
 educators. Generative models can create accompaniments for music,
-transfer styles between clips of music, and even generate entirely new
-music. Music is challenging to model because it exhibits a complex
-hierarchy of recurring patterns and long-range temporal dependencies,
-and because both musical scores and performances have multiple
-possible digital representations.
+blend and transfer styles between clips of music, and even generate
+entirely new music. Music is challenging to model because it is
+inherently high-dimensional, exhibiting a complex hierarchy of
+recurring patterns and long-range temporal dependencies, and because
+musical scores have multiple possible digital representations with
+distinct advantages and disadvantages.
 
 Depending on the task, machine learning models of music may be trained
-on the audio signal itself, either in a time domain or a frequency
-domain representation, or they may be trained on a digital symbolic
-representation of music, the most common of which is MIDI (Musical
-Instrument Digital Interface) notation. MIDI is an encoding of music
-as streams of bytes in one or more tracks or channels, each
-representing a sequence of 128 possible pitch values (where 0 is the
-lowest and 127 is the highest), along with timing, pressure and
-instrument identifier values. A music transcription model may
-transcribe an audio signal as a MIDI score, which can easily be
-converted into other symbolic representations such as sheet music for
-human performers to read from, while a synthesizer model can convert
-MIDI representations into audio signals. A generative music model can
-be trained either to generate raw audio, or to produce a symbolic
-score that must be played by a synthesizer or by humans to produce an
-audio music performance. This project focuses on the latter type of
-modeling: generative modeling of symbolic (MIDI) music to compose
-original musical scores.
+on the audio signal of a musical performance, either in a time domain
+or a frequency domain representation, or they may be trained on a
+digital symbolic representation of music, the most common of which is
+MIDI (Musical Instrument Digital Interface) notation. MIDI is an
+encoding of music as streams of bytes in one or more tracks or
+channels, each representing a sequence of 128 possible pitch values
+(where 0 is the lowest pitch and 127 is the highest), along with
+timing, pressure and instrument identifier values. A generative
+symbolic music model can produce a symbolic score in MIDI format,
+which must be played by synthesizer software or by humans to produce
+an audio music performance. This project focuses on generative
+modeling of symbolic (MIDI) music to compose original musical scores
+by blending input scores through continuous latent space interpolation.
 
@@
 
-The state of the art in music generation has a long way to go before
-it can consistently generate music scores or performances that would
-be enjoyable and popular for humans to listen to, but a number of
-recent research projects have shown promising progress in this area.
+The state of the art in music generation still has a long way to go
+before it can consistently generate music scores or performances that
+would be enjoyable and popular for humans to listen to, but for this
+reason it is an area of significant opportunity, where a number of
+recent research projects have shown promising progress.
 
@@
 
-Two primary datasets will be used for this research project: MusicNet,
-which is a collection of 330 freely licensed European classical music
-recordings with aligned MIDI scores \cite{thickstun2017learning}, and
-the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
-Commons licensed MIDI files of popular music songs that have been
-matched and aligned to MP3 recordings, and of which we plan to use a
-subset for model training and evaluation
-\cite{raffel_learning-based_2016}. Including this second dataset may
-help to generalize the modeling approach beyond European classical
-music to include other popular genres and associated instruments.
+The dataset used for this research project is MusicNet, which is a
+collection of 330 freely licensed European classical music recordings
+with aligned MIDI scores \cite{thickstun2017learning}.  The model is
+trained on 36 string quartets (four-part arrangements with two
+violins, one viola and one cello) by composers Bach, Beethoven,
+Dvorak, Haydn, Mozart, and Ravel.
 
@@
 The result of this preprocessing is that each training example will be
-a 3D tensor of shape (tracks x ticks x pitches) and stacking the
-training examples will produce a 4D tensor.
+a 2D matrix of shape (time steps x instrument tracks) and stacking the
+training examples will produce a 3D tensor. Information regarding which
+track corresponds to which musical instrument is lost and therefore
+must be maintained separately to reconstruct the MIDI representation.
 
-While we plan to model music with multiple instrument tracks, we
-anticipate the need to model only a fixed selection of instrument
-parts, similar to how MusicVAE models three-part (drum, bass and
-melody) \cite{roberts_hierarchical_2018} and MuseGAN models five-part
-(drum, bass, guitar, string, piano) arrangements.
+As the model will produce one output per instrument track, it can
+incorporate only a fixed selection of instrument parts, similar to how
+MusicVAE models three-part (drum, bass and melody)
+\cite{roberts_hierarchical_2018} and MuseGAN models five-part (drum,
+bass, guitar, string, piano) arrangements. Therefore this model is
+purpose-built for string quartets, which are the most common type of
+arrangement found in the MusicNet dataset.
 
@@
   \caption{Pianoroll visualization of the first measure of
-    Beethoven's Serioso String Quartet}
+    Beethoven's Serioso String Quartet in F Minor}
   \label{pianoroll}
@@
     parts from a Beethoven piece.}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{first_bar_sheet.png}
+  \caption{Sheet music of the first measure of Beethoven's Serioso
+    String Quartet in F Minor}
+  \label{sheet}
+  \Description{A measure of sheet music cello, viola, and violin parts
+    from Beethoven piece.}
 \end{figure}
@@
 \subsection{Model Design}
+
+The model is a recurrent variational autoencoder; both Long Short-Term
+Memory (LSTM) and Gated Recurrent Unit (GRU) cell types were tested.
 

### 2021-06-02 musiclearn/models/vae.py
add bidirectional hparam to vae

@@ -80,6 +80,7 @@ def build_multi_track_vae(
     n_notes,
     dropout_rate=0.2,
     gru=False,
+    bidirectional=False,
 ):
     """Build the multi-track LSTM-VAE."""
     # define encoder model
@@ -91,11 +92,20 @@ def build_multi_track_vae(
     if embedding_dim > 0:
         encoder = layers.Embedding(n_notes, embedding_dim, input_length=n_timesteps)(inputs)
         encoder = layers.Reshape((n_timesteps, embedding_dim * n_tracks))(encoder)
-        encoder = rnn(lstm_units, return_sequences=True)(encoder)
+        if bidirectional:
+            encoder = layers.Bidirectional(rnn(lstm_units, return_sequences=True)(encoder))
+        else:
+            encoder = rnn(lstm_units, return_sequences=True)(encoder)
     else:
-        encoder = rnn(lstm_units, return_sequences=True)(inputs)
+        if bidirectional:
+            encoder = layers.Bidirectional(rnn(lstm_units, return_sequences=True)(inputs))
+        else:
+            encoder = rnn(lstm_units, return_sequences=True)(inputs)
     encoder = layers.Dropout(dropout_rate)(encoder)
-    encoder = rnn(lstm_units, return_sequences=False)(encoder)
+    if bidirectional:
+        encoder = layers.Bidirectional(rnn(lstm_units, return_sequences=False)(encoder))
+    else:
+        encoder = rnn(lstm_units, return_sequences=False)(encoder)
     mu = layers.Dense(latent_dim, name="mu")(encoder)
     sigma = layers.Dense(latent_dim, name="sigma")(encoder)
     # Latent space sampling
@@ -143,8 +153,10 @@ class MultiTrackVAE:
         learning_rate,
         dropout_rate,
         gru=False,
+        bidirectional=False,
     ):
         self.lstm_units = lstm_units
+        self.bidirectional = bidirectional
         self.n_timesteps, self.n_tracks, self.n_notes = (None, None, None)
         self.optimizer = optimizers.Adam(learning_rate)
         self.embedding_dim = embedding_dim
@@ -170,6 +182,7 @@ class MultiTrackVAE:
             learning_rate=self.learning_rate,
             dropout_rate=self.dropout_rate,
             gru=self.gru,
+            bidirectional=self.bidirectional,
         )
         if self.trained_epochs > 0:
             joblib.dump(self.ord_enc, directory / "ordinal_encoder.joblib")

### 2021-06-05 musiclearn/training.py
remove models folder, move vae.py out, consolidate CLIs

@@ -9,8 +9,7 @@ from pathlib import Path
 
 from tensorflow.keras import callbacks
 
-from musiclearn import processing
-from musiclearn.models import vae
+from musiclearn import processing, vae_models
 
 LOG = logging.getLogger(__name__)
 LOG.setLevel(logging.DEBUG)
@@ -67,7 +66,7 @@ def train_mtvae(
     """"""
     # String quartet MIDI programs
     x = processing.get_string_quartets(ticks_per_beat)
-    mtvae = vae.MultiTrackVAE(
+    mtvae = vae_models.MultiTrackVAE(
         lstm_units,
         embedding_dim,
         latent_dim,

### 2021-06-05 musiclearn/vae_models.py

@@ -1,5 +1,6 @@
-"""models/vae.py
-Music Variational Autoencoders
+"""vae_models.py
+Music LSTM Variational Autoencoder models
+Author: Alex Kyllo
 """
 import itertools
 import os
@@ -8,14 +9,16 @@ from pathlib import Path
 import joblib
 import numpy as np
 import tensorflow as tf
-from musiclearn import processing
 from sklearn.preprocessing import OrdinalEncoder
 from tensorflow import keras
 from tensorflow.keras import backend as K
 from tensorflow.keras import layers, models, optimizers
 
+from musiclearn import processing
+
 
 def sample_normal(inputs):
+    """Draw a sample from a multidimensional Gaussian distribution."""
     mu, sigma = inputs
     batch = K.shape(mu)[0]
     dim = K.int_shape(mu)[1]
@@ -32,7 +35,7 @@ def build_one_track_vae(
     n_notes,
     dropout_rate=0.2,
 ):
-    """Build the one-track LSTM-VAE"""
+    """Build a one-track LSTM-VAE Keras model for autoencoding monophonic music."""
     # define encoder model
     inputs = layers.Input(shape=(n_timesteps, 1))
     encoder = layers.Embedding(n_notes, embedding_dim, input_length=n_timesteps)(inputs)
@@ -82,7 +85,7 @@ def build_multi_track_vae(
     gru=False,
     bidirectional=False,
 ):
-    """Build the multi-track LSTM-VAE."""
+    """Build a multi-track LSTM-VAE Keras model for autoencoding polyphonic music."""
     # define encoder model
     inputs = layers.Input(shape=(n_timesteps, n_tracks))
     if gru:
@@ -262,6 +265,7 @@ class MultiTrackVAE:
         return self
 
     def reconstruct(self, x, ticks_per_beat, beats_per_phrase):
+        """Reconstruct an input phrase through the autoencoder."""
         x = self.ord_enc.transform(x).astype(int)
         x = processing.split_array(
             x, beats_per_phrase=beats_per_phrase, resolution=ticks_per_beat, fill=self.rest_code
@@ -274,7 +278,7 @@ class MultiTrackVAE:
         return x
 
     def interpolate(self, start, stop, n, ticks_per_beat, beats_per_phrase):
-        """Interpolate n samples from the latent space between two inputs."""
+        """Interpolate n samples from the latent space between two input phrases."""
         start = self.ord_enc.transform(start).astype(int)
[…]

### 2021-06-05 musiclearn_cli.py

@@ -1,4 +1,7 @@
-"""Command line interface for the musiclearn project."""
+"""musiclearn_cli.py
+Command line interface for the musiclearn project.
+Authors: Alex Kyllo and Jack Phan
+"""
 import logging
 import random
 
@@ -6,7 +9,7 @@ import click
 import numpy as np
 from tensorflow.random import set_seed
 
-from musiclearn import training
+from musiclearn import sequential_models, single_note_processing, training
 
 LOG = logging.getLogger("musiclearn")
 LOG.setLevel(logging.DEBUG)
@@ -61,7 +64,8 @@ def fit_mtvae(
     bidirectional,
     patience,
 ):
-    """Run MultiTrackVAE experiment named EXP_NAME with hyperparameter options."""
+    """Run MultiTrackVAE experiment named EXP_NAME with hyperparameter options.
+    Author: Alex Kyllo"""
     training.train_mtvae(
         exp_name,
         ticks_per_beat,
@@ -79,6 +83,76 @@ def fit_mtvae(
     )
 
 
+@click.command()
+@click.option(
+    "--model-type",
+    type=click.Choice(["lstm", "bidirect", "att", "wavenet"], case_sensitive=False),
+    help="Type of model",
+)
+@click.option("--data-path", type=click.STRING, help="Path to folder stores dataset")
+@click.option("--logs-dir", type=click.STRING, help="Folder that stores training logs")
+@click.option(
+    "--sequence_length",
+    type=click.INT,
+    default=100,
+    help="Length of the sequences of notes used in training",
+)
+def fit_sequential(model_type, data_path, logs_dir, sequence_length):
+    """Fit a sequential model of choice on the specified dataset.
+    Author: Jack Phan"""
+    notes = single_note_processing.read_midi(data_path)
+
+    n_vocab = single_note_processing.get_num_unique_notes(notes)
+
+    network_input, network_output = single_note_processing.prepare_sequences(sequence_length, notes)
+
+    if model_type == "lstm":
+        model = sequential_models.lstm_model(network_input.shape, n_vocab)
+    elif model_type == "bidirect":
+        model = sequential_models.bidirectional_lstm_model(network_input.shape, n_vocab)
+    elif model_type == "att":
+        model = sequential_models.attention_lstm_model(network_input.shape, n_vocab)
+    elif model_type == "wavenet":
+        model = sequential_models.simplified_wavenet(network_input.shape, n_vocab)
+    else:
+        raise ValueError("Invalid model type")
+
[…]

### 2021-06-05 notebooks/midi_quartet_inference.py

@@ -8,11 +8,7 @@ from pathlib import Path
 
 sys.path.append("..")
 import numpy as np
-from musiclearn import config, processing
-from musiclearn.models import vae
-
-# %%
-exp_name = "mtvae_0017"
+from musiclearn import config, processing, vae_models
 
 # %%
 midis = Path(config.MUSICNET_MIDI_DIR)
@@ -31,12 +27,10 @@ moz_score = processing.midi_to_music21(mozart_1788)
 moz = processing.score_to_array(moz_score, ticks_per_beat)
 
 # %%
-# saved_model_path = "../experiments/mtvae_0009/2021-05-31T20:24:32/saved_model"
-# saved_model_path = "../experiments/mtvae_0013/2021-05-31T22:15:37/saved_model"
-# saved_model_path = "../experiments/mtvae_0015/2021-05-31T22:38:59/saved_model"
-# saved_model_path = "../experiments/mtvae_0016/2021-05-31T23:01:43/saved_model"
-saved_model_path = f"../experiments/{exp_name}/2021-06-02T22:55:43/saved_model"
-model = vae.MultiTrackVAE.from_saved(saved_model_path)
+exp_name = "mtvae_0016"
+exp_time = "2021-05-31T23:01:43"
+saved_model_path = f"../experiments/{exp_name}/{exp_time}/saved_model"
+model = vae_models.MultiTrackVAE.from_saved(saved_model_path)
 
 # %% [markdown]
 # ## Reconstruction
@@ -74,9 +68,10 @@ beemoz = model.interpolate(bee_8, moz_8, 5, ticks_per_beat, beats_per_phrase)
 for i, score in enumerate(beemoz):
     beemoz_score = processing.array_to_score(score, programs=programs, resolution=ticks_per_beat)
     beemoz_score.write("midi", f"../outputs/{exp_name}/beemoz_8_{i}.mid")
+    beemoz_score.write("musicxml", f"../outputs/{exp_name}/beemoz_8_{i}.xml")
 
 # %% [markdown]
-# another example
+# another example: Dvorak and Bach
 
 # %%
 dvorak_1916 = midis / "Dvorak" / "1916_dvq10m1.mid"
@@ -93,6 +88,25 @@ for i, score in enumerate(dvorbach):
     dvorbach_score.write("midi", f"../outputs/{exp_name}/dvorbach_8_{i}.mid")
 
 # %% [markdown]
+# another example: Beethoven and Bach
+
+# %%
[…]

### 2021-06-05 notebooks/midi_quartet_model_no_chords.py

@@ -24,8 +24,7 @@ from tensorflow.keras import backend as K
 from tensorflow.keras import layers, losses, metrics, optimizers, utils
 
 sys.path.append("..")
-from musiclearn import processing
-from musiclearn.models import vae
+from musiclearn import processing, vae_models
 
 # %%
 physical_devices = tf.config.list_physical_devices("GPU")
@@ -92,7 +91,7 @@ embedding_dim = 8
 opt = optimizers.Adam(learning_rate=0.0002)
 
 # %%
-lstm_vae, encoder, decoder = vae.build_one_track_vae(
+lstm_vae, encoder, decoder = vae_models.build_one_track_vae(
     opt, latent_dim, embedding_dim, n_timesteps, n_notes, dropout_rate=0.2
 )
 history = lstm_vae.fit(x_0, x_0, batch_size=32, epochs=500, validation_split=0.1)
@@ -105,14 +104,8 @@ pred_0
 
 # %% [markdown]
 # Variational LSTM-AE (four track)
-mtvae, mencoder, mdecoder = vae.build_multi_track_vae(
+mtvae, mencoder, mdecoder = vae_models.build_multi_track_vae(
     opt, lstm_units, latent_dim, embedding_dim, n_timesteps, n_features, n_notes, dropout_rate=0.2
 )
 mhistory = mtvae.fit(x, tf.unstack(x, axis=2), batch_size=32, epochs=100, validation_split=0.1)
 mtvae.save_weights("mtvae.hdf5")
-# %% [markdown]
-#
-# - [] TODO: try one-hot encoding instead?
-# - [] TODO: Try higher dimension embedding?
-# - [X] TODO: Try Higher latent dimension
-# - [X] TODO: Implement VAE for multi-track

### 2021-06-05 generate_music_cli.py
remove separate CLI files

@@ -1,37 +1,0 @@
-import click
-from musiclearn.sequential_models import *
-from musiclearn.single_note_processing import *
-
-
-@click.command()
-@click.argument("output-name", type=click.STRING)
-@click.option('--data-path', type=click.STRING, help='Path to folder stores dataset')
-@click.option('--model-type',
-              type=click.Choice(['lstm', 'bidirect', 'att', 'wavenet'], case_sensitive=False),
-              help='Type of model')
-@click.option('--weights-path', type=click.STRING, help='Path to saved weights of the model')
-@click.option('--num-notes', type=click.INT, help='Number of notes to generate')
-def generate_music(output_name,
-                   data_path,
-                   model_type,
-                   weights_path,
-                   num_notes):
-    """ Generate a short piece of music with a fixed number of notes """
-    # load the model
-    notes = read_midi(data_path)
-    n_vocab = get_num_unique_notes(notes)
-    network_input, _ = prepare_sequences(SEQUENCE_LENGTH, notes)
-    if model_type == 'lstm' or model_type == 'bidirect':
-        model = load_lstm_model(weights_path)
-    elif model_type == 'att':
-        model = load_attention_lstm_model(weights_path, network_input.shape, n_vocab)
-    elif model_type == 'wavenet':
-        model = load_wavenet_model(weights_path)
-    else:
-        raise ValueError("Invalid model type")
-    
-    generate_midi_sample(model, data_path, output_name, num_notes)
-
-    
-if __name__ == "__main__":
-    generate_music()

### 2021-06-05 sequential_models_cli.py

@@ -1,40 +1,0 @@
-import click
-from musiclearn.sequential_models import *
-from musiclearn.single_note_processing import *
-
-
-@click.command()
-@click.option('--model-type',
-              type=click.Choice(['lstm', 'bidirect', 'att', 'wavenet'], case_sensitive=False),
-              help='Type of model')
-@click.option('--data-path', type=click.STRING, help='Path to folder stores dataset')
-@click.option('--logs-dir', type=click.STRING, help='Folder that stores training logs')
-@click.option('--sequence_length', type=click.INT, default=100,
-              help='Length of the sequences of notes used in training')
-def fit_sequential(model_type,
-                   data_path,
-                   logs_dir,
-                   sequence_length):
-    """ Fit a sequential model of choice on the specified dataset """
-    notes = read_midi(data_path)
-    
-    n_vocab = get_num_unique_notes(notes)
-    
-    network_input, network_output = prepare_sequences(sequence_length, notes)
-    
-    if model_type == 'lstm':
-        model = lstm_model(network_input.shape, n_vocab)
-    elif model_type == 'bidirect':
-        model = bidirectional_lstm_model(network_input.shape, n_vocab)
-    elif model_type == 'att':
-        model = attention_lstm_model(network_input.shape, n_vocab)
-    elif model_type == 'wavenet':
-        model = simplified_wavenet(network_input.shape, n_vocab)
-    else:
-        raise ValueError("Invalid model type")
-    
-    train_model(model, sequence_length, model_type, 10, 64)           
-
-
-if __name__ == "__main__":
-    fit_sequential()

### 2021-06-05 musiclearn/config.py
code cleanup

@@ -1,10 +1,9 @@
 import os
+
 import dotenv
 
 dotenv.load_dotenv()
 
-MUSICNET_DIR = str(os.getenv("MUSICNET_DIR"))
 MUSICNET_MIDI_DIR = str(os.getenv("MUSICNET_MIDI_DIR"))
-MUSICNET_SAMPLE_RATE = str(os.getenv("MUSICNET_SAMPLE_RATE"))
 FF_MIDI_DIR = str(os.getenv("FF_MIDI_DIR"))
-LOGS_DIR=str(os.getenv("LOGS_DIR"))
+LOGS_DIR = str(os.getenv("LOGS_DIR"))

### 2021-06-05 musiclearn/processing.py

@@ -1,5 +1,6 @@
 """processing.py
 Preprocessing audio data (.wav files) for analysis.
+Author: Alex Kyllo
 """
 import logging
 import os

### 2021-06-05 musiclearn/sequential_models.py

@@ -1,22 +1,26 @@
-import sys
+"""sequential_models.py
+Author: Jack Phan
+"""
 import os
 import pathlib
+import sys
+
 import tensorflow as tf
-from tensorflow.keras.models import Sequential
-from tensorflow.keras.layers import *
-from tensorflow.keras.callbacks import ModelCheckpoint, TensorBoard
 from tensorflow.keras import backend as K
+from tensorflow.keras.callbacks import ModelCheckpoint, TensorBoard
+from tensorflow.keras.layers import *
+from tensorflow.keras.models import Sequential
 from tensorflow.keras.utils import plot_model
 
 sys.path.append("..")
-from musiclearn.single_note_processing import *
 from musiclearn import config
+from musiclearn.single_note_processing import *
 
 # path to Schubert collection
 schubert_dir = pathlib.Path(config.MUSICNET_MIDI_DIR) / "Schubert"
 
 # path to location of tensorboard logs
-#logs_dir = pathlib.Path(config.LOGS_DIR)
+# logs_dir = pathlib.Path(config.LOGS_DIR)
 logs_dir = pathlib.Path("/csslab-localdata/csslab-si/jack_working_dir/test_logs")
 
 # length of an input sequence
@@ -26,11 +30,7 @@ SEQUENCE_LENGTH = 100
 def lstm_model(input_shape, n_vocab):
     """ Build a simple LSTM model """
     model = Sequential()
-    model.add(LSTM(
-        512,
-        input_shape=(input_shape[1], input_shape[2]),
-        return_sequences=True
-    ))
+    model.add(LSTM(512, input_shape=(input_shape[1], input_shape[2]), return_sequences=True))
     model.add(Dropout(0.3))
     model.add((LSTM(512, return_sequences=True)))
     model.add(Dropout(0.3))
@@ -38,22 +38,26 @@ def lstm_model(input_shape, n_vocab):
     model.add(BatchNormalization())
     model.add(Dropout(0.3))
     model.add(Dense(256))
-    model.add(Activation('relu'))
+    model.add(Activation("relu"))
     model.add(BatchNormalization())
     model.add(Dropout(0.3))
     model.add(Dense(n_vocab))
-    model.add(Activation('softmax'))
-    
-    model.compile(loss='sparse_categorical_crossentropy', optimizer='rmsprop')
-    
+    model.add(Activation("softmax"))
+
+    model.compile(loss="sparse_categorical_crossentropy", optimizer="rmsprop")
[…]

### 2021-06-05 musiclearn/single_note_processing.py

@@ -1,18 +1,20 @@
+"""single_note_processing.py
+Author: Jack Phan
+"""
 import glob
-import pathlib
-import pickle
-import numpy as np
-import sys
 import os
-from music21 import *
+import pickle
+
+import numpy as np
+from music21 import chord, converter, instrument, note, stream
 
 
 def read_midi(midi_dir):
     """ Read all midi file in a collection and return the notes """
     notes = []
-    
+
     noteFile_dir = str(midi_dir) + "/notes"
-    
+
     if not os.path.exists(noteFile_dir):
         files_dir = str(midi_dir) + "/*.mid"
         for file in glob.glob(files_dir):
@@ -22,30 +24,30 @@ def read_midi(midi_dir):
 
             notes_to_parse = None
 
-            try: # file has instrument parts
+            try:  # file has instrument parts
                 s2 = instrument.partitionByInstrument(midi)
-                
+
                 # looping over all the instrument parts
                 for part in s2.parts:
-                    #select elements of piano
+                    # select elements of piano
                     if "Piano" in str(part):
                         notes_to_parse = part.recurse()
-            except: # file has notes in a flat structure
+            except:  # file has notes in a flat structure
                 notes_to_parse = midi.flat.notes
 
             # if there are notes to parse
             if notes_to_parse is not None:
                 for element in notes_to_parse:
                     if isinstance(element, note.Note):
-                         notes.append(str(element.pitch))
+                        notes.append(str(element.pitch))
                     elif isinstance(element, chord.Chord):
-                        notes.append('.'.join(str(n) for n in element.normalOrder))
+                        notes.append(".".join(str(n) for n in element.normalOrder))
 
         # save the note to file
-        with open(noteFile_dir, 'wb') as fp:
+        with open(noteFile_dir, "wb") as fp:
             pickle.dump(notes, fp)
     else:
-        with open(noteFile_dir, 'rb') as fp:
+        with open(noteFile_dir, "rb") as fp:
             notes = pickle.load(fp)
     return notes
 
@@ -54,7 +56,7 @@ def get_unique_notes(notes):
     """ Return a sorted set of unique notes """
     pitchNames = sorted(set(item for item in notes))
     return pitchNames
-    
+
 
 def get_num_unique_notes(notes):
     """ Return the number of unique notes"""
@@ -82,32 +84,32 @@ def map_int_to_notes(notes):
 
 
 def prepare_sequences(sequence_length, notes):
[…]

### 2021-06-05 musiclearn/training.py

@@ -63,7 +63,7 @@ def train_mtvae(
     bidirectional,
     patience=10,
 ):
-    """"""
+    """Train the MultiTrack VAE once with a given set of hyperparameters."""
     # String quartet MIDI programs
     x = processing.get_string_quartets(ticks_per_beat)
     mtvae = vae_models.MultiTrackVAE(

### 2021-06-05 notebooks/midi_model_playground.py

@@ -1,165 +1,0 @@
-# %% [markdown]
-# # MIDI Model Playground
-#
-# An exploratory notebook for playing with model architecture ideas.
-# %%
-%load_ext autoreload
-%autoreload 2
-
-# %%
-import pathlib
-import sys
-
-import matplotlib.pyplot as plt
-import numpy as np
-import pypianoroll
-from tensorflow.keras import Model, Sequential, layers, optimizers, utils
-
-sys.path.append("..")
-from musiclearn import config
-
-# %%
-# Picking one song from MusicNet dataset
-midi_dir = pathlib.Path(config.MUSICNET_MIDI_DIR)
-mid_2494 = midi_dir / "Beethoven" / "2494_qt11_1.mid"
-multitrack = pypianoroll.read(mid_2494, resolution=24)
-
-# %% [markdown]
-#
-# ## Training data representation
-#
-# Because an entire song is too long music is generally composed in bars or
-# multi-bar phrases, we plan to read MIDI files and segment them into phrases,
-# one training example per phrase.
-
-# %%
-def get_num_beats(multitrack, resolution):
-    return len(multitrack.downbeat) // resolution
-
-
-def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
-    start = bar_index * resolution
-    end = start + (num_bars * beats_per_bar) * resolution
-    return (start, end)
-
-
-# %%
-def bars(multitrack, start_index, num_bars, beats_per_bar, resolution):
-    start, end = get_bar_bounds(start_index, num_bars, beats_per_bar, resolution)
-    tracks = []
-    for track in multitrack.tracks:
-        tracks.append(
-            pypianoroll.Track(
-                name=track.name,
-                program=track.program,
-                pianoroll=track[start:end],
-            )
-        )
-    return pypianoroll.Multitrack(tracks=tracks, resolution=resolution)
-
-
-# %% [markdown]
-# Plot a bar of Beethoven:
-
-# %%
-resolution = 24
-first_bar = bars(multitrack, 0, 1, 4, resolution)
-# %%
-viola_track = first_bar.tracks[2]
-
-# %% [markdown]
-#
-# ## Pianoroll sequence autoencoder model
-#
-# Testing out training a simple autoencoder model on a single track.
-
-# %%
[…]

### 2021-06-05 notebooks/midi_quartet_model.py

@@ -1,103 +1,0 @@
-# %% [markdown]
-# # MIDI MusicNet string quartet model exploration with pypianoroll representation
-#
-# Alex Kyllo
-# May 9, 2021
-#
-# An exploratory notebook for playing with model architecture ideas.
-
-# %%
-#%load_ext autoreload
-#%autoreload 2
-
-import itertools
-# %%
-import os
-import pathlib
-import sys
-
-import joblib
-import numpy as np
-from sklearn.preprocessing import OrdinalEncoder
-from tensorflow.keras import (Model, Sequential, layers, metrics, optimizers,
-                              utils)
-
-sys.path.append("..")
-from musiclearn import processing
-from musiclearn.models import vae
-
-# %% [markdown]
-# Get all string quartets from MusicNet dataset
-
-# %%
-f = "quartets_music21.npy"
-program_ids = [40, 40, 41, 42]
-ticks_per_beat = 4
-if os.path.isfile(f):
-    x = np.load(f, allow_pickle=True)
-else:
-    scores_file = "scores.joblib"
-    if os.path.isfile(scores_file):
-        scores, fnames = joblib.load(scores_file)
-    else:
-        # Read all MusicNet string quartets
-        scores, fnames = processing.musicnet_quartets_to_music21(program_ids=program_ids)
-        joblib.dump((scores, fnames), scores_file)
-    # Stack them into one big 2D array
-    x = np.vstack(
-        [processing.score_to_str_array(score, resolution=ticks_per_beat) for score in scores]
-    )
-    np.save(f, x)
-
-# %%
-notes = np.unique(x)
-n_notes = notes.shape[0]
-enc = OrdinalEncoder(categories=list(itertools.repeat(notes, 4)))
-x = enc.fit_transform(x).astype(int)
-REST_CODE = np.argwhere(enc.categories_[0] == "REST")[0][0]
-
-# %%
-x = processing.split_array(x, beats_per_phrase=8, resolution=ticks_per_beat, fill=REST_CODE)
-n_timesteps = x.shape[1]
-n_features = x.shape[2]
-shape = (n_timesteps, n_features)
-print(shape)
-
-# %% [markdown]
-# First, one track autoencoder
-
-# %%
-model = vae.OneTrackAE(
-    optimizer=optimizers.Adam(learning_rate=0.0001),
-    n_timesteps=n_timesteps,
-    n_notes=n_notes,
-)
[…]

### 2021-06-05 notebooks/midi_quartet_model_pianoroll.py

@@ -1,79 +1,0 @@
-# %% [markdown]
-# # MIDI MusicNet string quartet model exploration with pypianoroll representation
-#
-# Alex Kyllo
-# May 9, 2021
-#
-# An exploratory notebook for playing with model architecture ideas.
-
-# %%
-%load_ext autoreload
-%autoreload 2
-
-import os
-# %%
-import pathlib
-import sys
-
-import matplotlib.pyplot as plt
-import numpy as np
-import pypianoroll
-from tensorflow.keras import (Model, Sequential, layers, metrics, optimizers,
-                              utils)
-
-sys.path.append("..")
-from musiclearn import config, processing
-
-# %% [markdown]
-# Get all string quartets from MusicNet dataset in shape
-# (
-
-# %%
-f = "quartets.npy"
-if os.path.isfile(f):
-    x = np.load(f)
-else:
-    x = processing.musicnet_quartets_to_numpy(bars_per_phrase=4)
-    np.save(f, x)
-
-n_timesteps = x.shape[1]
-n_features = x.shape[2]
-shape = (n_timesteps, n_features)
-print(shape)
-
-# %% [markdown]
-#
-# We are predicting what notes each instrument is playing, so this is
-# either a multiclass classification problem, or perhaps we need an
-# architecture with multiple outputs.
-#
-# This model fails to learn the problem, probably due to sparsity of
-# the pianoroll data.
-
-# %%
-opt = optimizers.Adam(learning_rate=0.0001)
-act = "tanh"
-model = Sequential()
-model.add(
-    layers.LSTM(
-        512,
-        activation="tanh",
-        input_shape=(n_timesteps, n_features),
-        return_sequences=True,
-    )
-)
-model.add(layers.LSTM(256, activation=act, return_sequences=True))
-model.add(layers.LSTM(256, activation=act, return_sequences=True))
-model.add(layers.LSTM(512, activation=act, return_sequences=True))
-model.add(layers.TimeDistributed(layers.Dense(n_features, activation="sigmoid")))
-model.compile(
-    optimizer=opt,
-    loss="binary_crossentropy",
-    metrics=[metrics.Precision(), metrics.Recall()],
-)
-utils.plot_model(model, show_shapes=True)
-
-# %%
-history = model.fit(x, x, epochs=200, validation_split=0.2)
-history.model.save("quartet_lstm_model")
-np.savez(history.history, "quartet_lstm_history.npz")

### 2021-06-05 notebooks/midi_quartet_training.py

@@ -7,21 +7,15 @@
 # An exploratory notebook for playing with model architecture ideas.
 
 # %%
-#%load_ext autoreload
-#%autoreload 2
-# %%
 import itertools
 import os
-import pathlib
 import sys
 
 import joblib
 import numpy as np
 import tensorflow as tf
 from sklearn.preprocessing import OrdinalEncoder
-from tensorflow.keras import Model, Sequential
-from tensorflow.keras import backend as K
-from tensorflow.keras import layers, losses, metrics, optimizers, utils
+from tensorflow.keras import optimizers
 
 sys.path.append("..")
 from musiclearn import processing, vae_models

### 2021-06-05 notebooks/tempCodeRunnerFile.py

@@ -1,10 +1,0 @@
-# %%
-import pathlib
-import sys
-import numpy as np
-import pypianoroll
-import matplotlib.pyplot as plt
-from tensorflow.keras import utils, layers, Model, optimizers, Sequential
-
-sys.path.append("..")
-from musiclearn import config

### 2021-06-05 musiclearn/plotting.py
add plot_losses command to CLI

@@ -1,0 +1,31 @@
+"""plotting.py
+Functions for plotting results.
+Author: Alex Kyllo
+"""
+import os
+
+import matplotlib
+import matplotlib.pyplot as plt
+import pandas as pd
+
+
+def plot_learning_curves(history_file: os.PathLike, dest_path: os.PathLike):
+    """Plot the train vs. validation learning curves from a history file."""
+    if dest_path.endswith(".pgf"):
+        matplotlib.use("pgf")
+    df = pd.read_csv(history_file)
+    fig, ax = plt.subplots()
+    y = df.loss
+    y_val = df.val_loss
+    label = "Loss"
+    ax.plot(df.epoch, y, label=f"Training {label}")
+    ax.plot(df.epoch, y_val, label=f"Validation {label}")
+    ax.set_xlabel("Training Epochs")
+    ax.set_ylabel(f"{label} Score")
+    ax.set_ylim(0.0)
+    ax.legend(loc="best")
+    fig.tight_layout()
+    plt.savefig(
+        dest_path,
+    )
+    return (fig, ax)

### 2021-06-05 musiclearn_cli.py

@@ -9,7 +9,8 @@ import click
 import numpy as np
 from tensorflow.random import set_seed
 
-from musiclearn import sequential_models, single_note_processing, training
+from musiclearn import (plotting, sequential_models, single_note_processing,
+                        training)
 
 LOG = logging.getLogger("musiclearn")
 LOG.setLevel(logging.DEBUG)
@@ -153,6 +154,14 @@ def generate_music(output_name, data_path, model_type, weights_path, num_notes):
     sequential_models.generate_midi_sample(model, data_path, output_name, num_notes)
 
 
+@click.command()
+@click.argument("historyfile", type=click.Path(exists=True))
+@click.argument("outputfile", type=click.Path())
+def plot_losses(historyfile, outputfile):
+    """Plot model training & validation loss curves from HISTORYFILE and save to OUTPUTFILE."""
+    plotting.plot_learning_curves(historyfile, outputfile)
+
+
 @click.group()
 def cli():
     """Command line interface for the musiclearn project"""
@@ -163,6 +172,7 @@ def main():
     cli.add_command(fit_mtvae)
     cli.add_command(fit_sequential)
     cli.add_command(generate_music)
+    cli.add_command(plot_losses)
     cli()
 
 

### 2021-06-05 notebooks/working_with_midi.py

@@ -1,169 +1,0 @@
-# %% [markdown]
-# # Exploring MIDI Music Data
-#
-# An exploratory notebook for reading and manipulating MIDI music data.
-# %%
-import pathlib
-import sys
-
-import matplotlib.pyplot as plt
-import music21
-import note_seq
-import numpy as np
-import pandas as pd
-import pretty_midi
-import pypianoroll
-from miditoolkit import midi, pianoroll
-from tensorflow.python.keras.layers.core import RepeatVector
-from tensorflow.python.keras.layers.wrappers import TimeDistributed
-
-sys.path.append("..")
-from musiclearn import config
-
-# %%
-# Picking one song from MusicNet dataset
-midi_dir = pathlib.Path(config.MUSICNET_MIDI_DIR)
-mid_2494 = midi_dir / "Beethoven" / "2494_qt11_1.mid"
-
-# %% [markdown]
-# ## Using the pretty_midi package
-#
-# `pretty_midi` encodes the note pitch values, velocities and start and end
-# times of each instrument track. Gaps in timing between notes are rests. It
-# includes methods for estimating the tempo and time signature, and converting
-# between representations.
-# %%
-pretty_2494 = pretty_midi.PrettyMIDI(str(mid_2494))
-
-# %%
-print(pretty_2494.instruments)
-print(pretty_2494.instruments[0].notes[0:50])
-
-# %% [markdown]
-#
-# One representation is a piano roll, which is a numpy 2D array of notes over
-# time steps at a given frames per second rate (symbolic timing). There are 128
-# notes (0-127) and the values are the note velocities. Or, if binarized, the
-# values are 1 for playing, else 0 for resting. This format allows chords but
-# assumes a single instrument. So it loses instrument information unless you
-# produce separate per-instrument piano rolls. It produces a sparse matrix since
-# most notes aren't being played at most time steps.
-
-# %%
-# returns a (notes, time) dimension numpy array
-piano_2494 = pretty_2494.get_piano_roll(fs=10)
-# Show as pandas DataFrame where rows are time steps and columns are note values
-pd.DataFrame(piano_2494.T)
-
-# %% [markdown]
-# ## Using miditoolkit
-#
-# miditoolkit is similar to pretty_midi but uses symbolic (tick) timing instead
[…]

### 2021-06-05 scripts/quartet_hierarchical.py

@@ -1,0 +1,25 @@
+import numpy as np
+from tensorflow.keras import layers, models
+
+latent_dim = 128
+state_size = 128
+z = np.zeros((1, latent_dim))
+
+# The latent vector z is passed through a fully connected layer with tanh activation
+# to set the initial state of a "conductor" RNN which produces one embedding vector
+# per subsequence (measure). First conductor is given zero inputs and its h, c states
+# are passed to subsequent conductors.
+# Each embedding vector is individually pass through a shared dense layer with tanh activation
+# to produce initial states for a final bottom-layer decoder rnn with softmax activation.
+# At each step of the bottom level decoder, the current embedding is concatenated with the
+# previous output to be used as the input (note: conditioning)
+# paraphrased from MusicVAE paper: https://arxiv.org/pdf/1803.05428.pdf
+# Explanation: https://medium.com/@musicvaeubcse/musicvae-understanding-of-the-googles-work-for-interpolating-two-music-sequences-621dcbfa307c
+
+
+decoder_input = layers.Input(shape=(latent_dim,))
+decoder = layers.Dense(latent_dim * state_size * 4, activation="tanh")(decoder_input)
+decoder, state_h, state_c = layers.LSTM(
+    latent_dim, initial_state=[], return_sequences=True, return_state=True
+)(decoder)
+# TODO

### 2021-06-05 musiclearn/processing.py
clean up unused code in processing.py

@@ -4,294 +4,62 @@ Author: Alex Kyllo
 """
 import logging
 import os
-import sys
-from functools import reduce
 from pathlib import Path
-from typing import Any, List, Tuple
+from typing import Any, List
 
 import joblib
 import numpy as np
 import pandas as pd
-import pypianoroll
 from music21 import chord, converter, instrument, note, stream
-from tensorflow import keras
 
 from musiclearn import config
 
 LOG = logging.getLogger(__name__)
 LOG.setLevel(logging.DEBUG)
 
+# Special note pitch values for rest and sustain
+REST = 128
+SUSTAIN = 129
 
-class MIDIDataGenerator(keras.utils.Sequence):
-    """Generates multitrack pianoroll tensors from MIDI files."""
-
-    def __init__(self, directory: os.PathLike):
-        """constructor"""
-        self.directory = directory
-
-    def __getitem__(self, index):
-        """[] accessor. Get one batch as a numpy array."""
-
-
-def midi_to_multitrack(path: os.PathLike, resolution: int) -> pypianoroll.Multitrack:
-    """"""
-    if os.path.isdir(path):
-        raise IsADirectoryError(f"{path} is a directory, expected a file.")
-    if not os.path.isfile(path):
-        raise FileNotFoundError(f"{path} is not a file.")
-    try:
-        mt = pypianoroll.read(str(path))
-    except:
-        return None
-    return mt
-
-
-def num_beats(multitrack: pypianoroll.Multitrack) -> int:
-    """Get the integer number of beats in a Multitrack piano roll"""
-    assert multitrack.downbeat is not None
-    assert multitrack.resolution is not None
-    return len(multitrack.downbeat) // multitrack.resolution
-
-
-def test_num_beats():
-    path = Path(config.MUSICNET_MIDI_DIR)
-    mid_file = list(path.glob("Beethoven/2494*.mid"))[0]
-    mt = midi_to_multitrack(mid_file, 24)
-    assert num_beats(mt) == 609
-
-
-def get_bar_bounds(bar_index, num_bars, beats_per_bar, resolution):
-    start = bar_index * resolution
-    end = start + (num_bars * beats_per_bar) * resolution
-    return (start, end)
-
-
-def get_phrase(
-    multitrack: pypianoroll.Multitrack,
-    start_index: int,
-    num_bars: int,
[…]

### 2021-06-05 musiclearn/vae_models.py
add interpolation command to CLI

@@ -161,6 +161,7 @@ class MultiTrackVAE:
         self.lstm_units = lstm_units
         self.bidirectional = bidirectional
         self.n_timesteps, self.n_tracks, self.n_notes = (None, None, None)
+        self.ticks_per_beat, self.beats_per_phrase = (None, None)
         self.optimizer = optimizers.Adam(learning_rate)
         self.embedding_dim = embedding_dim
         self.latent_dim = latent_dim
@@ -200,6 +201,8 @@ class MultiTrackVAE:
                 learning_rate=self.learning_rate,
                 trained_epochs=self.trained_epochs,
                 rest_code=self.rest_code,
+                ticks_per_beat=self.ticks_per_beat,
+                beats_per_phrase=self.beats_per_phrase,
             )
             joblib.dump(train_state, directory / "train_state.joblib")
         joblib.dump(hparams, directory / "hparams.joblib")
@@ -223,11 +226,15 @@ class MultiTrackVAE:
         model.batch_size = train_state["batch_size"]
         model.learning_rate = train_state["learning_rate"]
         model.trained_epochs = train_state["trained_epochs"]
+        model.ticks_per_beat = train_state.get("ticks_per_beat", 4)
+        model.beats_per_phrase = train_state.get("beats_per_phrase", 4)
         return model
 
     def train(self, x, ticks_per_beat, beats_per_phrase, epochs, callbacks=None):
         """Train the model on a dataset."""
         # Dataset prep, ordinal encoding
+        self.ticks_per_beat = ticks_per_beat
+        self.beats_per_phrase = beats_per_phrase
         notes = np.unique(x)
         self.n_notes = notes.shape[0]
         self.n_tracks = x.shape[1]
@@ -277,15 +284,21 @@ class MultiTrackVAE:
         x = self.ord_enc.inverse_transform(x)
         return x
 
-    def interpolate(self, start, stop, n, ticks_per_beat, beats_per_phrase):
+    def interpolate(self, start, stop, n):
         """Interpolate n samples from the latent space between two input phrases."""
         start = self.ord_enc.transform(start).astype(int)
         stop = self.ord_enc.transform(stop).astype(int)
         start = processing.split_array(
-            start, beats_per_phrase=beats_per_phrase, resolution=ticks_per_beat, fill=self.rest_code
+            start,
+            beats_per_phrase=self.beats_per_phrase,
+            resolution=self.ticks_per_beat,
+            fill=self.rest_code,
         )
[…]

### 2021-06-05 musiclearn_cli.py

@@ -4,13 +4,14 @@ Authors: Alex Kyllo and Jack Phan
 """
 import logging
 import random
+from pathlib import Path
 
 import click
 import numpy as np
 from tensorflow.random import set_seed
 
-from musiclearn import (plotting, sequential_models, single_note_processing,
-                        training)
+from musiclearn import (plotting, processing, sequential_models,
+                        single_note_processing, training, vae_models)
 
 LOG = logging.getLogger("musiclearn")
 LOG.setLevel(logging.DEBUG)
@@ -85,6 +86,27 @@ def fit_mtvae(
 
 
 @click.command()
+@click.argument("model_path", type=click.Path(exists=True))
+@click.argument("midi_1", type=click.Path(exists=True))
+@click.argument("midi_2", type=click.Path(exists=True))
+@click.argument("output_dir", type=click.Path())
+@click.option("--n", type=click.INT, default=3)
+def interpolate(model_path, midi_1, midi_2, output_dir, n):
+    """Use MODEL_PATH to interpolate n"""
+    programs = [40, 40, 41, 42]  # violin x2, viola, cello
+    model = vae_models.MultiTrackVAE.from_saved(model_path)
+    scores = [processing.midi_to_music21(f) for f in [midi_1, midi_2]]
+    arrays = [processing.score_to_array(sc, model.ticks_per_beat) for sc in scores]
+    output_arrays = model.interpolate(arrays[0], arrays[1], n)
+    output_scores = [
+        processing.array_to_score(arr, programs=programs, resolution=model.ticks_per_beat)
+        for arr in output_arrays
+    ]
+    for i, score in enumerate(output_scores):
+        score.write("midi", Path(output_dir) / f"interpolation_{i}.mid")
+
+
+@click.command()
 @click.option(
     "--model-type",
     type=click.Choice(["lstm", "bidirect", "att", "wavenet"], case_sensitive=False),
@@ -155,11 +177,11 @@ def generate_music(output_name, data_path, model_type, weights_path, num_notes):
 
 
 @click.command()
-@click.argument("historyfile", type=click.Path(exists=True))
-@click.argument("outputfile", type=click.Path())
-def plot_losses(historyfile, outputfile):
-    """Plot model training & validation loss curves from HISTORYFILE and save to OUTPUTFILE."""
-    plotting.plot_learning_curves(historyfile, outputfile)
+@click.argument("history_file", type=click.Path(exists=True))
[…]

### 2021-06-05 run_experiments.sh

@@ -22,4 +22,5 @@
 #python musiclearn_cli.py fit-mtvae mtvae_0014 --beats-per-phrase 8 --lstm-units 256 --latent-dim 512 --dropout-rate 0.5 --patience 20
 #python musiclearn_cli.py fit-mtvae mtvae_0015 --beats-per-phrase 8 --lstm-units 128 --latent-dim 128 --dropout-rate 0.5 --patience 20
 #python musiclearn_cli.py fit-mtvae mtvae_0016 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20
-python musiclearn_cli.py fit-mtvae mtvae_0017 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --bidirectional --patience 20
+#python musiclearn_cli.py fit-mtvae mtvae_0017 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --bidirectional --patience 20
+python musiclearn_cli.py fit-mtvae mtvae_0018 --beats-per-phrase 8 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20

### 2021-06-05 scripts/midi_quartet_inference.py

@@ -1,13 +1,13 @@
 # %% [markdown]
-# # Quartet model generation and interpolation
+# # String quartet MTVAE model for interpolation
+#
+# Author: Alex Kyllo
 
 # %%
-import os
 import sys
 from pathlib import Path
 
 sys.path.append("..")
-import numpy as np
 from musiclearn import config, processing, vae_models
 
 # %%
@@ -64,7 +64,7 @@ bee_8 = processing.score_to_array(bee_score.measures(0, 8), ticks_per_beat)
 moz_8 = processing.score_to_array(moz_score.measures(0, 8), ticks_per_beat)
 
 # %%
-beemoz = model.interpolate(bee_8, moz_8, 5, ticks_per_beat, beats_per_phrase)
+beemoz = model.interpolate(bee_8, moz_8, 5)
 for i, score in enumerate(beemoz):
     beemoz_score = processing.array_to_score(score, programs=programs, resolution=ticks_per_beat)
     beemoz_score.write("midi", f"../outputs/{exp_name}/beemoz_8_{i}.mid")
@@ -82,7 +82,7 @@ dvorak_8 = processing.score_to_array(dvorak_score.measures(0, 8), ticks_per_beat
 bach_8 = processing.score_to_array(bach_score.measures(0, 8), ticks_per_beat)
 
 # %%
-dvorbach = model.interpolate(dvorak_8, bach_8, 5, ticks_per_beat, beats_per_phrase)
+dvorbach = model.interpolate(dvorak_8, bach_8, 5)
 for i, score in enumerate(dvorbach):
     dvorbach_score = processing.array_to_score(score, programs=programs, resolution=ticks_per_beat)
     dvorbach_score.write("midi", f"../outputs/{exp_name}/dvorbach_8_{i}.mid")
@@ -91,7 +91,7 @@ for i, score in enumerate(dvorbach):
 # another example: Beethoven and Bach
 
 # %%
-beebach = model.interpolate(bee_8, bach_8, 5, ticks_per_beat, beats_per_phrase)
+beebach = model.interpolate(bee_8, bach_8, 5)
 for i, score in enumerate(beebach):
     beebach_score = processing.array_to_score(score, programs=programs, resolution=ticks_per_beat)
     beebach_score.write("midi", f"../outputs/{exp_name}/beebach_8_{i}.mid")
@@ -100,7 +100,7 @@ for i, score in enumerate(beebach):
[…]

### 2021-06-05 papers/final/alex/paper.tex
remove models folder, move vae.py out, consolidate CLIs
add interpolate command to CLI

@@
 
-In order to accommodate polyphonic music, we convert each MIDI file
-into a pianoroll representation, as visualized in Figure
-\ref{pianoroll}, wherein each instrument track is a sparse matrix that
-multi-hot encodes the velocity values for each of 128 possible pitch
-levels at each timestep. To reduce dimensionality, I will clip the
-note pitch values to a narrower range, excluding the rarely used notes
-in the extreme high and low octaves, and encode chords (combinations
-of notes played simultaneously on the same instrument) as distinct
-tokens, so that they can be one-hot encoded
+In order to accommodate polyphonic music, we each MIDI file is
+converted into a modified pianoroll representation, an example of
+which is visualized in Figure \ref{pianoroll}. In this modified
+representation, each instrument track is a sequence of integers from
+0-129 representing the MIDI note pitch value being played at the given
+time step, where values 0-127 are pitches, 128 is a rest (silence) and
+129 is sustain, indicating that the previously played pitch is held
+continuously. Because chords (multiple notes being played
+simultaneously by the same instrument) are relatively rare in string
+quartets, the lowest note of each chord is taken and the rest
+discarded, to reduce dimensionality. Because notes at the extreme low
+and high end are rare and are out of range of common instruments, the
+note pitch values are clipped to a narrower range and ordinal
+encoded. For the 36 string quartets found in the MusicNet database,
+this results in 66 distinct values, where 0-63 represent notes played
+by the violins, viola and cello, 64 is rest and 65 is sustain. Our
+software package also includes functions to convert this
+representation back into a Music21 Score object, which can then be
+written to a MIDI file.
 
@@
 
-The result of this preprocessing is that each training example will be
-a 2D matrix of shape (time steps x instrument tracks) and stacking the
-training examples will produce a 3D tensor. Information regarding which
+The result of this preprocessing is that each training example isbe a
+2D matrix of shape (time steps x instrument tracks) and stacking the
+training examples produces a 3D tensor. Information regarding which
 track corresponds to which musical instrument is lost and therefore
-must be maintained separately to reconstruct the MIDI representation.
+must be maintained separately to reconstruct the MIDI
+representation. For string quartets this instrument ensemble is always
+\texttt{[40, 40, 41, 42]}, which is an array of MIDI instrument codes
+representing two violins, one viola and one cello. Tempo information
+is also lost from this representation, so the outputs are normalized
+to the default tempo of 120 beats per minute.
 
@@
 
+The encoder model utilizes an embedding layer to encode the note pitch
+integers as vectors; embedding vector length is a tunable
+hyperparameter and we found the best results with it set to 8. Then
+two LSTM layers, with a dropout layer in between, convert the
+sequences into latent codes, which are modeled as a multidimensional
+Gaussian distribution.
+
+The decoder model samples from the latent space $z$ with the
+distribution parameterized by the $\mu$ and $log(\sigma)$ values
+learned by the encoder.
+
+TODO: model architecture plots
+
 \subsection{Model Evaluation}
@@
 
+TODO: learning curve plot
+
 \section{Discussion}

### 2021-06-05 musiclearn_cli.py

@@ -92,7 +92,7 @@ def fit_mtvae(
 @click.argument("output_dir", type=click.Path())
 @click.option("--n", type=click.INT, default=3)
 def interpolate(model_path, midi_1, midi_2, output_dir, n):
-    """Use MODEL_PATH to interpolate n"""
+    """Use MODEL_PATH to interpolate n points between MIDI_1 and MIDI_2."""
     programs = [40, 40, 41, 42]  # violin x2, viola, cello
     model = vae_models.MultiTrackVAE.from_saved(model_path)
     scores = [processing.midi_to_music21(f) for f in [midi_1, midi_2]]
@@ -102,6 +102,9 @@ def interpolate(model_path, midi_1, midi_2, output_dir, n):
         processing.array_to_score(arr, programs=programs, resolution=model.ticks_per_beat)
         for arr in output_arrays
     ]
+    LOG.info(
+        f"Generating {n} interpolations between {midi_1} and {midi_2} and saving to {output_dir}..."
+    )
     for i, score in enumerate(output_scores):
         score.write("midi", Path(output_dir) / f"interpolation_{i}.mid")
 
@@ -192,6 +195,7 @@ def cli():
 def main():
     """"""
     cli.add_command(fit_mtvae)
+    cli.add_command(interpolate)
     cli.add_command(fit_sequential)
     cli.add_command(generate_music)
     cli.add_command(plot_losses)

### 2021-06-05 run_experiments.sh

@@ -23,4 +23,4 @@
 #python musiclearn_cli.py fit-mtvae mtvae_0015 --beats-per-phrase 8 --lstm-units 128 --latent-dim 128 --dropout-rate 0.5 --patience 20
 #python musiclearn_cli.py fit-mtvae mtvae_0016 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20
 #python musiclearn_cli.py fit-mtvae mtvae_0017 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --bidirectional --patience 20
-python musiclearn_cli.py fit-mtvae mtvae_0018 --beats-per-phrase 8 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20
+#python musiclearn_cli.py fit-mtvae mtvae_0018 --beats-per-phrase 8 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20

### 2021-06-06 papers/final/alex/paper.tex
hparam tuning for vae model

@@
 %% \documentclass[manuscript,screen,review]{acmart}
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
+A survey of related works in generative modeling of polyphonic music
+turned up several models focused on multi-instrument pop/rock music
+arrangements or classical piano performances. In contrast, this
+project focuses on variational autoencoding and latent space
+interpolation of classical string quartets (arrangements for an
+ensemble of two violins, one viola and one cello) and is, to the
+author's knowledge, the first project to apply this modeling approach
+to this type of musical arrangement.
+
 \section{Related Work}
@@
 avoid the problem of ``posterior/mode collapse'' whereby a generative
-model learns to ignore its latent code and rely on autoregression
-\cite{roberts_hierarchical_2018}. This model is trained on 16-bar
-paragraphs of music and is capable of generating new melodies that
-blend two given melodies via latent space interpolation.
+model learns to ignore its latent codes and rely on autoregression to
+generate output sequences \cite{roberts_hierarchical_2018}. This model
+is trained on 16-bar paragraphs of music and is capable of generating
+new melodies that blend two given melodies together via latent space
+interpolation. Among the literature surveyed, MusicVAE's approach is
+the most similar to this work; in comparison our model is
+significantly simpler and works on a single bar of music at a time.
 
@@
 ratings from human listeners for its resemblance to human-composed
-music \cite{huang_music_2018}.
+music \cite{huang_music_2018}. In contrast to MusicVAE and this work,
+Music Transformer is capable of generating much longer sequences of
+music, but does so via sequence extrapolation rather than
+interpolation, so it can continue a given priming melody but does not
+blend multiple given melodies together. It also assumes a single
+instrument track (piano).
 
@@
 longer-term structural patterns.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{song_structure.png}
-  \caption{Diagram of the hierarchical structure of a musical
-    composition, from the MuseGAN paper \cite{dong2017musegan}.}
-  \label{song_structure}
-  \Description{A diagram showing the hierarchical structure of songs,
-    to paragraphs, to phrases, to bars, to, beats, to pixels.}
-\end{figure}
 
@@
 
+While there may be significant opportunity in future approaches that
+combining the two, contemporary research generally treats symbolic and
+audio music modeling as separate problem domains and this work does
+the same, choosing to focus on symbolic music.
+
 \section{Methods}
@@
 129 is sustain, indicating that the previously played pitch is held
-continuously. Because chords (multiple notes being played
-simultaneously by the same instrument) are relatively rare in string
-quartets, the lowest note of each chord is taken and the rest
-discarded, to reduce dimensionality. Because notes at the extreme low
-and high end are rare and are out of range of common instruments, the
-note pitch values are clipped to a narrower range and ordinal
-encoded. For the 36 string quartets found in the MusicNet database,
-this results in 66 distinct values, where 0-63 represent notes played
-by the violins, viola and cello, 64 is rest and 65 is sustain. Our
-software package also includes functions to convert this
-representation back into a Music21 Score object, which can then be
-written to a MIDI file.
+continuously. MIDI data contains pressure values that indicate how
+hard or soft each note is played; this information is discarded in our
+preprocessing and all notes are represented with constant
+pressure. Because chords (multiple notes being played simultaneously
+by the same instrument) are relatively rare in string quartets, the
+lowest note of each chord is taken and the rest discarded, to reduce
+dimensionality. Because notes at the extreme low and high end are rare
+and are out of range of common instruments, the note pitch values are
+clipped to a narrower range and ordinal encoded. For the 36 string
+quartets found in the MusicNet database, this results in 66 distinct
+values, where 0-63 represent notes played by the violins, viola and
+cello, 64 is rest and 65 is sustain. Our software package also
+includes functions to convert this representation back into a Music21
+Score object, which can then be written to a MIDI file.
 
@@
 examples, so we will crop songs into phrases of equal numbers of
-measures to use as training data.
+beats to use as training data.
 
@@
 The model is a recurrent variational autoencoder; both Long Short-Term
-Memory (LSTM) and Gated Recurrent Unit (GRU) cell types were tested.
+Memory (LSTM) and Gated Recurrent Unit (GRU) cell types were tested
+but this choice did not make a significant difference in the results.
 
@@
 distribution parameterized by the $\mu$ and $log(\sigma)$ values
-learned by the encoder.
+learned by the encoder, and then splits into four parallel output
+paths, one per instrument track. Each path contains again two LSTM
+layers with a dropout layer between them, and then a fully connected
+layer with softmax activation is used to select the most probable note
+value for the current track at the current time step.
+
+The autoencoder reconstruction loss function is sparse categorical
+cross-entropy, which is added to the Kullback-Leibler divergence of
+the approximate posterior distribution from a multivariate Gaussian
+distribution with mean 0 and standard deviation 1, resulting in the
+(negative) evidence lower bound (ELBO) loss, which is a standard
+variational autoencoder loss function. This loss function is applied
+to each track independently and then summed across all four tracks
+with equal weight.
 
 TODO: model architecture plots
+
+\subsection{Model Experimentation}
+
+The following hyperparameters were tuned, with tested values listed
+and best parameters in \textbf{bold}:
+
+\begin{itemize}
+  \tightlist
+  \item batch size: \textbf{32}, 64, 96
+  \item learning rate: 0.001, \textbf{0.0002}
+  \item recurrent layer type: LSTM, GRU (made no difference)
+  \item recurrent encoder layer directions: \textbf{unidirectional}, bidirectional
+  \item recurrent layer cells: 64, \textbf{128}, 256
+  \item latent code dimension: 128, 256, \textbf{512}
+  \item embedding dimension: 4, \textbf{8}, 16
+  \item dropout rate: 0.2, 0.4, \textbf{0.5}
+\end{itemize}
+
+Each model was trained for up to 500 epochs with an early stopping
+patience of 20 epochs. The model was cross-validated on 10\% of the
+training data after each epoch. The best model was selected by a
+combination of validation loss and listening to hand-picked sample
+reconstructions and interpolations.
+
+\section{Results}
+
+The reconstructed music produced by the model sounds reminiscent of
+the original inputs, preserving the key signature and general rhythmic
+and melodic structure, albeit with noticeable differences.
+
+A website with several samples of the original input MIDI tracks, the
+autoencoded outputs, and interpolations, is presented at: TODO
+
+TODO: learning curve plot
+
+TODO: links to audio samples
 
@@
 
-\section{Results}
-
-TODO: learning curve plot
-
 \section{Discussion}

### 2021-06-06 musiclearn/training.py

@@ -50,6 +50,7 @@ def log_end(exp_name, model_name, start_time):
 
 def train_mtvae(
     exp_name,
+    model_class,
     ticks_per_beat,
     beats_per_phrase,
     epochs,
@@ -66,7 +67,7 @@ def train_mtvae(
     """Train the MultiTrack VAE once with a given set of hyperparameters."""
     # String quartet MIDI programs
     x = processing.get_string_quartets(ticks_per_beat)
-    mtvae = vae_models.MultiTrackVAE(
+    mtvae = model_class(
         lstm_units,
         embedding_dim,
         latent_dim,
@@ -95,6 +96,7 @@ def train_mtvae(
         dropout_rate=dropout_rate,
         patience=patience,
         gru=gru,
+        bidirectional=bidirectional,
     )
     mtvae.train(x, ticks_per_beat, beats_per_phrase, epochs, callbacks=cbacks)
     log_end(exp_name, model_name, start_time)

### 2021-06-06 musiclearn/vae_models.py

@@ -144,6 +144,75 @@ def build_multi_track_vae(
     return vae_model, encoder_model, decoder_model
 
 
+def build_multi_track_split_vae(
+    optimizer,
+    lstm_units,
+    latent_dim,
+    embedding_dim,
+    n_timesteps,
+    n_tracks,
+    n_notes,
+    dropout_rate=0.2,
+    gru=False,
+    bidirectional=False,
+):
+    """Build a multi-track LSTM-VAE Keras model for autoencoding polyphonic music."""
+    # define encoder model
+    inputs = layers.Input(shape=(n_timesteps, n_tracks))
+    if gru:
+        rnn = layers.GRU
+    else:
+        rnn = layers.LSTM
+    if embedding_dim > 0:
+        encoder = layers.Embedding(n_notes, embedding_dim, input_length=n_timesteps)(inputs)
+        encoder = layers.Reshape((n_timesteps, embedding_dim * n_tracks))(encoder)
+        if bidirectional:
+            encoder = layers.Bidirectional(rnn(lstm_units, return_sequences=True))(encoder)
+        else:
+            encoder = rnn(lstm_units, return_sequences=True)(encoder)
+    else:
+        if bidirectional:
+            encoder = layers.Bidirectional(rnn(lstm_units, return_sequences=True))(inputs)
+        else:
+            encoder = rnn(lstm_units, return_sequences=True)(inputs)
+    encoder = layers.Dropout(dropout_rate)(encoder)
+    if bidirectional:
+        encoder = layers.Bidirectional(rnn(lstm_units, return_sequences=False))(encoder)
+    else:
+        encoder = rnn(lstm_units, return_sequences=False)(encoder)
+    mu = layers.Dense(latent_dim, name="mu")(encoder)
+    sigma = layers.Dense(latent_dim, name="sigma")(encoder)
+    # Latent space sampling
+    z = layers.Lambda(sample_normal, output_shape=(latent_dim,))([mu, sigma])
+    encoder_model = keras.Model(inputs, [mu, sigma, z])
+
+    # define decoder model
+    decoder_input = layers.Input(shape=(latent_dim,))
+    decoders = [layers.RepeatVector(n_timesteps)(decoder_input) for _ in n_tracks]
+    decoders = [rnn(lstm_units, return_sequences=True)(d) for d in decoders]
+    decoders = [layers.Dropout(dropout_rate)(d) for d in decoders]
+    decoders = [rnn(lstm_units, return_sequences=True)(d) for d in decoders]
+    outputs = [
[…]

### 2021-06-06 musiclearn_cli.py

@@ -3,6 +3,7 @@ Command line interface for the musiclearn project.
 Authors: Alex Kyllo and Jack Phan
 """
 import logging
+import os
 import random
 from pathlib import Path
 
@@ -70,6 +71,64 @@ def fit_mtvae(
     Author: Alex Kyllo"""
     training.train_mtvae(
         exp_name,
+        vae_models.MultiTrackVAE,
+        ticks_per_beat,
+        beats_per_phrase,
+        epochs,
+        batch_size,
+        learning_rate,
+        lstm_units,
+        latent_dim,
+        embedding_dim,
+        dropout_rate,
+        gru,
+        bidirectional,
+        patience,
+    )
+
+
+@click.command()
+@click.argument("exp_name", type=click.STRING)
+@click.option("--ticks-per-beat", type=click.INT, default=4, help="Time steps per quarter note.")
+@click.option("--beats-per-phrase", type=click.INT, default=4, help="Quarter notes per phrase.")
+@click.option("--epochs", type=click.INT, default=500, help="The training batch size.")
+@click.option("--batch-size", type=click.INT, default=32, help="The training batch size.")
+@click.option(
+    "--learning-rate", type=click.FLOAT, default=0.0002, help="The optimizer learning rate."
+)
+@click.option("--lstm-units", type=click.INT, default=256, help="Number of LSTM units per layer.")
+@click.option("--latent-dim", type=click.INT, default=100, help="The latent vector dimension.")
+@click.option("--embedding-dim", type=click.INT, default=8, help="The note embedding dimension.")
+@click.option(
+    "--dropout-rate", type=click.FLOAT, default=0.4, help="The dropout rate between LSTM layers"
+)
+@click.option("--gru/--lstm", default=False, help="Use GRU layer instead of LSTM.")
+@click.option(
+    "--bidirectional/--unidirectional",
+    default=False,
+    help="Use bidirectional LSTM layer in encoder.",
+)
+@click.option("--patience", type=click.INT, default=10, help="The early stopping patience.")
+def fit_mtsvae(
+    exp_name,
+    ticks_per_beat,
+    beats_per_phrase,
+    epochs,
+    batch_size,
+    learning_rate,
+    lstm_units,
+    latent_dim,
+    embedding_dim,
+    dropout_rate,
+    gru,
+    bidirectional,
+    patience,
+):
[…]

### 2021-06-06 run_experiments.sh

@@ -24,3 +24,17 @@
 #python musiclearn_cli.py fit-mtvae mtvae_0016 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20
 #python musiclearn_cli.py fit-mtvae mtvae_0017 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --bidirectional --patience 20
 #python musiclearn_cli.py fit-mtvae mtvae_0018 --beats-per-phrase 8 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20
+#python musiclearn_cli.py fit-mtvae mtsvae_0001 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20
+#python musiclearn_cli.py fit-mtvae mtsvae_0002 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --learning-rate 0.001 --patience 20
+#python musiclearn_cli.py fit-mtvae mtsvae_0003 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --batch-size 96
+#python musiclearn_cli.py fit-mtvae mtsvae_0004 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --batch-size 64
+#python musiclearn_cli.py fit-mtvae mtsvae_0005 --beats-per-phrase 8 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20
+#python musiclearn_cli.py fit-mtvae mtsvae_0006 --beats-per-phrase 4 --lstm-units 128 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --gru
+#python musiclearn_cli.py fit-mtvae mtsvae_0007 --beats-per-phrase 4 --lstm-units 64 --latent-dim 128 --embedding-dim 8 --dropout-rate 0.5 --patience 20
[…]

### 2021-06-06 papers/final/alex/paper.tex
data augmentation for vae model

@@
 \documentclass[sigconf,authorversion]{acmart}
+\usepackage{pgf}
+\usepackage{pgfplots}
 %% NOTE that a single column version may be required for
@@
 converted into a modified pianoroll representation, an example of
-which is visualized in Figure \ref{pianoroll}. In this modified
-representation, each instrument track is a sequence of integers from
-0-129 representing the MIDI note pitch value being played at the given
-time step, where values 0-127 are pitches, 128 is a rest (silence) and
-129 is sustain, indicating that the previously played pitch is held
-continuously. MIDI data contains pressure values that indicate how
-hard or soft each note is played; this information is discarded in our
-preprocessing and all notes are represented with constant
-pressure. Because chords (multiple notes being played simultaneously
-by the same instrument) are relatively rare in string quartets, the
-lowest note of each chord is taken and the rest discarded, to reduce
-dimensionality. Because notes at the extreme low and high end are rare
-and are out of range of common instruments, the note pitch values are
-clipped to a narrower range and ordinal encoded. For the 36 string
-quartets found in the MusicNet database, this results in 66 distinct
-values, where 0-63 represent notes played by the violins, viola and
-cello, 64 is rest and 65 is sustain. Our software package also
-includes functions to convert this representation back into a Music21
-Score object, which can then be written to a MIDI file.
+which is visualized in Figure \ref{pianoroll}. While the standard
+pianoroll represents each time-step as a multi-hot encoded vector of
+note values, because most notes are not being played at most timesteps
+there is an extreme class imbalance and sparsity in the matrix. This
+is reported in \cite{kumar2019polyphonic} to cause problems in model
+fitting and metrics calculation, so they introduce an alternative
+``multi-stream'' representation of the pianoroll, which we closely,
+(but not exactly) emulate in this work.
+
+In our modified representation, each instrument track is a sequence of
+integers from 0-129 representing the MIDI note pitch value being
+played at the given time step, where values 0-127 are pitches, 128 is
+a rest (silence) and 129 is sustain, indicating that the previously
+played pitch is held continuously. MIDI data contains pressure values
+that indicate how hard or soft each note is played; this information
+is discarded in our preprocessing and all notes are represented with
+constant pressure. Because chords (multiple notes being played
+simultaneously by the same instrument) are relatively rare in string
+quartets, the lowest note of each chord is taken and the rest
+discarded, to reduce dimensionality. Because notes at the extreme low
+and high end are rare and are out of range of common instruments, the
+note pitch values are clipped to a narrower range and ordinal
+encoded. For the 36 string quartets found in the MusicNet database,
+this results in 66 distinct values, where 0-63 represent notes played
+by the violins, viola and cello, 64 is rest and 65 is sustain. Our
+software package also includes functions to convert this
+representation back into a Music21 Score object, which can then be
+written to a MIDI file.
+
+TODO: plot histogram of notes per instrument
 
@@
 
-Data augmentation is also possible and we assess its impact on
-the results--the literature suggests augmentation via pitch shifting
-each training example up or down by up to six semitones, and
-increasing or reducing the speed by up to 10\% in order to create
-additional training examples and reduce overfitting
-\cite{oore_this_2018}.
-
-The multi-stream representation discussed in
-\cite{kumar2019polyphonic} addresses the issue of extreme class
-imbalance and sparsity in the pianoroll representation.
-
 \subsection{Model Design}
@@
 distribution with mean 0 and standard deviation 1, resulting in the
-(negative) evidence lower bound (ELBO) loss, which is a standard
-variational autoencoder loss function. This loss function is applied
-to each track independently and then summed across all four tracks
-with equal weight.
+(negative) evidence lower bound (ELBO) loss (Equation \ref{nelbo}),
+which is a standard variational autoencoder loss function. This loss
+function is applied to each track independently and then summed across
+all four tracks with equal weight.
+
+\begin{equation}
+  \label{nelbo}
+\mathbb{E}[log{p_\theta}(x|z)]-\mathbb{KL}(q_\lambda(z|x)||p(z))
+\end{equation}
+
+where $\lambda$ and $\theta$ represent the encoder and decoder
+parameters respectively, while $q_\lambda(z|x)$ represents the
+encoder, which approximates the true posterior probability $p(z|x)$,
+and $p_\theta(x|z)$ represents the decoder, which parameterizes the
+likelihood $p(x|z)$ \cite{roberts_hierarchical_2018}.
 
@@
   \item dropout rate: 0.2, 0.4, \textbf{0.5}
+  \item time steps per beat (quarter note): \textbf{4}, 8
+  \item beats per phrase: \textbf{4}, 8
 \end{itemize}
@@
 
+Data augmentation was also tested--\cite{oore_this_2018} suggests
+augmentation via pitch shifting each training example up or down by up
+to six semitones in order to create additional training examples and
+reduce overfitting, so we followed that recommendation and used it to
+double the size of the training dataset. As a result, we saw much
+lower validation losses (compare Figure \ref{loss} and Figure
+\ref{loss_augment} and a small but noticeable improvement in the
+fidelity of the song reconstructions, suggesting that this method was
+effective in reducing overfitting.
+
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.5}{\input{figures/loss.pgf}}
+    \end{center}
+    \caption{Training and validation loss for the best performing model - without data augmentation}
+    \label{loss}
+\end{figure}
+
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.5}{\input{figures/loss_augment.pgf}}
+    \end{center}
+    \caption{Training and validation loss for the best performing model - with data augmentation}
+    \label{loss_augment}
+\end{figure}
+
 \section{Results}
@@
 the original inputs, preserving the key signature and general rhythmic
-and melodic structure, albeit with noticeable differences.
+and melodic structure, albeit with noticeable differences at the
+individual note level.
 
 A website with several samples of the original input MIDI tracks, the
-autoencoded outputs, and interpolations, is presented at: TODO
-
-TODO: learning curve plot
-
-TODO: links to audio samples
+autoencoded outputs, and interpolations, is presented at: \textbf{TODO}
 

### 2021-06-06 musiclearn/plotting.py

@@ -28,4 +28,5 @@ def plot_learning_curves(history_file: os.PathLike, dest_path: os.PathLike):
     plt.savefig(
         dest_path,
     )
+
     return (fig, ax)

### 2021-06-06 musiclearn/processing.py

@@ -271,3 +271,16 @@ def get_string_quartets(ticks_per_beat: int):
         x = np.vstack([score_to_array(score, resolution=ticks_per_beat) for score in scores])
         np.save(f, x)
     return x
+
+
+def random_transpose(arr):
+    """Pitch shift the array up or down by +/- 6 semitones"""
+    shift = np.random.choice(range(1, 7))
+    return arr + shift * ~np.isin(arr, [REST, SUSTAIN])
+
+
+def augment(arr):
+    """Random transpose each training example and append to the original dataset."""
+    augs = np.vectorize(random_transpose)(arr)
+    arr = np.vstack([arr, augs])
+    return arr

### 2021-06-06 musiclearn/training.py

@@ -63,10 +63,13 @@ def train_mtvae(
     gru,
     bidirectional,
     patience=10,
+    augment=False,
 ):
     """Train the MultiTrack VAE once with a given set of hyperparameters."""
     # String quartet MIDI programs
     x = processing.get_string_quartets(ticks_per_beat)
+    if augment:
+        x = processing.augment(x)
     mtvae = model_class(
         lstm_units,
         embedding_dim,
@@ -97,6 +100,7 @@ def train_mtvae(
         patience=patience,
         gru=gru,
         bidirectional=bidirectional,
+        augment=augment,
     )
     mtvae.train(x, ticks_per_beat, beats_per_phrase, epochs, callbacks=cbacks)
     log_end(exp_name, model_name, start_time)

### 2021-06-06 musiclearn/vae_models.py

@@ -188,7 +188,7 @@ def build_multi_track_split_vae(
 
     # define decoder model
     decoder_input = layers.Input(shape=(latent_dim,))
-    decoders = [layers.RepeatVector(n_timesteps)(decoder_input) for _ in n_tracks]
+    decoders = [layers.RepeatVector(n_timesteps)(decoder_input) for _ in range(n_tracks)]
     decoders = [rnn(lstm_units, return_sequences=True)(d) for d in decoders]
     decoders = [layers.Dropout(dropout_rate)(d) for d in decoders]
     decoders = [rnn(lstm_units, return_sequences=True)(d) for d in decoders]

### 2021-06-06 musiclearn_cli.py

@@ -51,6 +51,11 @@ def set_seeds(seed):
     default=False,
     help="Use bidirectional LSTM layer in encoder.",
 )
+@click.option(
+    "--augment/--no-augment",
+    default=False,
+    help="Augment the training set with random pitch shifts.",
+)
 @click.option("--patience", type=click.INT, default=10, help="The early stopping patience.")
 def fit_mtvae(
     exp_name,
@@ -66,6 +71,7 @@ def fit_mtvae(
     gru,
     bidirectional,
     patience,
+    augment,
 ):
     """Run MultiTrackVAE experiment named EXP_NAME with hyperparameter options.
     Author: Alex Kyllo"""
@@ -84,6 +90,7 @@ def fit_mtvae(
         gru,
         bidirectional,
         patience,
+        augment,
     )
 
 
@@ -109,6 +116,11 @@ def fit_mtvae(
     help="Use bidirectional LSTM layer in encoder.",
 )
 @click.option("--patience", type=click.INT, default=10, help="The early stopping patience.")
+@click.option(
+    "--augment/--no-augment",
+    default=False,
+    help="Augment the training set with random pitch shifts.",
+)
 def fit_mtsvae(
     exp_name,
     ticks_per_beat,
@@ -123,6 +135,7 @@ def fit_mtsvae(
     gru,
     bidirectional,
     patience,
+    augment,
 ):
     """Run MultiTrackSplitVAE experiment named EXP_NAME with hyperparameter options.
     Author: Alex Kyllo"""
@@ -141,6 +154,7 @@ def fit_mtsvae(
         gru,
         bidirectional,
         patience,
+        augment,
     )
 
 
@@ -269,6 +283,7 @@ def generate_music(output_name, data_path, model_type, weights_path, num_notes):
 @click.argument("output_file", type=click.Path())
 def plot_losses(history_file, output_file):
     """Plot model training & validation loss curves from HISTORY_FILE and save to OUTPUT_FILE."""
+    os.makedirs(os.path.dirname(output_file), exist_ok=True)
     plotting.plot_learning_curves(history_file, output_file)
 
 

### 2021-06-06 run_experiments.sh

@@ -37,4 +37,10 @@
 #python musiclearn_cli.py fit-mtvae mtsvae_0011 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 4 --dropout-rate 0.5 --patience 20
 #python musiclearn_cli.py fit-mtvae mtsvae_0012 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 16 --dropout-rate 0.5 --patience 20
 #python musiclearn_cli.py fit-mtvae mtsvae_0013 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 32 --dropout-rate 0.5 --patience 20
-python musiclearn_cli.py fit-mtvae mtsvae_0014 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --bidirectional
+#python musiclearn_cli.py fit-mtvae mtsvae_0014 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --bidirectional
+#python musiclearn_cli.py fit-mtsvae mtsvae_0015 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 20
+#python musiclearn_cli.py fit-mtvae mtvae_0019 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --augment
+#python musiclearn_cli.py fit-mtsvae mtsvae_0016 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 10 --augment --epochs 100 --batch-size 128 --learning-rate 0.0005
+#python musiclearn_cli.py fit-mtsvae mtsvae_0017 --beats-per-phrase 8 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --augment --patience 10 --batch-size 128
+#python musiclearn_cli.py fit-mtvae mtvae_0020 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --batch-size 64 --learning-rate 0.0004 --augment
[…]

### 2021-06-06 musiclearn/plotting.py
add pitch class histogram

@@ -6,7 +6,11 @@ import os
 
 import matplotlib
 import matplotlib.pyplot as plt
+import numpy as np
 import pandas as pd
+import seaborn as sns
+
+from musiclearn.processing import REST, SUSTAIN
 
 
 def plot_learning_curves(history_file: os.PathLike, dest_path: os.PathLike):
@@ -25,8 +29,23 @@ def plot_learning_curves(history_file: os.PathLike, dest_path: os.PathLike):
     ax.set_ylim(0.0)
     ax.legend(loc="best")
     fig.tight_layout()
+    os.makedirs(os.path.dirname(dest_path), exist_ok=True)
     plt.savefig(
         dest_path,
     )
 
     return (fig, ax)
+
+
+def plot_pitch_class_histogram(x: np.array, dest_path: os.PathLike, by_instrument: bool = True):
+    """Plot a histogram of pitches"""
+    if dest_path.endswith(".pgf"):
+        matplotlib.use("pgf")
+    xp = np.where(np.isin(x, [REST, SUSTAIN]), np.nan, x)
+    tracks = ["violin 1", "violin 2", "viola", "cello"]
+    df = pd.DataFrame(xp, columns=tracks)
+    ax = sns.histplot(data=df)
+    ax.set_xlabel("MIDI pitch value")
+    os.makedirs(os.path.dirname(dest_path), exist_ok=True)
+    plt.savefig(dest_path)
+    return ax

### 2021-06-06 scripts/midi_quartet_training.py
delete outdated notebook scripts

@@ -1,105 +1,0 @@
-# %% [Markdown]
-# # MIDI MusicNet string quartet model exploration with pypianoroll representation
-#
-# Alex Kyllo
-# May 9, 2021
-#
-# An exploratory notebook for playing with model architecture ideas.
-
-# %%
-import itertools
-import os
-import sys
-
-import joblib
-import numpy as np
-import tensorflow as tf
-from sklearn.preprocessing import OrdinalEncoder
-from tensorflow.keras import optimizers
-
-sys.path.append("..")
-from musiclearn import processing, vae_models
-
-# %%
-physical_devices = tf.config.list_physical_devices("GPU")
-tf.config.experimental.set_memory_growth(physical_devices[0], True)
-
-
-# %% [markdown]
-# Get all string quartets from MusicNet dataset
-
-# %%
-f = "../data/quartets_music21_4.npy"
-program_ids = [40, 40, 41, 42]
-ticks_per_beat = 4
-if os.path.isfile(f):
-    x = np.load(f, allow_pickle=True)
-else:
-    scores_file = "../data/quartets.joblib"
-    if os.path.isfile(scores_file):
-        scores, fnames = joblib.load(scores_file)
-    else:
-        # Read all MusicNet string quartets
-        scores, fnames = processing.musicnet_quartets_to_music21(program_ids=program_ids)
-        joblib.dump((scores, fnames), scores_file)
-    # Stack them into one big 2D array
-    x = np.vstack([processing.score_to_array(score, resolution=ticks_per_beat) for score in scores])
-    np.save(f, x)
-
-# %%
-notes = np.unique(x)
-n_notes = notes.shape[0]
-enc = OrdinalEncoder(categories=list(itertools.repeat(notes, 4)))
-x = enc.fit_transform(x).astype(int)
-REST_CODE = np.argwhere(enc.categories_[0] == processing.REST)[0][0]
-
-# %%
-x = processing.split_array(x, beats_per_phrase=4, resolution=ticks_per_beat, fill=REST_CODE)
-n_timesteps = x.shape[1]
-n_features = x.shape[2]
-shape = (n_timesteps, n_features)
-
-# %% [markdown]
-# Remove the phrases that consist only of rests
-
-# %%
[…]

### 2021-06-06 scripts/quartet_hierarchical.py

@@ -1,25 +1,0 @@
-import numpy as np
-from tensorflow.keras import layers, models
-
-latent_dim = 128
-state_size = 128
-z = np.zeros((1, latent_dim))
-
-# The latent vector z is passed through a fully connected layer with tanh activation
-# to set the initial state of a "conductor" RNN which produces one embedding vector
-# per subsequence (measure). First conductor is given zero inputs and its h, c states
-# are passed to subsequent conductors.
-# Each embedding vector is individually pass through a shared dense layer with tanh activation
-# to produce initial states for a final bottom-layer decoder rnn with softmax activation.
-# At each step of the bottom level decoder, the current embedding is concatenated with the
-# previous output to be used as the input (note: conditioning)
-# paraphrased from MusicVAE paper: https://arxiv.org/pdf/1803.05428.pdf
-# Explanation: https://medium.com/@musicvaeubcse/musicvae-understanding-of-the-googles-work-for-interpolating-two-music-sequences-621dcbfa307c
-
-
-decoder_input = layers.Input(shape=(latent_dim,))
-decoder = layers.Dense(latent_dim * state_size * 4, activation="tanh")(decoder_input)
-decoder, state_h, state_c = layers.LSTM(
-    latent_dim, initial_state=[], return_sequences=True, return_state=True
-)(decoder)
-# TODO

### 2021-06-06 papers/final/alex/paper.tex
add pitch class histogram
add script to run interpolation and convert to wav

@@
 this results in 66 distinct values, where 0-63 represent notes played
-by the violins, viola and cello, 64 is rest and 65 is sustain. Our
-software package also includes functions to convert this
-representation back into a Music21 Score object, which can then be
-written to a MIDI file.
+by the violins, viola and cello, 64 is rest and 65 is sustain. Figure
+\ref{pitch_dist} depicts the distribution of these pitch values per
+instrument. Our software package also includes functions to convert
+this representation back into a Music21 Score object, which can then
+be written to a MIDI file.
 
-TODO: plot histogram of notes per instrument
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.6}{\input{figures/pitch_dist.pgf}}
+    \end{center}
+    \caption{Distribution of pitch values per instrument in the MusicNet string quartets}
+    \label{pitch_dist}
+\end{figure}
 
@@
   \centering
-  \includegraphics[width=\linewidth]{first_bar.png}
+  \includegraphics[width=\linewidth]{figures/first_bar.png}
   \caption{Pianoroll visualization of the first measure of
@@
   \centering
-  \includegraphics[width=\linewidth]{first_bar_sheet.png}
+  \includegraphics[width=\linewidth]{figures/first_bar_sheet.png}
   \caption{Sheet music of the first measure of Beethoven's Serioso
@@
 sequences into latent codes, which are modeled as a multidimensional
-Gaussian distribution.
+Gaussian distribution $z$, produced by the encoder output Lambda layer
+(Figure \ref{decoder}).
 
@@
 distribution parameterized by the $\mu$ and $log(\sigma)$ values
-learned by the encoder, and then splits into four parallel output
-paths, one per instrument track. Each path contains again two LSTM
-layers with a dropout layer between them, and then a fully connected
-layer with softmax activation is used to select the most probable note
-value for the current track at the current time step.
+learned by the encoder, and then utilizes two LSTM layers with a
+dropout layer between them to generate sequences from the latent
+codes. Then the network splits into four parallel outputs, one per
+instrument track, where a fully connected layer with softmax
+activation is used to select the most probable note value for the
+the current track at the each time step (Figure \ref{decoder}).
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{figures/encoder.png}
+  \caption{A directed graph diagram of the encoder network.}
+  \label{encoder}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{figures/decoder.png}
+  \caption{A directed graph diagram of the decoder network (only one
+    of four parallel output layers shown).}
+  \label{decoder}
+\end{figure}
 
@@
 likelihood $p(x|z)$ \cite{roberts_hierarchical_2018}.
-
-TODO: model architecture plots
 

### 2021-06-06 musiclearn/vae_models.py

@@ -384,48 +384,3 @@ class MultiTrackVAE:
             x = self.ord_enc.inverse_transform(x)
             results.append(x)
         return results
-
-
-class MultiTrackSplitVAE(MultiTrackVAE):
-    """Split LSTM version of MultiTrackVAE."""
-
-    def train(self, x, ticks_per_beat, beats_per_phrase, epochs, callbacks=None):
-        """Train the model on a dataset."""
-        # Dataset prep, ordinal encoding
-        self.ticks_per_beat = ticks_per_beat
-        self.beats_per_phrase = beats_per_phrase
-        notes = np.unique(x)
-        self.n_notes = notes.shape[0]
-        self.n_tracks = x.shape[1]
-        self.ord_enc = OrdinalEncoder(categories=list(itertools.repeat(notes, self.n_tracks)))
-        x = self.ord_enc.fit_transform(x).astype(int)
-        self.rest_code = np.argwhere(self.ord_enc.categories_[0] == processing.REST)[0][0]
-        # Split songs into phrases
-        x = processing.split_array(
-            x, beats_per_phrase=beats_per_phrase, resolution=ticks_per_beat, fill=self.rest_code
-        )
-        self.n_timesteps = x.shape[1]
-        # Remove phrases that are only rests
-        all_rests = self.rest_code * self.n_timesteps * self.n_tracks
-        x = x[x.sum(axis=(1, 2)) != all_rests]
-        self.vae_model, self.encoder_model, self.decoder_model = build_multi_track_split_vae(
-            self.optimizer,
-            self.lstm_units,
-            self.latent_dim,
-            self.embedding_dim,
-            self.n_timesteps,
-            self.n_tracks,
-            self.n_notes,
-            self.dropout_rate,
-        )
-        self.history = self.vae_model.fit(
-            x,
-            tf.unstack(x, axis=2),
-            batch_size=self.batch_size,
-            epochs=epochs,
-            validation_split=0.1,
-            callbacks=callbacks,
-            initial_epoch=self.trained_epochs,
-        )
-        self.trained_epochs = self.trained_epochs + epochs
-        return self

### 2021-06-06 musiclearn_cli.py

@@ -95,70 +95,6 @@ def fit_mtvae(
 
 
 @click.command()
-@click.argument("exp_name", type=click.STRING)
-@click.option("--ticks-per-beat", type=click.INT, default=4, help="Time steps per quarter note.")
-@click.option("--beats-per-phrase", type=click.INT, default=4, help="Quarter notes per phrase.")
-@click.option("--epochs", type=click.INT, default=500, help="The training batch size.")
-@click.option("--batch-size", type=click.INT, default=32, help="The training batch size.")
-@click.option(
-    "--learning-rate", type=click.FLOAT, default=0.0002, help="The optimizer learning rate."
-)
-@click.option("--lstm-units", type=click.INT, default=256, help="Number of LSTM units per layer.")
-@click.option("--latent-dim", type=click.INT, default=100, help="The latent vector dimension.")
-@click.option("--embedding-dim", type=click.INT, default=8, help="The note embedding dimension.")
-@click.option(
-    "--dropout-rate", type=click.FLOAT, default=0.4, help="The dropout rate between LSTM layers"
-)
-@click.option("--gru/--lstm", default=False, help="Use GRU layer instead of LSTM.")
-@click.option(
-    "--bidirectional/--unidirectional",
-    default=False,
-    help="Use bidirectional LSTM layer in encoder.",
-)
-@click.option("--patience", type=click.INT, default=10, help="The early stopping patience.")
-@click.option(
-    "--augment/--no-augment",
-    default=False,
-    help="Augment the training set with random pitch shifts.",
-)
-def fit_mtsvae(
-    exp_name,
-    ticks_per_beat,
-    beats_per_phrase,
-    epochs,
-    batch_size,
-    learning_rate,
-    lstm_units,
-    latent_dim,
-    embedding_dim,
-    dropout_rate,
-    gru,
-    bidirectional,
-    patience,
-    augment,
-):
-    """Run MultiTrackSplitVAE experiment named EXP_NAME with hyperparameter options.
-    Author: Alex Kyllo"""
-    training.train_mtvae(
-        exp_name,
-        vae_models.MultiTrackSplitVAE,
-        ticks_per_beat,
-        beats_per_phrase,
-        epochs,
-        batch_size,
-        learning_rate,
-        lstm_units,
-        latent_dim,
-        embedding_dim,
-        dropout_rate,
-        gru,
-        bidirectional,
[…]

### 2021-06-06 run_experiments.sh

@@ -43,4 +43,5 @@
 #python musiclearn_cli.py fit-mtsvae mtsvae_0016 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 10 --augment --epochs 100 --batch-size 128 --learning-rate 0.0005
 #python musiclearn_cli.py fit-mtsvae mtsvae_0017 --beats-per-phrase 8 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --augment --patience 10 --batch-size 128
 #python musiclearn_cli.py fit-mtvae mtvae_0020 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --batch-size 64 --learning-rate 0.0004 --augment
-python musiclearn_cli.py fit-mtvae mtvae_0021 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --augment
+#python musiclearn_cli.py fit-mtvae mtvae_0021 --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --augment
+python musiclearn_cli.py fit-mtvae mtvae --beats-per-phrase 4 --lstm-units 128 --latent-dim 512 --embedding-dim 8 --dropout-rate 0.5 --patience 20 --augment

### 2021-06-06 scripts/midi_quartet_inference.py

@@ -1,132 +1,0 @@
-# %% [markdown]
-# # String quartet MTVAE model for interpolation
-#
-# Author: Alex Kyllo
-
-# %%
-import sys
-from pathlib import Path
-
-sys.path.append("..")
-from musiclearn import config, processing, vae_models
-
-# %%
-midis = Path(config.MUSICNET_MIDI_DIR)
-beethoven_2494 = midis / "Beethoven" / "2494_qt11_1.mid"
-mozart_1788 = midis / "Mozart" / "1788_kv_465_1.mid"
-
-# %%
-ticks_per_beat = 4
-beats_per_phrase = 4
-programs = [40, 40, 41, 42]  # Violin x2, Viola, Cello
-
-# %%
-bee_score = processing.midi_to_music21(beethoven_2494)
-bee = processing.score_to_array(bee_score, ticks_per_beat)
-moz_score = processing.midi_to_music21(mozart_1788)
-moz = processing.score_to_array(moz_score, ticks_per_beat)
-
-# %%
-exp_name = "mtvae_0016"
-exp_time = "2021-05-31T23:01:43"
-saved_model_path = f"../experiments/{exp_name}/{exp_time}/saved_model"
-model = vae_models.MultiTrackVAE.from_saved(saved_model_path)
-
-# %% [markdown]
-# ## Reconstruction
-#
-# Use the model to reconstruct an entire piece of music
-
-# %%
-bee_reconst = model.reconstruct(bee, ticks_per_beat, beats_per_phrase)
-bee_reconst_score = processing.array_to_score(
-    bee_reconst, programs=programs, resolution=ticks_per_beat
-)
-
-bee_reconst_score.write("midi", f"../outputs/{exp_name}/2494_qt11_1_reconst.mid")
-
-# %%
-moz_reconst = model.reconstruct(moz, ticks_per_beat, beats_per_phrase)
-moz_reconst_score = processing.array_to_score(
-    moz_reconst, programs=programs, resolution=ticks_per_beat
-)
-moz_reconst_score.write("midi", f"../outputs/{exp_name}/1788_kv_465_1_reconst.mid")
-
-# %% [markdown]
-# ## Interpolation
-#
-# Use the model to blend two pieces of music by taking the mean of their
-# latent codes
-
-# %%
-# Get first 8 measures of each
-bee_8 = processing.score_to_array(bee_score.measures(0, 8), ticks_per_beat)
[…]

### 2021-06-06 scripts/string_quartet_inference.py

@@ -1,0 +1,122 @@
+# %% [markdown]
+# # String quartet MTVAE model for interpolation
+#
+# Author: Alex Kyllo
+# Date: 2021-06-06
+
+# %% [markdown]
+# This notebook demonstrates string quartet music interpolation
+# with a trained Multi-Track Variational Autoencoder (MTVAE) model.
+
+# %%
+import os
+import sys
+from pathlib import Path
+
+import numpy as np
+
+sys.path.append("..")
+from musiclearn import config, processing, vae_models
+
+# %% [markdown]
+# Check the `musicnet_midis` directory path
+# and show the 36 string quartet MIDI filenames:
+
+# %%
+midis = Path(config.MUSICNET_MIDI_DIR)
+str(midis)
+
+# %%
+sq = processing.STRING_QUARTETS
+sq
+
+# %% [markdown]
+# Configure the string quartet MIDI program numbers and
+# beat/measure resolution and length
+
+# %%
+ticks_per_beat = 4
+beats_per_phrase = 4
+programs = [40, 40, 41, 42]  # Violin x2, Viola, Cello
+num_measures = 16  # length of each sample in measures
+num_interpolations = 5  # Number of interpolations per pair
+
+# %% [markdown]
+# Load the best model from training experiments
+
+# %%
+exp_name = "mtvae_0021"
+exp_time = "2021-06-06T09:32:32"
+saved_model_path = f"../experiments/{exp_name}/{exp_time}/saved_model"
+model = vae_models.MultiTrackVAE.from_saved(saved_model_path)
+
+# %% [markdown]
+# To test the model's ability to reconstruct its inputs and
+# interpolate new music between them, we will create 18 pairs
+# from the 36 original tracks and process them with the
+# model's `interpolate` method.
+
+# %%
+# zip the list of filenames together to make pairs
+half = len(sq) // 2
+left = sq[0:half]
+right = sq[half : len(sq)]
+pairs = list(zip(left, reversed(right)))
+pairs
+
+# %% [markdown]
+# Now we'll use the model to reconstruct each of the pairs
+# and use linear interpolatation to generate 3 vectors in between them in the latent space.
+#
+# We'll truncate each track to the first 16 measures to save time and space.
+
+# %%
+output_dir = Path(f"../outputs/{exp_name}")
+
+# %%
+for pair in pairs:
+    print(f"Interpolating between {pair[0]} and {pair[1]}...")
+    pairname = "_".join(
[…]

### 2021-06-06 scripts/string_quartet_inference.py
save string quartet inference notebook

@@ -48,6 +48,8 @@ num_interpolations = 5  # Number of interpolations per pair
 exp_name = "mtvae_0021"
 exp_time = "2021-06-06T09:32:32"
 saved_model_path = f"../experiments/{exp_name}/{exp_time}/saved_model"
+
+# %%
 model = vae_models.MultiTrackVAE.from_saved(saved_model_path)
 
 # %% [markdown]
@@ -65,8 +67,10 @@ pairs = list(zip(left, reversed(right)))
 pairs
 
 # %% [markdown]
+# ## Interpolation
+#
 # Now we'll use the model to reconstruct each of the pairs
-# and use linear interpolatation to generate 3 vectors in between them in the latent space.
+# and use linear interpolation to generate 3 vectors in between them in the latent space.
 #
 # We'll truncate each track to the first 16 measures to save time and space.
 
@@ -102,12 +106,20 @@ for pair in pairs:
     # write the interpolations to disk as MusicXML format (for sheet music printing)
 
 # %% [markdown]
-# ## Convert MIDI files to WAV so we can listen to them anywhere without synthesizer software.
+# ## Convert MIDI files to WAV
+#
+# This way we can listen to them anywhere without synthesizer software.
+#
+# Requires the [fluidsynth](https://www.fluidsynth.org/) library installed on the system
+# and a sound font such as the
+# [Fluid Release 3 General-MIDI Soundfont](https://member.keymusician.com/Member/FluidR3_GM/index.html)
 
+# %%
+SOUND_FONT = "/usr/share/sounds/sf2/FluidR3_GM.sf2"
 # %%
 import midi2audio
 
-fs = midi2audio.FluidSynth(sound_font="/usr/share/sounds/sf2/FluidR3_GM.sf2")
+fs = midi2audio.FluidSynth(sound_font=SOUND_FONT)
 
 # %%
 for pair in pairs:
@@ -120,3 +132,51 @@ for pair in pairs:
         wav = mid.with_suffix(".wav")
         fs.midi_to_audio(mid, wav)
         print(f"{mid} converted to {wav}")
+
+# %% [markdown]
+# ## Listen to WAV files
+#
+# using the IPython Audio widget
+
+# %%
+from IPython.display import Audio
+
+# %%
+wavs = sorted(list(output_dir.rglob("*.wav")))
+
+
+# %% [markdown]
[…]

### 2021-06-07 papers/final/alex/paper.tex
add gdrive link to paper

@@
 interpolation. Among the literature surveyed, MusicVAE's approach is
-the most similar to this work; in comparison our model is
-significantly simpler and works on a single bar of music at a time.
+the most similar to this work; in comparison MusicVAE dedicates
+significant effort to devising an architecture that allows the model
+to generate very long sequences of notes, which is beneficial for
+other generation tasks but not strictly necessary for
+interpolation. Our model is significantly simpler and demonstrates
+that interpolation works well even on a single bar of music at a time.
 
@@
 
-A website with several samples of the original input MIDI tracks, the
-autoencoded outputs, and interpolations, is presented at: \textbf{TODO}
+A collection of several samples of the autoencoded outputs and
+interpolations, is presented at:
+\href{https://drive.google.com/drive/folders/12o04uubXtP8WdI62Xe5fjE6wQgzCPG8n?usp=sharing
+  (Google Drive link)}
 

### 2021-06-07 papers/final/alex/paper.tex
add gdrive outputs links

@@
 \usepackage{pgfplots}
+\usepackage{url}
 %% NOTE that a single column version may be required for
@@
   generating pieces of multi-instrument symbolic classical music
-  through interpolation.
+  through interpolation. The code utilized for this research is available at:
+  \url{https://github.com/jackPhn/CSS586_Project/}
 \end{abstract}
@@
 
-A collection of several samples of the autoencoded outputs and
-interpolations, is presented at:
-\href{https://drive.google.com/drive/folders/12o04uubXtP8WdI62Xe5fjE6wQgzCPG8n?usp=sharing
-  (Google Drive link)}
+A collection of samples of the autoencoded outputs and
+interpolations, converted to mp3 format, is presented at:
+\url{https://drive.google.com/drive/folders/1SlQOI_cL452PftO3KmOYO_qglE2RotCX?usp=sharing}.
+Each of the 18 directories contains five interpolations between two of
+the string quartet pieces in the MusicNet corpus, numbered 0-4.
 

### 2021-06-07 scripts/string_quartet_inference.py
update gitignore

@@ -45,7 +45,7 @@ num_interpolations = 5  # Number of interpolations per pair
 # Load the best model from training experiments
 
 # %%
-exp_name = "mtvae_0021"
+exp_name = "mtvae"
 exp_time = "2021-06-06T09:32:32"
 saved_model_path = f"../experiments/{exp_name}/{exp_time}/saved_model"
 

### 2021-06-07 scripts/string_quartet_evaluation.py
start evaluation notebook

@@ -1,0 +1,29 @@
+# %%
+import os
+import sys
+from pathlib import Path
+
+import numpy as np
+
+sys.path.append("..")
+from musiclearn import config, processing, vae_models
+
+# %%
+exp_name = "mtvae"
+exp_time = "2021-06-06T09:32:32"
+
+
+# %% [markdown]
+# ## Metrics
+#
+# (from Yang and Lerch 2020)
+#
+# - Pitch count
+# - Pitch class histogram
+# - Pitch class transition matrix
+# - Pitch range
+# - Average pitch interval
+# - Note count
+# - Average inter-onset interval
+# - Note length histogram
+# - Note length transition matrix

### 2021-06-07 papers/final/alex/paper.tex
quant eval figures for paper

@@
 
-A collection of samples of the autoencoded outputs and
-interpolations, converted to mp3 format, is presented at:
+A collection of samples of the autoencoded outputs and interpolations,
+converted to mp3 format, is presented at the following URL:
 \url{https://drive.google.com/drive/folders/1SlQOI_cL452PftO3KmOYO_qglE2RotCX?usp=sharing}.
 Each of the 18 directories contains five interpolations between two of
-the string quartet pieces in the MusicNet corpus, numbered 0-4.
+the string quartet pieces in the MusicNet corpus, numbered 0-4, where
+0 is the reconstruction of the first piece, 4 is the reconstruciton of
+the second piece, and 1-3 are linear interpolations between the two
+pieces. Each file is cropped to the first 16 measures only.
+
+The interpolations are between the following 18 pairs of pieces
+
+\begin{itemize}
+\tightlist
+\item \texttt{Haydn/2104\_op64n5\_1.mid} and \texttt{Ravel/2178\_gr\_rqtf2.mid}
+\item \texttt{Haydn/2105\_op64n5\_2.mid} and   \texttt{Ravel/2179\_gr\_rqtf3.mid}
+\item \texttt{Haydn/2106\_op64n5\_3.mid} and   \texttt{Ravel/2177\_gr\_rqtf1.mid}
+\item \texttt{Beethoven/2497\_qt11\_4.mid} and \texttt{Mozart/1822\_kv\_421\_1.mid}
+\item \texttt{Beethoven/2433\_qt16\_3.mid} and \texttt{Mozart/1859\_kv\_464\_2.mid}
+\item \texttt{Beethoven/2368\_qt12\_4.mid} and \texttt{Mozart/1807\_kv\_387\_3.mid}
+\item \texttt{Beethoven/2314\_qt15\_2.mid} and \texttt{Mozart/1791\_kv\_465\_4.mid}
+\item \texttt{Beethoven/2480\_qt05\_1.mid} and \texttt{Mozart/1792\_kv\_465\_1.mid}
+\item \texttt{Beethoven/2481\_qt05\_2.mid} and \texttt{Mozart/1835\_kv\_590\_3.mid}
+\item \texttt{Beethoven/2379\_qt08\_4.mid} and \texttt{Mozart/1805\_kv\_387\_1.mid}
+\item \texttt{Beethoven/2365\_qt12\_1.mid} and \texttt{Mozart/1793\_kv\_465\_2.mid}
+\item \texttt{Beethoven/2562\_qt02\_4.mid} and \texttt{Mozart/1790\_kv\_465\_3.mid}
+\item \texttt{Beethoven/2494\_qt11\_1.mid} and \texttt{Mozart/1789\_kv\_465\_2.mid}
+\item \texttt{Beethoven/2403\_qt01\_4.mid} and \texttt{Mozart/1788\_kv\_465\_1.mid}
+\item \texttt{Beethoven/2376\_qt08\_1.mid} and \texttt{Dvorak/1916\_dvq10m1.mid}
+\item \texttt{Beethoven/2384\_qt13\_4.mid} and \texttt{Bach/2242\_vs1\_2.mid}
+\item \texttt{Beethoven/2560\_qt02\_2.mid} and \texttt{Beethoven/2621\_qt07\_1.mid}
+\item \texttt{Beethoven/2377\_qt08\_2.mid} and \texttt{Beethoven/2381\_qt13\_1.mid}
+\end{itemize}
 
@@
 
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.6}{\input{figures/source_pitch_count.pgf}}
+    \end{center}
+    \caption{Distinct pitch counts per sample for source samples.}
+    \label{source_pitches}
+\end{figure}
+
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.6}{\input{figures/generated_pitch_count.pgf}}
+    \end{center}
+    \caption{Distinct pitch counts per sample for generated samples.}
+    \label{generated_pitches}
+\end{figure}
+
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.6}{\input{figures/pitch_class_histogram.pgf}}
+    \end{center}
+    \caption{Comparative pitch class histogram for source and generated samples.}
+    \label{pch}
+\end{figure}
+
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.6}{\input{figures/pitch_range_source.pgf}}
+    \end{center}
+    \caption{Pitch range per sample for source samples.}
+    \label{source_ranges}
+\end{figure}
+
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.6}{\input{figures/pitch_range_generated.pgf}}
+    \end{center}
+    \caption{Pitch range per sample for generated samples.}
+    \label{generated_ranges}
+\end{figure}
+
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.6}{\input{figures/avg_pitch_interval_source.pgf}}
+    \end{center}
+    \caption{Average pitch interval per sample for source samples.}
+    \label{source_interval}
+\end{figure}
+
+\begin{figure}[htbp]
+    \begin{center}
+        \scalebox{0.6}{\input{figures/avg_pitch_interval_generated.pgf}}
+    \end{center}
+    \caption{Average pitch interval per sample for generated samples.}
+    \label{generated_interval}
+\end{figure}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\linewidth]{figures/pctm_source.png}
+  \caption{Pitch class transition matrix for source samples.}
+  \label{pctm_source}
+\end{figure}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\linewidth]{figures/pctm_generated.png}
+  \caption{Pitch class transition matrix for generated samples.}
+  \label{pctm_generated}
+\end{figure}
+
 \section{Discussion}

### 2021-06-07 musiclearn/plotting.py

@@ -12,6 +12,8 @@ import seaborn as sns
 
 from musiclearn.processing import REST, SUSTAIN
 
+TRACKS = ["violin 1", "violin 2", "viola", "cello"]
+
 
 def plot_learning_curves(history_file: os.PathLike, dest_path: os.PathLike):
     """Plot the train vs. validation learning curves from a history file."""
@@ -42,8 +44,7 @@ def plot_pitch_class_histogram(x: np.array, dest_path: os.PathLike, by_instrumen
     if dest_path.endswith(".pgf"):
         matplotlib.use("pgf")
     xp = np.where(np.isin(x, [REST, SUSTAIN]), np.nan, x)
-    tracks = ["violin 1", "violin 2", "viola", "cello"]
-    df = pd.DataFrame(xp, columns=tracks)
+    df = pd.DataFrame(xp, columns=TRACKS)
     ax = sns.histplot(data=df)
     ax.set_xlabel("MIDI pitch value")
     os.makedirs(os.path.dirname(dest_path), exist_ok=True)

### 2021-06-07 scripts/string_quartet_evaluation.py

@@ -4,26 +4,196 @@ import sys
 from pathlib import Path
 
 import numpy as np
+import pandas as pd
+import seaborn as sns
+import matplotlib as mpl
+from matplotlib import pyplot as plt
 
 sys.path.append("..")
-from musiclearn import config, processing, vae_models
+from musiclearn import config, processing, vae_models, plotting
 
 # %%
 exp_name = "mtvae"
-exp_time = "2021-06-06T09:32:32"
-
 
 # %% [markdown]
 # ## Metrics
 #
 # (from Yang and Lerch 2020)
 #
+# Pitch-based metrics:
+#
 # - Pitch count
 # - Pitch class histogram
 # - Pitch class transition matrix
 # - Pitch range
 # - Average pitch interval
-# - Note count
-# - Average inter-onset interval
-# - Note length histogram
-# - Note length transition matrix
+#
+# By calculating these metrics for the source music and the generated music
+# we can make quantitative comparisons between them to assess generation quality.
+
+# %% [markdown]
+# First collect the original input data
+
+# %%
+x = processing.get_string_quartets(4)
+
+# %% [markdown]
+#
+# Then collect the generated music (interpolations).
+# If the .npz files don't exist, first run `python string_quartet_inference.py`
+# to create them.
+
+# %%
+outputs_dir = Path("../outputs") / exp_name
+npzs = list(outputs_dir.rglob("*.npz"))
+npz_files = [np.load(npz) for npz in npzs]
+xgen = np.vstack([np.concatenate([np.concatenate([v for k, v in f.items()])]) for f in npz_files])
+
+# %%
+def to_pandas(arr):
+    arr = processing.split_array(arr, 4, 4)
+    df = pd.DataFrame(
+        arr.reshape(arr.shape[0] * arr.shape[1], arr.shape[2]), columns=plotting.TRACKS
+    )
+    df["sample"] = np.divmod(np.arange(len(df)), 16)[0]
+    return df
+
+
+# %%
+x = to_pandas(x)
+xgen = to_pandas(xgen)
+
+# %% [markdown]
+#
+# ## Pitch count
+
+# %%
+def pitch_count(df):
+    df = (
+        pd.melt(df, id_vars=["sample"])
+        .replace(processing.REST, np.nan)
[…]

### 2021-06-07 papers/final/alex/paper.tex
paper edits

@@
 
-Data augmentation was also tested--\cite{oore_this_2018} suggests
-augmentation via pitch shifting each training example up or down by up
-to six semitones in order to create additional training examples and
-reduce overfitting, so we followed that recommendation and used it to
-double the size of the training dataset. As a result, we saw much
-lower validation losses (compare Figure \ref{loss} and Figure
-\ref{loss_augment} and a small but noticeable improvement in the
-fidelity of the song reconstructions, suggesting that this method was
-effective in reducing overfitting.
+Data augmentation was also tested--Oore et al (2018)
+\cite{oore_this_2018} suggests augmentation via pitch shifting each
+training example up or down by up to six semitones in order to create
+additional training examples and reduce overfitting, so we followed
+that recommendation and used it to double the size of the training
+dataset. As a result, we saw much lower validation losses (compare
+Figure \ref{loss} and Figure \ref{loss_augment} and a small but
+noticeable improvement in the fidelity of the song reconstructions,
+suggesting that this method was effective in reducing overfitting.
 
@@
 the string quartet pieces in the MusicNet corpus, numbered 0-4, where
-0 is the reconstruction of the first piece, 4 is the reconstruciton of
+0 is the reconstruction of the first piece, 4 is the reconstruction of
 the second piece, and 1-3 are linear interpolations between the two
@@
 \tightlist
-\item \texttt{Haydn/2104\_op64n5\_1.mid} and \texttt{Ravel/2178\_gr\_rqtf2.mid}
-\item \texttt{Haydn/2105\_op64n5\_2.mid} and   \texttt{Ravel/2179\_gr\_rqtf3.mid}
-\item \texttt{Haydn/2106\_op64n5\_3.mid} and   \texttt{Ravel/2177\_gr\_rqtf1.mid}
-\item \texttt{Beethoven/2497\_qt11\_4.mid} and \texttt{Mozart/1822\_kv\_421\_1.mid}
-\item \texttt{Beethoven/2433\_qt16\_3.mid} and \texttt{Mozart/1859\_kv\_464\_2.mid}
-\item \texttt{Beethoven/2368\_qt12\_4.mid} and \texttt{Mozart/1807\_kv\_387\_3.mid}
-\item \texttt{Beethoven/2314\_qt15\_2.mid} and \texttt{Mozart/1791\_kv\_465\_4.mid}
-\item \texttt{Beethoven/2480\_qt05\_1.mid} and \texttt{Mozart/1792\_kv\_465\_1.mid}
-\item \texttt{Beethoven/2481\_qt05\_2.mid} and \texttt{Mozart/1835\_kv\_590\_3.mid}
-\item \texttt{Beethoven/2379\_qt08\_4.mid} and \texttt{Mozart/1805\_kv\_387\_1.mid}
-\item \texttt{Beethoven/2365\_qt12\_1.mid} and \texttt{Mozart/1793\_kv\_465\_2.mid}
-\item \texttt{Beethoven/2562\_qt02\_4.mid} and \texttt{Mozart/1790\_kv\_465\_3.mid}
-\item \texttt{Beethoven/2494\_qt11\_1.mid} and \texttt{Mozart/1789\_kv\_465\_2.mid}
-\item \texttt{Beethoven/2403\_qt01\_4.mid} and \texttt{Mozart/1788\_kv\_465\_1.mid}
-\item \texttt{Beethoven/2376\_qt08\_1.mid} and \texttt{Dvorak/1916\_dvq10m1.mid}
-\item \texttt{Beethoven/2384\_qt13\_4.mid} and \texttt{Bach/2242\_vs1\_2.mid}
-\item \texttt{Beethoven/2560\_qt02\_2.mid} and \texttt{Beethoven/2621\_qt07\_1.mid}
-\item \texttt{Beethoven/2377\_qt08\_2.mid} and \texttt{Beethoven/2381\_qt13\_1.mid}
+\item \texttt{Haydn/2104\_op64n5\_1.mid} , \texttt{Ravel/2178\_gr\_rqtf2.mid}
+\item \texttt{Haydn/2105\_op64n5\_2.mid} ,   \texttt{Ravel/2179\_gr\_rqtf3.mid}
+\item \texttt{Haydn/2106\_op64n5\_3.mid} ,   \texttt{Ravel/2177\_gr\_rqtf1.mid}
+\item \texttt{Beethoven/2497\_qt11\_4.mid} , \texttt{Mozart/1822\_kv\_421\_1.mid}
+\item \texttt{Beethoven/2433\_qt16\_3.mid} , \texttt{Mozart/1859\_kv\_464\_2.mid}
+\item \texttt{Beethoven/2368\_qt12\_4.mid} , \texttt{Mozart/1807\_kv\_387\_3.mid}
+\item \texttt{Beethoven/2314\_qt15\_2.mid} , \texttt{Mozart/1791\_kv\_465\_4.mid}
+\item \texttt{Beethoven/2480\_qt05\_1.mid} , \texttt{Mozart/1792\_kv\_465\_1.mid}
+\item \texttt{Beethoven/2481\_qt05\_2.mid} , \texttt{Mozart/1835\_kv\_590\_3.mid}
+\item \texttt{Beethoven/2379\_qt08\_4.mid} , \texttt{Mozart/1805\_kv\_387\_1.mid}
+\item \texttt{Beethoven/2365\_qt12\_1.mid} , \texttt{Mozart/1793\_kv\_465\_2.mid}
+\item \texttt{Beethoven/2562\_qt02\_4.mid} , \texttt{Mozart/1790\_kv\_465\_3.mid}
+\item \texttt{Beethoven/2494\_qt11\_1.mid} , \texttt{Mozart/1789\_kv\_465\_2.mid}
+\item \texttt{Beethoven/2403\_qt01\_4.mid} , \texttt{Mozart/1788\_kv\_465\_1.mid}
+\item \texttt{Beethoven/2376\_qt08\_1.mid} , \texttt{Dvorak/1916\_dvq10m1.mid}
+\item \texttt{Beethoven/2384\_qt13\_4.mid} , \texttt{Bach/2242\_vs1\_2.mid}
+\item \texttt{Beethoven/2560\_qt02\_2.mid} , \texttt{Beethoven/2621\_qt07\_1.mid}
+\item \texttt{Beethoven/2377\_qt08\_2.mid} , \texttt{Beethoven/2381\_qt13\_1.mid}
 \end{itemize}

### 2021-06-07 papers/final/jack/paper.tex

--- 2021-05-17 papers/final/paper.tex
@@
 \documentclass[sigconf,authorversion]{acmart}
+\usepackage{pgf}
+\usepackage{pgfplots}
+\usepackage{url}
 %% NOTE that a single column version may be required for
@@
 \pagestyle{plain}
-\title{Musical Deep Learning}
+\title{Automatic Music Generation with Deep Learning}
 
@@
 
-\author{Alex Kyllo}
-\email{akyllo@uw.edu}
-\affiliation{%
-  \institution{University of Washington}
-  \city{Bothell}
-  \state{WA}
-  \country{USA}
-}
-
 \begin{abstract}
-  This paper presents a generative model of multi-instrument symbolic
-  music.
+Automatic music generation is the art of generating new music with little human 
+intervention. This art has a long history, and recently breakthrough in deep learning
+provides brand-new tools for artists and untrained curious laypersons to generate
+their own music with any level of music understanding. A music recording could be 
+presented in different data types: raw audio waveforms, frequency-domain spectrogram, 
+and symbolic representation of piano rolls. Any of these data types could be used 
+to train deep learning models. This paper presents the results of experiments with 
+recurrent neural networks (LSTM, bidirectional LSTM), and 1-D convolutional neural 
+networks for automatic music generation. Specifically, we investigate single-track 
+classical piano music generation at the musical note level. The code for preprocessing 
+data and building deep learning models are available at 
+\url{https://github.com/jackPhn/CSS586_Project/}.
 \end{abstract}
 
-\keywords{deep learning, neural networks, music}
+\keywords{deep learning, neural networks, recurrent models, music}
 
@@
 
-Machine learning models of music have interesting applications in
-music information retrieval and creative tools for musical artists and
-educators. Generative models can create accompaniments for music,
-transfer styles between clips of music, and even generate entirely new
-music. Music is challenging to model because it exhibits a complex
-hierarchy of recurring patterns and long-range temporal dependencies,
-and because both musical scores and performances have multiple
-possible digital representations.
+The poet Henry Wadsworth Longfellow once said that music is the universal 
+language of mankind. Truely, we use music to convey emotion and feeling without
+words. Music is an integral part of human culture, which
+came into existence since the dawn of civilization and has since continued to 
+proliferate into various forms. Composing a piece of music is a complicated process 
+requiring a deep understanding of music theory and creativity. To 
+generate new music, typically a composition is created in the form of music notation 
+like sheet music prior to transforming it to sound by the hand of one or more musicians.
+Because music generation is a specialized process that requires a lot of training,
+generating music is still preserved for a trained musicians. To promote the discovery of 
+new music, everyone should be able to generate their own music without spending
+years of training.
 
-Depending on the task, machine learning models of music may be trained
-on the audio signal itself, either in a time domain or a frequency
-domain representation, or they may be trained on a digital symbolic
-representation of music, the most common of which is MIDI (Musical
-Instrument Digital Interface) notation. MIDI is an encoding of music
-as streams of bytes in one or more tracks or channels, each
-representing a sequence of 128 possible pitch values (where 0 is the
-lowest and 127 is the highest), along with timing, pressure and
-instrument identifier values. A music transcription model may
-transcribe an audio signal as a MIDI score, which can easily be
-converted into other symbolic representations such as sheet music for
-human performers to read from, while a synthesizer model can convert
-MIDI representations into audio signals. A generative music model can
-be trained either to generate raw audio, or to produce a symbolic
-score that must be played by a synthesizer or by humans to produce an
-audio music performance. This project focuses on the latter type of
-modeling: generative modeling of symbolic (MIDI) music to compose
-original musical scores.
+There are multiple representations of music to consider. Music data are available
+as raw time-domain audio waveform, frequency-domain spectrogram, or symbolic
+representation called piano rolls. A piece of music consists of one or more musical
+tracks generated by one or more musicians performing on different musical
+instruments.
+
+Deep learning has been successfully used in applications ranging from speech
+recognition to image classification and segmentation, among many others tasks.
+The generative capacity of deep learning is recently discovered and has been
+explored extensively to generate novel images, videos, and texts.
+Recurrent Neural Network (RNN) has shown its capability to
+model time-series sequences. Advanced version of RNN such as Gated Recurrent 
+Unit (GRU) and Long Short-Term Memory (LSTM) have been successfully used for 
+applications such as voice recognition, text to speech, and other applications 
+that involve time-series data. RNN could also be used to generate novel 
+sequences of texts or sounds. In addition to RNN, 1D convolutional neural network
+could also model the dependencies between elements in a sequence.
+
+In this paper, we investigate the use of sequential recurrent models and 1-D 
+convolutional neural network to model the dependency among the notes and chords
+of musical sequences from classical music performed on piano. The trained models 
+are used to generate single-track piano performances.
 
@@
 
-The state of the art in music generation has a long way to go before
-it can consistently generate music scores or performances that would
-be enjoyable and popular for humans to listen to, but a number of
-recent research projects have shown promising progress in this area.
+Automatic music generation dates back to 1781 when Mozarts invented 
+a musical dice game in which a player generate new music by selecting
+a tone from a list of pre-composed tone based on the sum of two dices 
+\cite{enwiki:995309040}. In the past decade, deep learning has become
+a increasingly popular tool for artists and researchers who are interested in automatic 
+music generation. This section introduces several state-of-the-art deep learning 
+works in the field of automatic music generation.
 
-Google's \href{https://magenta.tensorflow.org/}{Magenta} is an
-umbrella project for music deep learning research and development of
-software tools to expose these models for use by creative artists and
-students.
+In the physical domain, a sound is essentially the displacement of air molecules,
+which results in fluctuation in air pressure. When sound or music is recorded,
+tens of thousands of air pressure data points are captured per second to
+construct the raw audio waveform. Since a piece of music spans anywhere between 
+a few seconds to several minutes, its raw audio waveform contains a large amount
+of data, making the modeling of long-distance dependencies difficult. Google DeepMind
+based in London, UK, introduced WaveNet, a generative deep neural network, in 2016
+to tackle this problem \cite{oord_wavenet_2016}. WaveNet is a fully probabilistic
+and autoprogressive model, in which the distribution of each audio sample depends on 
+the previous ones. It was used to generate spoken English speech and music.
 
-MusicVAE, part of the Magenta project, is a variational Long
-Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
-novel hierarchical structure using a ``conductor'' recurrent layer in
-its decoder model to better capture structure at multiple levels and
-avoid the problem of ``posterior/mode collapse'' whereby a generative
-model learns to ignore its latent code and rely on autoregression
-\cite{roberts_hierarchical_2018}. This model is trained on 16-bar
-paragraphs of music and is capable of generating new melodies that
-blend two given melodies via latent space interpolation.
+Another generative model that operates on the time-domain raw audio waveforms is 
+SampleRNN \cite{mehri_samplernn_2017}. This model consists of memory-less autoregressive 
+multilayer perceptron modules and stateful recurrent neural networks in a hierarchical 
+structure. This combination allows the model to capture the structure of temporal 
+sequences over long time spans.
 
-Another Magenta model called Music Transformer is a generative model
-that borrows its approach from the Natural Language Processing (NLP)
-domain, using a self-attention network to model MIDI music as a
-sequence of discrete tokens with relative positional dependencies
-\cite{huang_music_2018}. The focus of this model is on learning
-long-term dependencies in music to produce longer clips of music with
-coherent structure. Music Transformer was trained on a dataset of
-Piano-e-competition performances \cite{hawthorne2019enabling} and its
-generated piano music received favorable qualitative (Likert scale)
-ratings from human listeners for its resemblance to human-composed
-music \cite{huang_music_2018}.
+Modeling raw audio waveforms is challenging since a single second in
+modern recording spans thousands of timesteps. Most commonly, music is
+recorded at 44.1 kHz (44,100 samples per second). Since a piece of music
+is at least a few seconds in length, capturing long-range dependencies
+is difficult in the time domain. A model named MelNet was introduced to address
+this limitation of time-domain models \cite{vasquez_melnet_2019-1}. MelNet 
+leverages a two-dimentsional time-frequency representation of audio called 
+spectrogram, which reduces the dimensionality of the audio waveforms. This 
+model was able to generate high-fidelity audio samples using structures on
+long timescales.
 
-MuseGAN \cite{dong2017musegan} is an application of Generative
-Adversarial Networks (GAN) to polyphonic MIDI music generation,
-trained on four-bar phrases of a multi-track pianoroll representation
-of rock songs from the Lakh Midi Dataset
-\cite{raffel_learning-based_2016}. Like MusicVAE, MuseGAN includes a
-two-level generator that first samples latent codes at the phrase or
-bar level, then generates notes within the bars, to produce
-longer-term structural patterns.
+The discussion on generative models is not complete without mention of Generative
+Adversarial Networks (GAN). Google AI introduced GANSynth in 2019 following the 
+publication of WaveNet \cite{engel_gansynth_2019}. GANSynth generates log-magnitude
+spectrograms and phases prior to generating the audio waveform. It was shown to 
+outperform WaveNet in automatic and human evaluations and could generate music much
+faster than WaveNet.
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{song_structure.png}
-  \caption{Diagram of the hierarchical structure of a musical
-    composition, from the MuseGAN paper \cite{dong2017musegan}.}
-  \label{song_structure}
-  \Description{A diagram showing the hierarchical structure of songs,
-    to paragraphs, to phrases, to bars, to, beats, to pixels.}
-\end{figure}
-
-A major advantage of working with the symbolic representation of music
-is that it is of far lower dimensionality than the raw audio waveforms
-of a recorded performance, which makes it less computationally
-expensive. However, there are many stylistic aspects of musical
-performance that are not captured by a symbolic representation, and
-may be specific to a particular performer, so the expressiveness of
-symbolic generative models is limited in comparison
-\cite{manzelli_conditioning_2018}.
-
-Other research has focused on modeling raw audio waveforms directly. WaveNet is
-a causal convolutional neural network for generating raw audio waveforms,
-developed by Google DeepMind, which achieves state of the art performance in
-generating natural sounding speech from text, but is also capable of generating
-short, realistic snippets of audio music \cite{oord_wavenet_2016}.
-Another model named SampleRNN generates raw audio waveforms using a three-tier
-hierarchy of gated recurrent units (GRU) to model recurrent structure at
-multiple temporal resolutions \cite{mehri_samplernn_2017}.
-
-Prior work points out that the division between symbolic music notes
-and music performances, is analogous to the division between symbolic
-language and utterances in speech, which may inspire ideas for
-combining the two approaches \cite{hawthorne2019enabling}. A paper
-from Boston University describes an effort to combine the symbolic and
-waveform approaches to music modeling, by training an LSTM to learn
-melodic structure of different styles of music, then providing
-generations from this model as conditioning inputs to a WaveNet-based
-raw audio generator \cite{manzelli_conditioning_2018}.
+Music creation typically starts with the composition of a musical score prior 
+to the performance by a musician. The majority of works in automatic music 
+generation focus on training a model using performance data like audio sound or
+ its frequency transformation. Two researchers from Taiwan took another
+approach and published PerformanceNet, a score-to-audio music generation model
+\cite{wang_performancenet_2018}. This model consists of two subnets: a U-Net-like 
+convolutional neural network that maps musical scores in the form of piano rolls 
+to spectrograms; and a residual network that refines the outputs.
 
@@
 
-Two primary datasets will be used for this research project: MusicNet,
-which is a collection of 330 freely licensed European classical music
-recordings with aligned MIDI scores \cite{thickstun2017learning}, and
-the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
-Commons licensed MIDI files of popular music songs that have been
-matched and aligned to MP3 recordings, and of which we plan to use a
-subset for model training and evaluation
-\cite{raffel_learning-based_2016}. Including this second dataset may
-help to generalize the modeling approach beyond European classical
-music to include other popular genres and associated instruments.
+To train our models and to provide seeds for music generation, we used 
+the MusicNet, an open-source dataset published by the department of Computer 
+Science at the University of Washington \cite{thickstun_learning_2017}. This 
+dataset contains a collection of 330 freely-licensed classical music recordings 
+captured under various studio and microphone conditions. The recordings are
+available in .wav raw audio format and MIDI (Musical Instrument Digital
+Interface). A MIDI file contains the encoding of music, which is not playable
+like an audio file but is viewable using specialized software such as 
+Apple GarageBand. It contains the performance information such as musical
+instruments and the temporal location of notes and chords in a composition.
+
+MusicNet consists of music composed by Beethoven, Bach, Ravel, Faure, and Schubert,
+to name a few. Since the focus of this project is on piano music generation, only 
+Schubert's compositions are selected.
 
@@
 
-Several choices must be made in how to preprocess binary MIDI files
-into training examples for a neural network. There are multiple
-open-source Python packages that assist with the process of reading
-MIDI files from their binary on-disk representations into Python
-objects, such as pretty\_midi \cite{raffel_pretty_midi_2014},
-Pypianoroll \cite{dong_pypianoroll_2018} and music21
-\cite{cuthbert_music21_2010}.
+MIDI files are in a specialized format that requires extensive processing as 
+the data are encoded according to the MIDI standard. Several open-source Python
+packages exist to assist the works on MIDI files. A few examples are Pypianoroll 
+\cite{dong_pypianoroll_2018} and music21 \cite{cuthbert_music21_2010}. We
+decided to use music21 to process the MIDI files from Schubert's collection.
 
-In order to accommodate polyphonic music, we convert each MIDI file
-into a pianoroll representation, as visualized in Figure
-\ref{pianoroll}, wherein each instrument track is a sparse matrix that
-multi-hot encodes the velocity values for each of 128 possible pitch
-levels at each timestep. To reduce dimensionality, I will clip the
-note pitch values to a narrower range, excluding the rarely used notes
-in the extreme high and low octaves, and encode chords (combinations
-of notes played simultaneously on the same instrument) as distinct
-tokens, so that they can be one-hot encoded
+Although the main instrument used to perform the compositions in Schubert's
+collection is the piano, several recordings contain musical tracks generated by 
+string instruments and other instruments. Since our focus is on piano music,
+these tracks need to be filtered out. To do this, the MIDI files are partitioned 
+by the instruments. After partitioning, each file contains an array of different
+instrument tracks. Any track other than piano is removed.
 
-Because songs are typically at least a few minutes long and of varying
-length, it will not be feasible to train with entire songs as
-examples, so we will crop songs into phrases of equal numbers of
-measures to use as training data.
+A music21 instrument track is a Python object containing a list of notes and chords
+(multiple notes played at the same time) that are played at different time points in 
+a composition. It also contains rest (silence) and sustain (holding the
+previous note continuously). For simplicity, the rest and sustain are discarded
+from the dataset. The MIDI file format also specifies the amplitude values of 
+the air pressure, which are presented in the vertical axis of a plot of the raw 
+audio waveform. This information denotes the volume of each note in a performance.
+For this project, the information about note volume would unnecessarily increase the 
+dimensionality of the data. We decided to disregard the volume and to use the same 
+pressure value for all notes and chords.
 
-The result of this preprocessing is that each training example will be
-a 3D tensor of shape (tracks x ticks x pitches) and stacking the
-training examples will produce a 4D tensor.
+From the isolated piano tracks, we built a corpus of extracted notes and chords. 
+There is a maximum of 128 musical notes in any given piece of music. Chords, on the 
+other hand, are combination of two or more notes; thus the number of possible permutations 
+is very large. In Schubert's collection, the number of distinct notes and chords is 312. 
+We decided to use all 312 notes and chord for single-track
+piano music generation. We mapped unique integers to the notes and chords since deep 
+learning models do not work well with non-numeric data format.
 
-While we plan to model music with multiple instrument tracks, we
-anticipate the need to model only a fixed selection of instrument
-parts, similar to how MusicVAE models three-part (drum, bass and
-melody) \cite{roberts_hierarchical_2018} and MuseGAN models five-part
-(drum, bass, guitar, string, piano) arrangements.
+Each song in Schubert's collection spans a few minutes. However, our goal for this 
+project is to model short sequences of music. To this end, we partitioned the corpus
+into sequences of 100 notes/chords. The sequences are created by sliding a 100-note-long
+window on the corpus, moving it one element at a time. The label for each sequence is 
+the immediate element in the corpus that is next to its last element. 
+Figure \ref{preprocessing} provides a summary for the data preprocessing pipeline.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{first_bar.png}
-  \caption{Pianoroll visualization of the first measure of
-    Beethoven's Serioso String Quartet}
-  \label{pianoroll}
-  \Description{A pianoroll representation of cello, viola and violin
-    parts from a Beethoven piece.}
+  \includegraphics[width=0.5\linewidth]{preprocess.png}
+  \caption{The data preprocessing pipeline consists of five steps.}
+  \label{preprocessing}
+  \Description{Preprocessing pipeline.}
 \end{figure}
 
-Data augmentation is also possible and we assess its impact on
-the results--the literature suggests augmentation via pitch shifting
-each training example up or down by up to six semitones, and
-increasing or reducing the speed by up to 10\% in order to create
-additional training examples and reduce overfitting
-\cite{oore_this_2018}.
-
-The multi-stream representation discussed in
-\cite{kumar2019polyphonic} addresses the issue of extreme class
-imbalance and sparsity in the pianoroll representation.
+To do: figure for encoding of notes and chords to numbers
 
@@
 
+In this project, we treat a short piece of music like 
+a sentence, in which the vocabulary is the collection of distinct notes 
+and chords, and the grammar is the local and global dependencies between
+the notes and chords. We explore the effectiveness of four types of models 
+in automatic music generation: Long Short Term Memory (LSTM), Bidirectional
+LSTM, Bidirectional LSTM with attention, and a 1D convolutional neural 
+network inspired by WaveNet's architecture. These models are many-to-one
+models since the input is a sequence of multiple musical values and the output 
+is a single note or chord. The models are built using the TensorFlow library.
+
+\subsubsection{LSTM Model}
+
+The first model is a recurrent neural network that contains LSTM layer as 
+the basic building block. The LSTM model is selected because it is able to encode
+the dependencies between the elements of a long sequence via a gating
+mechanism. The main components of the model are LSTM layers and fully connected
+layers. The output layer has 312 perceptrons corresponding to the number
+of distinct musical notes and chords in the dataset. An output perceptron 
+calculates the probabilities that its corresponding note or chord is the next
+element in the sequence. The output activation function is softmax and the
+model is trained by minimizing the sparse categorical cross-entropy loss.
+The architecture of this model is shown in Figure \ref{lstm_model}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.7\linewidth]{lstm_model.png}
+  \caption{Architecture of the LSTM model.}
+  \label{lstm_model}
+  \Description{LSTM model}
+\end{figure}
+
+\subsubsection{Bidirectional LSTM Model}
+
+LSTM models capture the information in a sequence only in the forward direction,
+from the beginning to end. An enhanced version of LSTM models is the bidirectional 
+LSTM model. A bidirectional LSTM model trains two LSTMs simultaneously, one on the 
+input sequence and another on the inverted input sequence. Effectively, the model 
+encodes the structure of the past as well as the future. The bidirectional LSTM
+model shares the same architecture as LSTM model, with the only difference being
+the replacement of the LSTM layer by the bidirectional LSTM layer. Figure 
+\ref{bidirect_lstm_model} depicts the structure of this model.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.7\linewidth]{bidirect_lstm_model.png}
+  \caption{Architecture of the bidirectional LSTM model.}
+  \label{bidirect_lstm_model}
+  \Description{Bidirectional LSTM model}
+\end{figure}
+
+\subsubsection{Bidirectional LSTM Model with Attention}
+
+The bidirectional LSTM model introduces an additional complexity by using the 
+reverse of the input sequence. It is still not complete because of a fundamental
+problem: its hidden states consider the elements of the input sequence with 
+equal importance. To solve this problem, we added an attention layer following 
+a bidirectional LSTM layer. The attention mechanism is shown in Figure 
+\ref{attention_mechanism}. The importance of the hidden states of a bidirectional 
+LSTM layer is captured by a set of attention weights. These weights enhance or 
+supress a hidden state according to its contribution to the output of an input
+sequence. Figure \ref{att_lstm_model} demonstrates how the attention mechanism
+fits into the structure of the bidirectional LSTM model.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{attention_mec.png}
+  \caption{The attention mechanism is encapsulated in an attention block that has
+  a set of attention weights (W) and attention biases (b). The attention biases are
+  added to the dot product result of the previous LSTM layer's hidden state and 
+  attention weights. The results are then fed into a tanh function and a softmax
+  function to generate attention outputs (A).}
+  \label{attention_mechanism}
+  \Description{Attention mechanism.}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.5\linewidth]{attention_model.png}
+  \caption{Architecture of the bidirectional LSTM model with attention mechanism.
+  The Customized Attention layer is derived from Keras Layer object.}
+  \label{att_lstm_model}
+  \Description{Bidirectional LSTM model with attention mechanism.}
+\end{figure}
+
+\subsubsection{WaveNet-Style 1-D Convolutional Neural Network}
+
+The original WaveNet model is designed for the raw audio waveforms, which contain
+tens of thousands of discrete points per second of sound. WaveNet consists of 
+ multiple 1-D dilated causal convolutional layers, which is visualized in Figure 
+ \ref{causal_conv}, enabling the network to have a large receptive field. Consequently,
+ the network can learn the long-term dependency in extremely long sequences. We are
+ interested in this property of WaveNet and constructed our 1-D dilated causal 
+ convolution network as seen in Figure \ref{wavenet_model}. The core of this network
+ is three 1-D dilated causal convolutional layers that have dilation rates in ascending order 
+ (1, 2, and 4). The output layer is similar to that of the LSTM models.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{causal_conv.png}
+  \caption{The core of WaveNet is dilated causal convolutuon. (A) Causal convolution 
+  ensures that the output at time step t is the result of only previous time steps 
+  with no input from the future. (B) With dilated convolution, the convolutional
+  kernel has a larger receptive field, increasing the number of input elements
+  that an output captures. Figures are adopted from the WaveNet paper \cite{oord_wavenet_2016}.}
+  \label{causal_conv}
+  \Description{Causal dilated convolution.}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.7\linewidth]{wavenet_model.png}
+  \caption{Architecture of the WaveNet-like 1-D dilated causal convolutional neural network.}
+  \label{wavenet_model}
+  \Description{Architecture of the 1D CNN.}
+\end{figure}
+
+\subsection{Generating New Music}
+
+The same method of generating a new sequence of notes and chords is used on all trained
+modules. This section describes this method, which is illustrated in Figure \ref{generate_music}. 
+An input sequence containing 100 notes/chords is randomly taken from the corpus and
+fed to a trained model as the initial condition. The model predicts the most probable note
+or chord that comes after this input sequence. The output is appended to an output array, which is
+empty at the beginning. The output note is also appended back to the input sequence. To maintain 
+the input length, the first note of the input sequence is cropped out. The new input sequence
+is fed into the model to gerate a new note/chord. This process is repeated until the output
+array reaches a desirable length.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{generate_music.png}
+  \caption{The process of generating a new sequence of musical notes and chords 
+  using an initial condition. The solid lines represent the forward path to the output
+  sequence. The dashed line indicates the backward path via which the output note/chord at
+  a step is appends to the end of the input sequence.}
+  \label{generate_music}
+  \Description{Generating music with initial condition.}
+\end{figure}
+
 \subsection{Model Evaluation}
 
-Evaluation of generative models is challenging because there is no
-equivalent of an accuracy metric like what is used in supervised
-learning. For autoencoder models we can measure how accurately the
-model can reconstruct its own inputs, but this does not tell us the
-quality of the interpolated examples. Generative models are typically
-evaluated using a combination of qualitative metrics whereby human
-judges rate the quality of the generated examples (essentially a
-Turing test), and quantitative metrics that assess the differences in
-the parametric distributions of generated and real examples. Yang and
-Lerch (2020) proposes a set of metrics informed by music theory, for
-probabilistically evaluating how similar the generations are to known
-sample distributions of real music \cite{yang_evaluation_2020}. These
-metrics include counts, ranges, histograms and transition matrices of
-pitches and note lengths, then the Kullback-Leibler divergence and
-overlapping area of the probability density functions are used to
-compare against known reference distributions per musical genre
-\cite{yang_evaluation_2020}. Due to the cost and time requirements
-associated with designing a human subjects experiment, we
-utilize this quantitative approach to the quality assessment of
-generated samples.
+Similar to the evaluation of GANs, evaluating music generating models is a challenging 
+task since there is a lack of metrics that are used for supervised machine learning models.
 
-\section{Results}
+\section{Experiments and Results}
 
-\section{Discussion}
+For LSTM-based models, hyperparameter tuning was performed on the basic LSTM model 
+that does not have the bidirectional connection or the attention mechanism. 
+The intention is that the hyperparameters that result in the best performance 
+are used for the bidirectional LSTM models with and without the attention mechanism
+since these more advanced models have a similar structure with the basic LSTM
+model. We performed hyperparameter tuning on the following values, with the 
+best parameters in \textbf{bold}:
+
+\begin{itemize}
+  \tightlist
+  \item batch size: 32, \textbf{64}, 128
+  \item learning rate: 0.01, \textbf{0.001}, 0.0001
+  \item dimension of output space in the first LSTM layer: 64, 128, 256, \textbf{512} 
+  \item dropout rate: 0.2, \textbf{0.3}, 0.4, 0.5
+\end{itemize}
+
+For the 1-D dilated causal neural network, the following hyperparameter were tested:
+
+\begin{itemize}
+  \tightlist
+  \item batch size: 32, \textbf{64}, 128
+  \item learning rate: 0.01, \textbf{0.001}, 0.0001
+  \item number of filters in the first Conv1D layer: 32, \textbf{64}, 128, 256
+  \item number of filters in the first Conv1D layer: 32, 64, \textbf{128}, 256
+  \item number of filters in the first Conv1D layer: 32, 64, 128, \textbf{256}
+  \item number of filters in the first Conv1D layer: 32, 64, 128, \textbf{256}
+  \item number of filters in the first Conv1D layer: 32, 64, 128, \textbf{256}
+  \item dropout rate: \textbf{0.2}, 0.3, 0.4, 0.5
+\end{itemize}
+
+The models were trained for 100 epochs.
+
+\section{Sample Output}
+
+A collection of music generated by the LSTM models and 1-D dilated
+causal convolutional neural network is available at 
+\url{https://drive.google.com/drive/folders/1SlQOI_cL452PftO3KmOYO_qglE2RotCX?usp=sharing}.
+
+\section{Conclusion and Discussion}
 

### 2021-06-07 run_experiments.sh
gitignore shell script

@@ -1,47 +1,0 @@
-#python musiclearn_cli.py fit-mtvae mtvae_0001 --beats-per-phrase 4 --lstm-units 128
-#python musiclearn_cli.py fit-mtvae mtvae_0001 --beats-per-phrase 4 --dropout-rate 0.5
-#python musiclearn_cli.py fit-mtvae mtvae_0001 --beats-per-phrase 8
-#python musiclearn_cli.py fit-mtvae mtvae_0001 --beats-per-phrase 8 --lstm-units=512
-#python musiclearn_cli.py fit-mtvae mtvae_0001 --beats-per-phrase 8 --lstm-units=512 --dropout-rate 0.2
-#python musiclearn_cli.py fit-mtvae mtvae_0002 --beats-per-phrase 16 --lstm-units=256 --embedding-dim=16 --dropout-rate 0.4 --patience 50
-#python musiclearn_cli.py fit-mtvae mtvae_0003 --beats-per-phrase 8 --gru --patience 50
-#python musiclearn_cli.py fit-mtvae mtvae_0003 --beats-per-phrase 8 --gru --embedding-dim=32 --patience 50
-#python musiclearn_cli.py fit-mtvae mtvae_0003 --beats-per-phrase 16 --gru --embedding-dim=32 --patience 50
-#python musiclearn_cli.py fit-mtvae mtvae_0003 --beats-per-phrase 16 --gru --embedding-dim=48 --patience 50
-#python musiclearn_cli.py fit-mtvae mtvae_0003 --beats-per-phrase 16 --lstm --embedding-dim=48 --patience 50
-#python musiclearn_cli.py fit-mtvae mtvae_0004 --beats-per-phrase 16 --lstm-units=256 --embedding-dim=16 --dropout-rate 0.4 --patience 50
-#python musiclearn_cli.py fit-mtvae mtvae_0005 --beats-per-phrase 4 --lstm-units 512 --embedding-dim 0 --patience 50
-#python musiclearn_cli.py fit-mtvae mtvae_0006 --beats-per-phrase 4 --lstm-units 256 --latent-dim 1024 --embedding-dim 128 --patience 100
-#python musiclearn_cli.py fit-mtvae mtvae_0007 --beats-per-phrase 4 --lstm-units 512 --latent-dim 1024 --embedding-dim 32 --patience 200
[…]

### 2021-06-08 papers/final/alex/paper.tex
metric analysis in paper

@@
 Data augmentation was also tested--Oore et al (2018)
+
 \cite{oore_this_2018} suggests augmentation via pitch shifting each
@@
 the original inputs, preserving the key signature and general rhythmic
-and melodic structure, albeit with noticeable differences at the
-individual note level.
+and melodic structure, albeit with small but noticeable differences at
+the individual note level, such as more silence and less variation in
+pitch.
 
@@
 
-The interpolations are between the following 18 pairs of pieces
+The interpolations are between the following 18 pairs of pieces:
 
@@
 \tightlist
-\item \texttt{Haydn/2104\_op64n5\_1.mid} , \texttt{Ravel/2178\_gr\_rqtf2.mid}
-\item \texttt{Haydn/2105\_op64n5\_2.mid} ,   \texttt{Ravel/2179\_gr\_rqtf3.mid}
-\item \texttt{Haydn/2106\_op64n5\_3.mid} ,   \texttt{Ravel/2177\_gr\_rqtf1.mid}
-\item \texttt{Beethoven/2497\_qt11\_4.mid} , \texttt{Mozart/1822\_kv\_421\_1.mid}
-\item \texttt{Beethoven/2433\_qt16\_3.mid} , \texttt{Mozart/1859\_kv\_464\_2.mid}
-\item \texttt{Beethoven/2368\_qt12\_4.mid} , \texttt{Mozart/1807\_kv\_387\_3.mid}
-\item \texttt{Beethoven/2314\_qt15\_2.mid} , \texttt{Mozart/1791\_kv\_465\_4.mid}
-\item \texttt{Beethoven/2480\_qt05\_1.mid} , \texttt{Mozart/1792\_kv\_465\_1.mid}
-\item \texttt{Beethoven/2481\_qt05\_2.mid} , \texttt{Mozart/1835\_kv\_590\_3.mid}
-\item \texttt{Beethoven/2379\_qt08\_4.mid} , \texttt{Mozart/1805\_kv\_387\_1.mid}
-\item \texttt{Beethoven/2365\_qt12\_1.mid} , \texttt{Mozart/1793\_kv\_465\_2.mid}
-\item \texttt{Beethoven/2562\_qt02\_4.mid} , \texttt{Mozart/1790\_kv\_465\_3.mid}
-\item \texttt{Beethoven/2494\_qt11\_1.mid} , \texttt{Mozart/1789\_kv\_465\_2.mid}
-\item \texttt{Beethoven/2403\_qt01\_4.mid} , \texttt{Mozart/1788\_kv\_465\_1.mid}
-\item \texttt{Beethoven/2376\_qt08\_1.mid} , \texttt{Dvorak/1916\_dvq10m1.mid}
-\item \texttt{Beethoven/2384\_qt13\_4.mid} , \texttt{Bach/2242\_vs1\_2.mid}
-\item \texttt{Beethoven/2560\_qt02\_2.mid} , \texttt{Beethoven/2621\_qt07\_1.mid}
-\item \texttt{Beethoven/2377\_qt08\_2.mid} , \texttt{Beethoven/2381\_qt13\_1.mid}
+\item \texttt{Haydn/2104\_op64n5\_1.mid} \& \texttt{Ravel/2178\_gr\_rqtf2.mid}
+\item \texttt{Haydn/2105\_op64n5\_2.mid} \&   \texttt{Ravel/2179\_gr\_rqtf3.mid}
+\item \texttt{Haydn/2106\_op64n5\_3.mid} \&   \texttt{Ravel/2177\_gr\_rqtf1.mid}
+\item \texttt{Beethoven/2497\_qt11\_4.mid} \& \texttt{Mozart/1822\_kv\_421\_1.mid}
+\item \texttt{Beethoven/2433\_qt16\_3.mid} \& \texttt{Mozart/1859\_kv\_464\_2.mid}
+\item \texttt{Beethoven/2368\_qt12\_4.mid} \& \texttt{Mozart/1807\_kv\_387\_3.mid}
+\item \texttt{Beethoven/2314\_qt15\_2.mid} \& \texttt{Mozart/1791\_kv\_465\_4.mid}
+\item \texttt{Beethoven/2480\_qt05\_1.mid} \& \texttt{Mozart/1792\_kv\_465\_1.mid}
+\item \texttt{Beethoven/2481\_qt05\_2.mid} \& \texttt{Mozart/1835\_kv\_590\_3.mid}
+\item \texttt{Beethoven/2379\_qt08\_4.mid} \& \texttt{Mozart/1805\_kv\_387\_1.mid}
+\item \texttt{Beethoven/2365\_qt12\_1.mid} \& \texttt{Mozart/1793\_kv\_465\_2.mid}
+\item \texttt{Beethoven/2562\_qt02\_4.mid} \& \texttt{Mozart/1790\_kv\_465\_3.mid}
+\item \texttt{Beethoven/2494\_qt11\_1.mid} \& \texttt{Mozart/1789\_kv\_465\_2.mid}
+\item \texttt{Beethoven/2403\_qt01\_4.mid} \& \texttt{Mozart/1788\_kv\_465\_1.mid}
+\item \texttt{Beethoven/2376\_qt08\_1.mid} \& \texttt{Dvorak/1916\_dvq10m1.mid}
+\item \texttt{Beethoven/2384\_qt13\_4.mid} \& \texttt{Bach/2242\_vs1\_2.mid}
+\item \texttt{Beethoven/2560\_qt02\_2.mid} \& \texttt{Beethoven/2621\_qt07\_1.mid}
+\item \texttt{Beethoven/2377\_qt08\_2.mid} \& \texttt{Beethoven/2381\_qt13\_1.mid}
 \end{itemize}
 
-\subsection{Model Evaluation}
+\subsection{Generated Sample Evaluation}
 
 Evaluation of generative models is challenging because there is no
-equivalent of an accuracy metric like what is used in supervised
-learning. For autoencoder models we can measure how accurately the
-model can reconstruct its own inputs, but this does not tell us the
-quality of the interpolated examples. Generative models are typically
-evaluated using a combination of qualitative metrics whereby human
-judges rate the quality of the generated examples (essentially a
-Turing test), and quantitative metrics that assess the differences in
-the parametric distributions of generated and real examples. Yang and
-Lerch (2020) proposes a set of metrics informed by music theory, for
+direct equivalent of an accuracy or error metric like what is used in
+supervised learning. For autoencoder models we can measure how
+accurately the model can reconstruct its own inputs, but this does not
+tell us the quality of the interpolated examples. Generative models
+are typically evaluated using a combination of qualitative metrics
+whereby human judges rate the quality of the generated examples and
+guess whether they were generated by a machine (essentially a Turing
+test), and quantitative metrics that assess the differences in the
+parametric distributions of generated and real examples. Yang and
+Lerch (2020) propose a set of metrics informed by music theory, for
 probabilistically evaluating how similar the generations are to known
@@
 pitches and note lengths, then the Kullback-Leibler divergence and
-overlapping area of the probability density functions are used to
-compare against known reference distributions per musical genre
-\cite{yang_evaluation_2020}. Due to the cost and time requirements
-associated with designing a human subjects experiment, we
-utilize this quantitative approach to the quality assessment of
-generated samples.
+overlapping area of the probability density or mass functions are
+used to compare against known reference distributions per musical
+genre \cite{yang_evaluation_2020}. Due to the cost and time
+requirements associated with designing a human subjects experiment for
+qualitative evaluation, we utilize this quantitative approach to the
+assessment of generated samples.
+
+Figure \ref{pitch_counts} depicts histograms of source and generated
+music samples respectively (1 sample = 1 measure) by the number of
+distinct pitch counts they contain. In both samples the mode is 9
+distinct pitches, but the source arrangements have a median of 11
+while the generations have a median of 8. This supports our own
+qualitative observation that the generated samples have slightly less
+pitch diversity than the genuine compositions.
 
@@
     \begin{center}
-        \scalebox{0.6}{\input{figures/source_pitch_count.pgf}}
+        \scalebox{0.6}{\input{figures/pitch_count_histogram.pgf}}
     \end{center}
-    \caption{Distinct pitch counts per sample for source samples.}
-    \label{source_pitches}
+    \caption{Distinct pitch counts per sample for generated and source samples.}
+    \label{pitch_counts}
 \end{figure}
 
-\begin{figure}[htbp]
-    \begin{center}
-        \scalebox{0.6}{\input{figures/generated_pitch_count.pgf}}
-    \end{center}
-    \caption{Distinct pitch counts per sample for generated samples.}
-    \label{generated_pitches}
-\end{figure}
+Figure \ref{pch} depicts the source and generated samples' proportion
+of notes by each of the 12 notes in the chromatic scale, regardless of
+octave, where 0 = C, 1 = C$\sharp$/D$\flat$, 2 = D, and so on. These
+distributions are very similar, with the generated samples having more
+Cs and fewer Ds than the source samples.
 

### 2021-06-08 scripts/string_quartet_evaluation.py

@@ -75,20 +75,23 @@ def pitch_count(df):
         .replace(processing.SUSTAIN, np.nan)[["sample", "value"]]
         .groupby("sample")
         .nunique()
-        .reset_index()["value"]
+        .reset_index()[["value"]]
     )
-    return df.loc[df != 0]
+    df["count"] = 1
+    df = df.groupby("value")["count"].sum().reset_index()
+    df["count"] = df["count"] / df["count"].sum()
+    return df.loc[df.value != 0]
 
 
 pc_x = pitch_count(x)
 pc_xgen = pitch_count(xgen)
+pc_x["class"] = "source"
+pc_xgen["class"] = "generated"
+pc_df = pd.concat([pc_x, pc_xgen])
 
 # %%
 FIGURES = Path("../papers/final/alex/figures/")
-sns.histplot(pc_x, stat="density", binrange=(0, 30))
-
-# %%
-sns.histplot(pc_xgen, stat="density", binrange=(0, 30))
+sns.barplot(data=pc_df, x="value", y="count", hue="class")
 
 # %% [markdown]
 #
@@ -104,6 +107,7 @@ def pitch_count_histogram(df):
     df["value"] = df["value"] % 12
     df = df.groupby(["value"]).count().reset_index()
     df["sample"] = df["sample"] / df["sample"].sum()
+    df["value"] = df["value"].astype(int)
     return df
 
 

### 2021-06-08 papers/final/alex/paper.tex
write discussion section of paper

@@
 metrics include counts, ranges, histograms and transition matrices of
-pitches and note lengths, then the Kullback-Leibler divergence and
-overlapping area of the probability density or mass functions are
-used to compare against known reference distributions per musical
-genre \cite{yang_evaluation_2020}. Due to the cost and time
+pitches and note lengths, then compare the probability density or mass
+functions of generated samples against that of known genuine source
+samples \cite{yang_evaluation_2020}. Due to the cost and time
 requirements associated with designing a human subjects experiment for
 qualitative evaluation, we utilize this quantitative approach to the
-assessment of generated samples.
+assessment of the pitch distributions of generated samples.
 
@@
 
+Figures \ref{source_ranges} and \ref{generated_ranges} depict the
+distribution of pitch range (highest pitch - lowest pitch) among
+samples. The distributions appear similar; source distribution has a
+mean of 30.5 while the generated distribution has a mean of 26.3, (the
+medians are similar to the means for both) so the generated
+distributions have a slightly narrower pitch range.
+
 \begin{figure}[htbp]
@@
 \end{figure}
+
+Figures \ref{source_interval} and \ref{generated_interval} describe
+the distributions average pitch intervals between consecutively played
+notes; these distributions are very similar, both are centered with
+approximate mean of 0 and standard deviation of 1.5.
 
@@
 
+Finally, figures \ref{pctm_source} and \ref{pctm_generated} depict
+heatmaps of transition probability between the twelve pitch classes in
+the chromatic scale. The source samples exhibit higher probabilities
+of transition between different notes, while in the generated sample
+there is more repetition of the same note, again confirming the
+relative lack of pitch diversity in the generated samples. However,
+the dark areas of very low probability in the source distribution are
+similarly dark in the generated distribution; the model is unlikely to
+generate pitch transitions that are rare in the source distribution.
+
 \begin{figure}[htbp]
@@
 
+Our research demonstrates that a relatively simple LSTM variational
+autoencoder network can learn a continuous latent space for generating
+realistic linear interpolations between symbolic, classical string
+quartet pieces, which are a notably complex genre of music.
+Furthermore, our model can achieve this while working
+measure-by-measure, on sequences just 16 time steps in length, without
+the need for a highly complex neural architecture capable of
+memorizing very long sequences. Listening to the generated samples and
+profiling their pitch distributions reveals that some of the pitch
+diversity of the original music is lost, but the overall melodic and
+rhythmic structure still generally resemble the originals. It was our
+aim to produce interpolations that retained enough aesthetic quality
+to be pleasing to listen to and useful in inspiring the creative
+process, and we believe our work meets that goal.
+
+Future work in this domain could involve disentanglement of the latent
+dimensions to enable music style transfer and semantic editing via
+interpolation in specific directions within the latent space. The VAE
+model could be adapted to other musical genres aside from string
+quartets, to interpolate within or across genres. The model could also
+be modified into a conditional VAE to allow conditioning on the genre,
+composer/artist name, or some other defining attribute label, to
+influence the generated results.
+
 \bibliographystyle{ACM-Reference-Format}

### 2021-06-08 scripts/string_quartet_evaluation.py

@@ -79,7 +79,7 @@ def pitch_count(df):
     )
     df["count"] = 1
     df = df.groupby("value")["count"].sum().reset_index()
-    df["count"] = df["count"] / df["count"].sum()
+    df["probability"] = df["count"] / df["count"].sum()
     return df.loc[df.value != 0]
 
 
@@ -91,8 +91,8 @@ pc_df = pd.concat([pc_x, pc_xgen])
 
 # %%
 FIGURES = Path("../papers/final/alex/figures/")
-sns.barplot(data=pc_df, x="value", y="count", hue="class")
-
+sns.barplot(data=pc_df, x="value", y="probability", hue="class")
+plt.savefig(FIGURES / "pitch_count_histogram.pgf")
 # %% [markdown]
 #
 # ## Pitch Class Histogram
@@ -106,7 +106,7 @@ def pitch_count_histogram(df):
     )
     df["value"] = df["value"] % 12
     df = df.groupby(["value"]).count().reset_index()
-    df["sample"] = df["sample"] / df["sample"].sum()
+    df["probability"] = df["sample"] / df["sample"].sum()
     df["value"] = df["value"].astype(int)
     return df
 
@@ -116,9 +116,12 @@ pch_x = pitch_count_histogram(x)
 pch_xgen = pitch_count_histogram(xgen)
 
 # %%
-pch = pd.DataFrame(dict(note=pch_x["value"], source=pch_x["sample"], generated=pch_xgen["sample"]))
+pch = pd.DataFrame(
+    dict(note=pch_x["value"], source=pch_x["probability"], generated=pch_xgen["probability"])
+)
 pch = pd.melt(pch, id_vars="note")
-sns.barplot(data=pch, x="note", y="value", hue="variable")
+pch = pch.rename(columns={"value": "probability", "variable": "class"})
+sns.barplot(data=pch, x="note", y="probability", hue="class")
 
 # %% [markdown]
 #

### 2021-06-08 papers/final/jack/paper.tex

@@
 
-To do: figure for encoding of notes and chords to numbers
-
 \subsection{Model Design}
@@
   \centering
-  \includegraphics[width=0.7\linewidth]{lstm_model.png}
+  \includegraphics[width=0.6\linewidth]{lstm_model.png}
   \caption{Architecture of the LSTM model.}
@@
   \centering
-  \includegraphics[width=0.7\linewidth]{bidirect_lstm_model.png}
+  \includegraphics[width=0.6\linewidth]{bidirect_lstm_model.png}
   \caption{Architecture of the bidirectional LSTM model.}
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{attention_model.png}
+  \includegraphics[width=0.6\linewidth]{attention_model.png}
   \caption{Architecture of the bidirectional LSTM model with attention mechanism.
@@
   \centering
-  \includegraphics[width=0.7\linewidth]{wavenet_model.png}
-  \caption{Architecture of the WaveNet-like 1-D dilated causal convolutional neural network.}
+  \includegraphics[width=0.6\linewidth]{wavenet_model.png}
+  \caption{Architecture of the WaveNet-style 1-D dilated causal convolutional neural network.}
   \label{wavenet_model}
@@
 Similar to the evaluation of GANs, evaluating music generating models is a challenging 
-task since there is a lack of metrics that are used for supervised machine learning models.
+due to the lack of metrics that are used for supervised machine learning models.
+In this paper, we evaluated the generative model by rating their outputs against our own
+reference of music.
 
@@
   \item learning rate: 0.01, \textbf{0.001}, 0.0001
-  \item number of filters in the first Conv1D layer: 32, \textbf{64}, 128, 256
-  \item number of filters in the first Conv1D layer: 32, 64, \textbf{128}, 256
-  \item number of filters in the first Conv1D layer: 32, 64, 128, \textbf{256}
-  \item number of filters in the first Conv1D layer: 32, 64, 128, \textbf{256}
-  \item number of filters in the first Conv1D layer: 32, 64, 128, \textbf{256}
+  \item number of filters in the first Conv1D layer: \textbf{64}, 128, 256
+  \item number of filters in the first Conv1D layer: 64, \textbf{128}, 256
+  \item number of filters in the first Conv1D layer: 64, 128, \textbf{256}
+  \item number of filters in the first Conv1D layer: 64, 128, \textbf{256}
+  \item number of filters in the first Conv1D layer: 64, 128, \textbf{256}
   \item dropout rate: \textbf{0.2}, 0.3, 0.4, 0.5
@@
 
-The models were trained for 100 epochs.
+The models were trained for 100 epochs. After training, the models were tested by evaluating
+ their generated outputs. The winning set of hyperparameters are the combinations that result
+in interestingly sounding music with a lot of variations of notes and chords. Ten samples
+were generated from each trained model and were assessed qualitatively. The Bidirectional
+LSTM model with attention mechamism generated the most interesting outputs with lots of
+variations in the notes and chords.
 
-\section{Sample Output}
+\section{Sample Outputs}
 
-A collection of music generated by the LSTM models and 1-D dilated
-causal convolutional neural network is available at 
-\url{https://drive.google.com/drive/folders/1SlQOI_cL452PftO3KmOYO_qglE2RotCX?usp=sharing}.
+A collection of music generated by the LSTM models and 1-D dilated causal convolutional 
+neural network is available at 
+\sloppy
+\url{https://drive.google.com/drive/u/1/folders/1IO3X1OMwHKwNm26uXmez6XGLeyNLG1PZ}.
 
 \section{Conclusion and Discussion}
+
+Automatic music generation is an interesting problem that has recently benefited from the
+development of advanced machine learning techniques. Despite being in infancy,
+research has shown that deep learning generative models could produce interesting music 
+with little to no human intervention. In this project, we experimented with four different
+deep learning architectures, LSTM, bidirectional LSTM, bidirectional LSTM with attention
+mechanism, and a WaveNet-inspired 1-D dilated causal convolutional neural network. By 
+qualitative evaluation, we determined that the bidirectional LSTM model with attention 
+mechanism generates the most interesting pieces of music compared to the other models.
+This work, although is not directly related to my partner Alex Kyllo work on generative
+models for string-quartet-based classical music, demonstrates a different perspective 
+on single-track piano music generation by treating a short sequence of musical notes
+and chords as a written language sentence with vocabulary and grammar.
+
+To expand upon this project, additional work is needed to search for more sophisticated
+deep learning architectures that can capture the short-term, mid-range, and long-term 
+dependencies among the notes of a musical composition. Furthermore, these architectures 
+should be generalized to music that is performed by multiple instruments (multi-track 
+polyphonic music)). Another improvement is to enable the generative models to work with
+multiple musical genres, not just classical music. Finally, we only relied on our own
+qualitative evaluation to assess the quality of the models due to the time constraint
+of the project. We also need a comprehensive evaluation plan, which involves qualitative evaluation 
+by the multiple human judges and quantitative evaluation on the structure of the generative 
+musical sequences.
 

### 2021-06-08 papers/final/alex/paper.tex
regenerate notebooks from updated scripts

@@
   generating pieces of multi-instrument symbolic classical music
-  through interpolation. The code utilized for this research is available at:
-  \url{https://github.com/jackPhn/CSS586_Project/}
+  through linear interpolation. The code utilized for this research is
+  available at: \url{https://github.com/jackPhn/CSS586_Project/}
 \end{abstract}
@@
 \keywords{deep learning, recurrent neural networks, sequential models,
-  music modeling, latent space modeling, generative modeling}
+  autoencoders, music modeling, latent space modeling, generative
+  modeling}
 
@@
 before it can consistently generate music scores or performances that
-would be enjoyable and popular for humans to listen to, but for this
+would be broadly enjoyable for humans to listen to, but for this
 reason it is an area of significant opportunity, where a number of
@@
 software tools to expose these models for use by creative artists and
-students.
-
-MusicVAE, part of the Magenta project, is a variational Long
+students. MusicVAE, part of the Magenta project, is a variational Long
 Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
@@
 While there may be significant opportunity in future approaches that
-combining the two, contemporary research generally treats symbolic and
+combine the two, contemporary research generally treats symbolic and
 audio music modeling as separate problem domains and this work does
@@
 
-In order to accommodate polyphonic music, we each MIDI file is
-converted into a modified pianoroll representation, an example of
-which is visualized in Figure \ref{pianoroll}. While the standard
-pianoroll represents each time-step as a multi-hot encoded vector of
-note values, because most notes are not being played at most timesteps
+In order to accommodate polyphonic music, each MIDI file is converted
+into a modified pianoroll representation, an example of which is
+visualized in Figure \ref{pianoroll}. While the standard pianoroll
+represents each time-step as a multi-hot encoded vector of note
+values, because most notes are not being played at most timesteps
 there is an extreme class imbalance and sparsity in the matrix. This
-is reported in \cite{kumar2019polyphonic} to cause problems in model
-fitting and metrics calculation, so they introduce an alternative
-``multi-stream'' representation of the pianoroll, which we closely,
-(but not exactly) emulate in this work.
+is reported in Kumar and Ravindran (2019) \cite{kumar2019polyphonic}
+to cause problems in model fitting and metrics calculation, so they
+introduce an alternative ``multi-stream'' representation of the
+pianoroll, which we closely, (but not exactly) emulate in this work.
 
@@
 by the violins, viola and cello, 64 is rest and 65 is sustain. Figure
-\ref{pitch_dist} depicts the distribution of these pitch values per
-instrument. Our software package also includes functions to convert
-this representation back into a Music21 Score object, which can then
-be written to a MIDI file.
+\ref{pitch_dist} depicts the distribution of MIDI pitch values per
+instrument and the degree to which they overlap.
+
+Our software package also includes functions to convert this
+representation back into a Music21 Score object, which can then be
+written to a MIDI file that can be synthesized for human listening.
 
@@
 Because songs are typically at least a few minutes long and of varying
-length, it will not be feasible to train with entire songs as
-examples, so we will crop songs into phrases of equal numbers of
-beats to use as training data.
-
-The result of this preprocessing is that each training example isbe a
-2D matrix of shape (time steps x instrument tracks) and stacking the
-training examples produces a 3D tensor. Information regarding which
-track corresponds to which musical instrument is lost and therefore
-must be maintained separately to reconstruct the MIDI
+length, it is not feasible to train with entire songs as examples, so
+we crop songs into phrases of equal numbers of beats to use as
+training data. The result of this preprocessing is that each training
+example is a 2D matrix of shape (time steps x instrument tracks) and
+stacking the training examples produces a 3D tensor. Information
+regarding which track corresponds to which musical instrument is lost
+and therefore must be maintained separately to reconstruct the MIDI
 representation. For string quartets this instrument ensemble is always
@@
 representing two violins, one viola and one cello. Tempo information
-is also lost from this representation, so the outputs are normalized
-to the default tempo of 120 beats per minute.
+is also lost from this representation, which uses symbolic time steps
+rather than wall clock time, so the outputs are standardized to the
+default tempo of 120 beats per minute.
 
@@
 \cite{roberts_hierarchical_2018} and MuseGAN models five-part (drum,
-bass, guitar, string, piano) arrangements. Therefore this model is
-purpose-built for string quartets, which are the most common type of
-arrangement found in the MusicNet dataset.
+bass, guitar, string, piano) arrangements. Therefore the number of
+tracks is a hyperparameter to the network architecture definition and
+this model is trained for string quartets, which are the most common
+type of arrangement found in the MusicNet dataset.
 
@@
 two LSTM layers, with a dropout layer in between, convert the
-sequences into latent codes, which are modeled as a multidimensional
+sequences into latent codes, which are modeled as a 512-dimensional
 Gaussian distribution $z$, produced by the encoder output Lambda layer
@@
 activation is used to select the most probable note value for the
-the current track at the each time step (Figure \ref{decoder}).
+the current track at each time step (Figure \ref{decoder}).
 
@@
 dataset. As a result, we saw much lower validation losses (compare
-Figure \ref{loss} and Figure \ref{loss_augment} and a small but
+Figure \ref{loss} and Figure \ref{loss_augment}) and a small but
 noticeable improvement in the fidelity of the song reconstructions,
@@
 
-The interpolations are between the following 18 pairs of pieces:
+The interpolations are between the following 18 pairs of pieces
+denoted by their filenames in the original MusicNet reference MIDI
+archive, which can be downloaded from:
+\url{https://homes.cs.washington.edu/~thickstn/media/musicnet_midis.tar.gz}
 
@@
 distribution of pitch range (highest pitch - lowest pitch) among
-samples. The distributions appear similar; source distribution has a
-mean of 30.5 while the generated distribution has a mean of 26.3, (the
-medians are similar to the means for both) so the generated
+samples. The distributions appear similar; the source distribution has
+a mean of 30.5 while the generated distribution has a mean of 26.3,
+(the medians are similar to the means for both) so the generated
 distributions have a slightly narrower pitch range.
@@
 interpolation in specific directions within the latent space. The VAE
-model could be adapted to other musical genres aside from string
-quartets, to interpolate within or across genres. The model could also
-be modified into a conditional VAE to allow conditioning on the genre,
-composer/artist name, or some other defining attribute label, to
-influence the generated results.
+model could also be adapted to other musical genres aside from string
+quartets, to interpolate within or across genres.
 

### 2021-06-08 scripts/string_quartet_evaluation.py

@@ -1,3 +1,11 @@
+# %% [markdown]
+# # Quantitative evaluation metrics for generated samples
+#
+# Author: Alex Kyllo
+#
+# Date: 2021-06-06
+
+
 # %%
 import os
 import sys

### 2021-06-08 scripts/string_quartet_inference.py

@@ -2,6 +2,7 @@
 # # String quartet MTVAE model for interpolation
 #
 # Author: Alex Kyllo
+#
 # Date: 2021-06-06
 
 # %% [markdown]

### 2021-06-08 musiclearn/config.py
comments

@@ -1,3 +1,8 @@
+"""config.py
+Configuration constants.
+Authors: Jack Phan and Alex Kyllo
+"""
+
 import os
 
 import dotenv

### 2021-06-08 musiclearn/plotting.py

@@ -1,6 +1,7 @@
 """plotting.py
 Functions for plotting results.
 Author: Alex Kyllo
+Date: 2021-06-05
 """
 import os
 

### 2021-06-08 musiclearn/processing.py

@@ -1,6 +1,7 @@
 """processing.py
 Preprocessing audio data (.wav files) for analysis.
 Author: Alex Kyllo
+Date: 2021-06-06
 """
 import logging
 import os

### 2021-06-08 musiclearn/training.py

@@ -1,6 +1,8 @@
 """training.py
 
 Helper functions for running experiments.
+Author: Alex Kyllo
+Date: 2021-06-06
 """
 import logging
 import os

### 2021-06-08 musiclearn/vae_models.py

@@ -1,6 +1,7 @@
 """vae_models.py
 Music LSTM Variational Autoencoder models
 Author: Alex Kyllo
+Date: 2021-06-06
 """
 import itertools
 import os
